### 2020-01-09 MANUSCRIPT/manuscript.Rmd
First commit

---
title: "Effects without a Cause: The Search for Demographic Anomalies"

# to produce blinded version set to 1
blinded: 0

authors: 
- name: Mathew E. Hauer
  thanks:  Thanks y'all!
  affiliation: Department of Sociology, Florida State University

- name: Stephanie A. Bohon
  affiliation: Department of Sociology, University of Tennessee - Knoxville

keywords:
- mortality
- fertility
- causal inference

abstract: |
  The proliferation of data, modern computing advances, and powerful statistical algorithms unlock the potential search for hidden, previously un- or understudied demographic anomalies. Here, we investigate US state-level fertility and mortality time series since 1999 to uncover the hidden baby booms/busts and mortality plagues?/non-plauges?. We find 38 states exhibited at least one mortality anomaly, totalling more than 318k anomalous deaths and an additional 6 states exhibited at least one mortality non-plague? totalling 164k protective deaths. 12 states exhibited baby busts, totalling more than 240k missing births and an additional 11 states exhibited baby booms totalling more than 134k additional births. These results suggest the widespread detection of demographic anomalies. Our analysis does not examine the *causes* of these anomalies and our results point to important further research on the causes of anomalous demographic behavior.

bibliography: mybibfile
output: rticles::asa_article
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}

---

\newpage

```{r setup, include=FALSE}
rm( list=ls() )
knitr::opts_chunk$set(echo  = TRUE)

library(knitr)
read_chunk('../R/SCRIPTS/000-Libraries.R')
```

```{r libraries, include=FALSE}
```

# Introduction

The proliferation of data, advances in high performance (“super”) computing, and the development of powerful statistical algorithms mark the era of "Big Data" or data science [@van2016data;@zikopoulos2011], with the potential for researchers to find hidden or understudied social phenomena [@bohon2018demography]. While the availability of Big Data and high performance computing allows novel exploration of data through causal inference [@bohon2018demography;@rcausalimpact;@shiffrin2016drawing], relatively few studies utilize causal inference techniques in the study of demographic phenomena. However, understanding social phenomena using these advances reveals important insights into society [@angrist1989lifetime;@mas2009peers] and allows us to better monitor population trends [@nobles2019;@torche2015hidden].

Causal inference approaches in social sciences have a long history [@grimmer2015ppsp]. Often, the identification of a casual mechanism requires either a randomized control trial (RCT) or expert knowledge applied to a natural experiment. For population-level analysis, expensive RCTs usually preclude their widespread adoption [@west2008ajph]. Using natural experiments, however, provides important and more widespread insights on the determinants of migration after Hurricane Katrina in 2006 [@fussellRecoveryMigrationCity2014;@horiDisplacementDynamicsSouthern2009], mini baby booms after electrical blackouts [@fetzer2018jpe], and the highly publicized estimates of excess mortality after Hurricane Maria in Puerto Rico in 2016 [@kishore2018mortality;@santos2018use]. Both RCTs and natural experiments follow traditional, hypothesis testing, inductive scientific paradigms where a question is first posed and scientists generate or find data to answer the question. Such approaches, while extensively utilized and time-tested, are likely to miss important phenomena that might go unnoticed.

Abductive modeling is the movement from the inductive to the deductive, and sometimes back and forth, to reach conclusions [@bryant2014realm] or “inferring cause from effect” [@Crowder2017]. In situations with large data sets, an abductive approach is far superior to deductive hypothesis testing, as p-values with an extremely large number of cases are far from revealing [@head2015extent;@nuzzo2014scientific] and we do not want to make purely inductive inferences from data of questionable generalizability [@ruggles2014big]. Owing to the long history of big datasets in demographic research (one of the original "big data" resources [@ruggles2014big]), the rich demographic data available in the United States make the potential revelation of interesting and important demographic phenomena not only possible, but extremely plausible – even if identification of the phenomena occurs without identifying the underlying cause. In other sciences, the identification of effects without causes has led to new theories of galactic migration of planets and other breakthroughs [@gomes2005n].

In this paper, we use modern statistical outlier detection algorithms [@chen1993joint] on nearly twenty years of mortality and fertility data at the US state-level to identify anomalous demographic behavior. In essence, we identify *effects without knowledge of the cause*. We ask two questions regarding demographic anomalies: What are the hidden baby booms/busts and mortality spikes/dips in the United States over the last twenty years? We do not necessarily know the causes of these anomalies but identifying them allows scientists with more detailed knowledge of local population dynamics, state-level policy making, or macro-economics to explain these phenomena post-hoc. From explanations of phenomena that may have previously gone unnoticed, demographers may be able to better forecast populations and provide policy solutions for impending problems.

# Materials and Methods

## Method
We use a statistical time series outlier detection algorithm [@chen1993joint], implemented in the R programming language [@rcore] via the tsoutliers package [@tsoutliers2019]. This algorithm iteratively uses ARIMA models to 1) produce a counter-factual time series to initially detect an outlier or anomaly, and 2) refit the ARIMA with the outliers removed. Here we briefly summarize and describe the method.

Often, the behavior of a time series can be described and summarized in ARIMA models. If a series of values, $y_t^*$, is subject to $m$ interventions or outliers at time points $t_1,t_2,…,t_m$, then $y_t^*$ can be defined as

$$y_t^* = \sum_{j=1}^{m} \omega_jL_j(B)I_t(t_j) + \frac{\theta(B)}{\phi(B)\alpha(B)}\alpha_t$$

Where
$I_t(t_j)$ is an indicator variable with a value of 1 at observation $t_j$ and where the $j$th outlier arises,
$\phi(B)$ is an autoregressive polynomial with all roots outside the unit circle,
$\theta(B)$ is a moving average polynomial with all roots outside the unit circle,
and $\alpha(B)$ is an autoregressive polynomial with all roots on the unit circle.

We examine three types of outliers at time point $t_m$:
1) additive outliers (AO), defined as $L_j(B)=1$;
2) level shift outliers (LS), defined as $L_j(B) = 1/(1-B)$; and
2) temporary change outliers (TC), defined as $L_j(B) = 1/(1-\delta B)$.

Colloquially, additive outliers arise when a single event causes the time series to unexpectedly increase/decrease for a single time period; level shift outliers arise when an event causes the time series to unexpectedly increase/decrease for multiple time periods; and temporary change outliers arise when an event causes the time series to unexpectedly increase/decrease with lingering effects that decay over multiple time periods.

An outlier is detected using a regression equation

$$ \pi(B)y_t^* \equiv \hat{e} = \sum_{j=1}^m \omega_j \pi(B)L_j(B)I_t(t_j) + \alpha_t $$

where $\pi(B)=\sum_{i=o}^{inf} \pi_iB^i$. The identification of outliers then involves a three step process to (1) identify all potential outliers, $t_j$ and $L_j (B)$, (2) joint estimates of model parameters and outlier effects are computed to identify potentially spurious outliers, and (3) the outliers and effects are re-estimated without spurious outliers. Of importance here is step 2, which relies on some critical value above which an outlier at time point m is considered spurious. Based on Chien and Liu’s [-@chen1993joint] recommendation, we set a critical value of 3.5 which generally minimizes the possibility of Type I errors or false positive outliers.

Outliers are then reported using a simple t-statistic.

## Data

We search for demographic anomalies using the Center for Disease Control and Prevention’s online WONDER monthly fertility (2003-2017) and mortality databases (1999-2016) for all fifty states and the District of Columbia
[@CDC_fert07;@CDC_mort]. These data sets contain every birth and death record in the United States over the time periods of interest, representing the universe of both mortality and fertility data in the US.  These data are considered the “gold standard” of data collections [@mahapatra2007civil] and have been considered “complete” since 1968 [@hetzel2016us]. We search over each state equivalent’s (n=51) mortality (n=228) and fertility (n=180) monthly time series for a total of 20,808 state-months of data.

# Results
We detect numerous anomalous mortality and fertility events at the US state-level since 1999. A full listing of these anomalies can be found in the **Supplementary Materials**. We begin with a summary of the anomalies we detect and then we highlight highly significant mortality and fertility anomalies across all three types of outliers (Additive Outliers, Level Shift Outliers, and Temporary Change Outliers) with plausible explanations.

## Overall Anomalies

## Mortality

We begin with mortality for New York State (Figure 1). We identify seven anomalies in the mortality time series for New York, all with t-statistics in excess of 3.91, making these anomalies highly significant. The algorithm correctly identifies September 2001 as an additive outlier (2001:09 t=6.40) where mortality in that month was 1,628 higher than anticipated. Clearly, this mortality event is likely caused by the September 11 tragedy and the detection of this mortality event provides confidence in our detection of other anomalies.

\newpage

### 2020-01-09 MANUSCRIPT/manuscript.tex
First commit

% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{textcomp}
\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

%% load any required packages here

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}

\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf Effects without a Cause: The Search for Demographic Anomalies}

  \author{
        Mathew E. Hauer \thanks{Thanks y'all!} \\
    Department of Sociology, Florida State University\\
     and \\     Stephanie A. Bohon \\
    Department of Sociology, University of Tennessee - Knoxville\\
      }
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Effects without a Cause: The Search for Demographic Anomalies}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
The proliferation of data, modern computing advances, and powerful
statistical algorithms unlock the potential search for hidden,
previously un- or understudied demographic anomalies. Here, we
investigate US state-level fertility and mortality time series since
1999 to uncover the hidden baby booms/busts and mortality
plagues?/non-plauges?. We find 38 states exhibited at least one
mortality anomaly, totalling more than 318k anomalous deaths and an
additional 6 states exhibited at least one mortality non-plague?
totalling 164k protective deaths. 12 states exhibited baby busts,
totalling more than 240k missing births and an additional 11 states
exhibited baby booms totalling more than 134k additional births. These
results suggest the widespread detection of demographic anomalies. Our
analysis does not examine the \emph{causes} of these anomalies and our
results point to important further research on the causes of anomalous
demographic behavior.\\
\end{abstract}

\noindent%
{\it Keywords:} mortality, fertility, causal inference
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!

\newpage

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The proliferation of data, advances in high performance (``super'')
computing, and the development of powerful statistical algorithms mark
the era of ``Big Data'' or data science
\citep{van2016data, zikopoulos2011}, with the potential for researchers
to find hidden or understudied social phenomena
\citep{bohon2018demography}. While the availability of Big Data and high
performance computing allows novel exploration of data through causal
inference
\citep{bohon2018demography, rcausalimpact, shiffrin2016drawing},
relatively few studies utilize causal inference techniques in the study
of demographic phenomena. However, understanding social phenomena using
these advances reveals important insights into society
\citep{angrist1989lifetime, mas2009peers} and allows us to better
monitor population trends \citep{nobles2019, torche2015hidden}.

Causal inference approaches in social sciences have a long history
\citep{grimmer2015ppsp}. Often, the identification of a casual mechanism
requires either a randomized control trial (RCT) or expert knowledge
applied to a natural experiment. For population-level analysis,
expensive RCTs usually preclude their widespread adoption
\citep{west2008ajph}. Using natural experiments, however, provides
important and more widespread insights on the determinants of migration
after Hurricane Katrina in 2006
\citep{fussellRecoveryMigrationCity2014, horiDisplacementDynamicsSouthern2009},
mini baby booms after electrical blackouts \citep{fetzer2018jpe}, and
the highly publicized estimates of excess mortality after Hurricane
Maria in Puerto Rico in 2016
\citep{kishore2018mortality, santos2018use}. Both RCTs and natural
experiments follow traditional, hypothesis testing, inductive scientific
paradigms where a question is first posed and scientists generate or
find data to answer the question. Such approaches, while extensively
utilized and time-tested, are likely to miss important phenomena that
might go unnoticed.

Abductive modeling is the movement from the inductive to the deductive,
and sometimes back and forth, to reach conclusions
\citep{bryant2014realm} or ``inferring cause from effect''
\citep{Crowder2017}. In situations with large data sets, an abductive
approach is far superior to deductive hypothesis testing, as p-values
with an extremely large number of cases are far from revealing
\citep{head2015extent, nuzzo2014scientific} and we do not want to make
purely inductive inferences from data of questionable generalizability
\citep{ruggles2014big}. Owing to the long history of big datasets in
demographic research (one of the original ``big data'' resources
\citep{ruggles2014big}), the rich demographic data available in the
United States make the potential revelation of interesting and important
demographic phenomena not only possible, but extremely plausible -- even
if identification of the phenomena occurs without identifying the
underlying cause. In other sciences, the identification of effects
without causes has led to new theories of galactic migration of planets
and other breakthroughs \citep{gomes2005n}.

In this paper, we use modern statistical outlier detection algorithms
\citep{chen1993joint} on nearly twenty years of mortality and fertility
data at the US state-level to identify anomalous demographic behavior.
In essence, we identify \emph{effects without knowledge of the cause}.
We ask two questions regarding demographic anomalies: What are the
hidden baby booms/busts and mortality spikes/dips in the United States
over the last twenty years? We do not necessarily know the causes of
these anomalies but identifying them allows scientists with more
detailed knowledge of local population dynamics, state-level policy
making, or macro-economics to explain these phenomena post-hoc. From
explanations of phenomena that may have previously gone unnoticed,
demographers may be able to better forecast populations and provide
policy solutions for impending problems.

\hypertarget{materials-and-methods}{%
\section{Materials and Methods}\label{materials-and-methods}}

\hypertarget{method}{%
\subsection{Method}\label{method}}

We use a statistical time series outlier detection algorithm
\citep{chen1993joint}, implemented in the R programming language
\citep{rcore} via the tsoutliers package \citep{tsoutliers2019}. This
algorithm iteratively uses ARIMA models to 1) produce a counter-factual
time series to initially detect an outlier or anomaly, and 2) refit the
ARIMA with the outliers removed. Here we briefly summarize and describe
the method.

Often, the behavior of a time series can be described and summarized in
ARIMA models. If a series of values, \(y_t^*\), is subject to \(m\)
interventions or outliers at time points \(t_1,t_2,…,t_m\), then
\(y_t^*\) can be defined as

\[y_t^* = \sum_{j=1}^{m} \omega_jL_j(B)I_t(t_j) + \frac{\theta(B)}{\phi(B)\alpha(B)}\alpha_t\]

Where \(I_t(t_j)\) is an indicator variable with a value of 1 at
observation \(t_j\) and where the \(j\)th outlier arises, \(\phi(B)\) is
an autoregressive polynomial with all roots outside the unit circle,
\(\theta(B)\) is a moving average polynomial with all roots outside the
unit circle, and \(\alpha(B)\) is an autoregressive polynomial with all
roots on the unit circle.

We examine three types of outliers at time point \(t_m\): 1) additive
outliers (AO), defined as \(L_j(B)=1\); 2) level shift outliers (LS),
defined as \(L_j(B) = 1/(1-B)\); and 2) temporary change outliers (TC),
defined as \(L_j(B) = 1/(1-\delta B)\).

Colloquially, additive outliers arise when a single event causes the
time series to unexpectedly increase/decrease for a single time period;
level shift outliers arise when an event causes the time series to
unexpectedly increase/decrease for multiple time periods; and temporary
change outliers arise when an event causes the time series to
unexpectedly increase/decrease with lingering effects that decay over
multiple time periods.

An outlier is detected using a regression equation

\[ \pi(B)y_t^* \equiv \hat{e} = \sum_{j=1}^m \omega_j \pi(B)L_j(B)I_t(t_j) + \alpha_t \]

where \(\pi(B)=\sum_{i=o}^{inf} \pi_iB^i\). The identification of
outliers then involves a three step process to (1) identify all
potential outliers, \(t_j\) and \(L_j (B)\), (2) joint estimates of
model parameters and outlier effects are computed to identify
potentially spurious outliers, and (3) the outliers and effects are
re-estimated without spurious outliers. Of importance here is step 2,
which relies on some critical value above which an outlier at time point
m is considered spurious. Based on Chien and Liu's
\citeyearpar{chen1993joint} recommendation, we set a critical value of
3.5 which generally minimizes the possibility of Type I errors or false
positive outliers.

Outliers are then reported using a simple t-statistic.

\hypertarget{data}{%
\subsection{Data}\label{data}}

We search for demographic anomalies using the Center for Disease Control
and Prevention's online WONDER monthly fertility (2003-2017) and
mortality databases (1999-2016) for all fifty states and the District of
Columbia \citep{CDC_fert07, CDC_mort}. These data sets contain every
birth and death record in the United States over the time periods of
interest, representing the universe of both mortality and fertility data
in the US. These data are considered the ``gold standard'' of data
collections \citep{mahapatra2007civil} and have been considered
``complete'' since 1968 \citep{hetzel2016us}. We search over each state
equivalent's (n=51) mortality (n=228) and fertility (n=180) monthly time
series for a total of 20,808 state-months of data.

\hypertarget{results}{%
\section{Results}\label{results}}

We detect numerous anomalous mortality and fertility events at the US
state-level since 1999. A full listing of these anomalies can be found
in the \textbf{Supplementary Materials}. Here, we highlight highly
significant mortality and fertility events across all three types of
outliers (Additive Outliers, Level Shift Outliers, and Temporary Change
Outliers) where we have identified plausible explanations for these
outliers.

\hypertarget{mortality}{%
\subsection{Mortality}\label{mortality}}

We begin with mortality for New York State (Figure 1). We identify seven
anomalies in the mortality time series for New York, all with
t-statistics in excess of 3.91, making these anomalies highly
significant. The algorithm correctly identifies September 2001 as an
additive outlier (2001:09 t=6.40) where mortality in that month was
1,628 higher than anticipated. Clearly, this mortality event is likely
caused by the September 11 tragedy and the detection of this mortality
event provides confidence in our detection of other anomalies.

\newpage

\bibliographystyle{agsm}
\bibliography{mybibfile}

\end{document}

### 2020-01-09 R/SANDBOX/PR_mort.R

@@ -1,0 +1,18 @@
+
+# Reading in the mortality data
+PRdeaths <- readxl::read_xlsx("DATA/Deaths0918.xlsx", sheet =1)
+
+# Generating the TS outliers using just the `Deaths`
+outlier.morttots <- tsoutliers::tso(ts(PRdeaths$Deaths),  maxit.iloop=10)
+
+# Generating the TS outliers using the `Deaths` and the regressor variable `Population_Estimate`
+outlier.mortots.popest <- tsoutliers::tso(ts(PRdeaths$Deaths), xreg = PRdeaths$Population_Estimate,  maxit.iloop=10)
+
+# Plotting the figure for the total `Deaths`
+plot(outlier.morttots)
+
+# Summing up the difference in the Mortality using only the `Deaths`
+sum(outlier.morttots$y - outlier.morttots$yadj)
+
+# Summing up the difference using the `Deaths` and the `Population_Estimate` regressor
+sum(outlier.mortots.popest$y - outlier.mortots.popest$yadj)

### 2020-01-09 R/SANDBOX/USmigrationLoop-test.R

@@ -1,0 +1,156 @@
+
+
+migdat <- read_tsv("DATA/county_migration_data.txt")
+
+
+
+inmig <- filter(migdat, !origin == destination) %>%
+  group_by(destination) %>%
+  dplyr::select(-origin) %>%
+  summarise_all(sum) %>%
+  mutate(direction = "inmigrants") %>%
+  gather(Years, Migrants, `1990`:`2010`) %>%
+  dplyr::select(location = destination, everything()) %>%
+  mutate(Year = as.Date(as.yearmon(as.numeric(Years))))
+
+outmig <- filter(migdat, !origin == destination) %>%
+  # mutate(state = substr(origin, 1,2))
+  group_by(origin) %>%
+  dplyr::select(-destination, -origin) %>%
+  summarise_all(sum)  %>%
+  mutate(direction = "outmigrants")%>%
+  gather(Years, Migrants, `1990`:`2010`) %>%
+  dplyr::select(location = origin, everything()) %>%
+  mutate(Year = as.Date(as.yearmon(as.numeric(Years))),
+         location = paste0("X", location))
+
+set.seed(1)
+
+outmig <- outmig %>%
+  dplyr::select(-Years, -direction) %>%
+  spread(location, Migrants) %>%
+  dplyr::select(-Year) 
+time.points <- seq.Date(as.Date("1990-01-01"), by ="years", length.out = 21)
+# df2 <- outmig[,i]
+# df3 <- outmig[,-(i+1)]
+# df <- cbind(df2, outmig[, c(sample(1:ncol(df3), 100, replace =F))])
+# outmigdat <- zoo(cbind(df), time.points)
+
+results <- data.frame(matrix( ncol =3))
+colnames(results) <- c("county", "p", "eventyear")
+
+# for (i in 1:ncol(outmig)) {
+#   df2 <- outmig[,i]
+#   df3 <- outmig[,-(i+1)]
+#   df <- cbind(df2, outmig[, c(sample(1:ncol(df3), 100, replace =F))])
+#   outmigdat <- zoo(cbind(df), time.points)
+#   
+#   for(j in 9:(length(time.points)-1)){
+#     temp <-  data.frame(matrix( ncol =3))
+#     colnames(temp) <- c("county", "p", "eventyear")
+#     pre.period <- as.Date(c(time.points[1], time.points[j]))
+#     post.period <- as.Date(c(time.points[j+1], time.points[length(time.points)]))
+#     
[…]

### 2020-01-09 R/SANDBOX/USmortalityLoop-test.R

@@ -1,0 +1,161 @@
+
+
+dat <- read_tsv("DATA/Compressed Mortality, 1999-2016.txt") %>%
+  mutate(Year = as.Date(as.yearmon(as.numeric(Year))),
+         location = paste0("X", `County Code`)) %>%
+  dplyr::select(Year, location, Deaths) %>%
+  na.omit %>%
+  spread(Year, Deaths) %>%
+  na.omit %>%
+  gather(Year, Deaths, `1999-01-01`:`2016-01-01`)
+
+timelength <- year(max(dat$Year)) - year(min(dat$Year))+1
+
+time.points <- seq.Date(as.Date(min(dat$Year)), by ="years", length.out = timelength )
+
+dat<- spread(dat, location, Deaths) 
+
+
+dat <- dplyr::select(dat, -Year)
+
+
+
+
+
+
+
+casualfunc<- function(datframe,j){
+  df2 <- datframe[,i]
+  df3 <- datframe[,-(i+1)]
+  # set.seed(seednum)
+  df <- cbind(df2, datframe[, c(sample(1:ncol(df3), sampsize, replace =F))])
+  outmigdat <- zoo(cbind(df), time.points)
+  
+  # for(j in 9:(length(time.points)-1)){
+  temp <-  data.frame(matrix( ncol =6))
+  colnames(temp) <- c("county", "p", "eventyear", "iteration", "abseffect", "samplesize")
+  # pre.period <- as.Date(c(time.points[1], time.points[j]))
+  # post.period <- as.Date(c(time.points[j+1], time.points[length(time.points)]))
+  pre.period <- as.Date(c(time.points[1], time.points[j]))
+  post.period <- as.Date(c(time.points[j+1], time.points[ifelse(j+2 < length(time.points), j+2, length(time.points))]))
+  
+  
+  tryCatch({#print(x)
+    outimpact <- CausalImpact(outmigdat, pre.period, post.period)
+    # return(outimpact)
+  }
+  , error=function(e){cat(i," ERROR :",conditionMessage(e), "\n")})
+  # endtime <- Sys.time()
+  # (time <- endtime - starttime)
+  temp[1,2] <- ifelse(is.null(outimpact$summary$p[1]),NA,outimpact$summary$p[1])
+  temp[1,1] <- colnames(df2)[1]
+  temp[1,3] <-  paste0(time.points[j])
+  temp[1,4] <-  i
+  temp[1,5] <- outimpact$summary$AbsEffect[2]
[…]

### 2020-01-09 R/SANDBOX/USmortality_monthly-test.R

@@ -1,0 +1,116 @@
+
+
+dat <- 
+ ###MORTALITY 
+  # read_tsv("DATA/Mortality/Underlying Cause of Death, 1999-2017,AL-CT.txt") %>%
+  read_tsv("DATA/Mortality/Underlying Cause of Death, 1999-2017,states.txt") %>%
+  mutate(Month2 = substr(`Month Code`,6,7),
+         Month3 = as.yearmon(paste0(Year,"-", Month2)),
+         # location = paste0("X", `County Code`)
+         location = paste0("X", `State Code`)
+         ) %>%
+  dplyr::select(Year = Month3, location, Target = Deaths) %>%
+  na.omit %>%
+  arrange(location, Year)
+
+dat <-
+  ###FERTILITY
+  bind_rows(read_tsv("DATA/Fertility/Natality, 2003-2006,states.txt"),
+    read_tsv("DATA/Fertility/Natality, 2007-2017,states.txt"))%>%
+  mutate(Month2 = str_pad(`Month Code`,2 , pad = "0"),
+         Month3 = as.yearmon(paste0(Year,"-", Month2)),
+         location = paste0("X", `State Code`)
+  ) %>%
+  dplyr::select(Year = Month3, location, Target = Births) %>%
+  na.omit %>%
+  arrange(location, Year)
+
+
+
+# timelength <- year(max(dat$Year)) - year(min(dat$Year))+1
+
+# time.points <- seq.Date(as.Date(min(dat$Year)), by ="years", length.out = timelength )
+# 
+# dat<- spread(dat, location, Deaths) 
+
+countylist = unique(dat$location)[1]
+
+df <- data.frame()
+df_sums <- data.frame()
+sigma <- 6
+
+for(i in 1:length(unique(countylist))){
+  tryCatch({
+df2<- setNames(data.frame(matrix(ncol = 5, nrow = 0)), c("type", "ind", "time", "coefhat", "tstat"))
+dat2 <- filter(dat, location == paste0(countylist[i]))
+dat3 <- ts(dat2$Target, start = year(min(dat$Year)), end = c(year(max(dat$Year)),12), frequency = 12)
+outlier.county <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),maxit.iloop=10)
+
+ df_sums2 <- data.frame(y = NA, yadj=NA)
+if(!is.null(outlier.county$times)){
+ 
+  df2<- outlier.county$outliers
[…]

### 2020-01-09 R/SCRIPTS/000-Libraries.R

@@ -1,6 +1,5 @@
 ###------LIBRARY SETUP-----
 ## @knitr libraries
-
 rm(list = ls()) # Remove Previous Workspace
 gc(reset = TRUE) # Garbage Collection
 
@@ -40,33 +39,30 @@ pkgs <- c(
   "openxlsx",      # Microsoft Excel Files
   "stringi",       #Character/String Editor
   "stringr",       # Character/String Editor
+  "zoo",           # Time Series
   "reshape2",      # Data Management/Manipulation
   "scales",        # Number formatting
   "cowplot",       # Plot Grids
   "tmap",          # Cartography
   "tmaptools",     # Cartographic tools
+  "tigris",        # US shapefiles
+  "censusapi",     # Census Data
   "sp",            # Spatial Objects
   "grid",          # Plot Grids
   "kableExtra",    # Pretty Tables
+  "LexisPlotR",    # Lexis Diagrams
   "pdftools",      # Load pdfs
   "R.utils",       # Utilities
+  "forecast",      # Forecasting
+  "pbmcapply",     # Progress Bar Multicore Apply
+  "parallelsugar", # Parallel apply
+  "rucm",          # UCM
   "IDPmisc",        # Quality na.rm
-  "ggrepel",
-  "raster",
-  "gridExtra",
-  "magick",
-  "pdftools",
-  "readxl",
-  "zoo",
-  "bsts",
+  "tidycensus",     # Census Data
   "CausalImpact",
-  "kableExtra",
-  "lettercase",
-  "philentropy",
-  "ggridges",
-  "janitor",
-  "gghighlight"
-  
+  "bsts",
+  "lubridate",
+  "tsoutliers"
 )
 
 # Install missing packages

### 2020-01-09 R/SCRIPTS/001-AnomaliesDetection.R

@@ -1,0 +1,126 @@
+###------Mort Loop-----
+## @knitr MortLoop
+
+dat <-
+  ###FERTILITY
+  bind_rows(read_tsv("./R/DATA-RAW/Fertility/Natality, 2003-2006,states.txt"),
+            read_tsv("./R/DATA-RAW/Fertility/Natality, 2007-2017,states.txt"))%>%
+  mutate(Month2 = str_pad(`Month Code`,2 , pad = "0"),
+         Month3 = as.yearmon(paste0(Year,"-", Month2)),
+         location = paste0("X", `State Code`)
+  ) %>%
+  dplyr::select(Year = Month3, location, Target = Births) %>%
+  na.omit %>%
+  arrange(location, Year)
+
+countylist = unique(dat$location)
+
+df <- data.frame()
+df_sums <- data.frame()
+sigma <- 3.5
+
+for(i in 1:length(unique(countylist))){
+  tryCatch({
+    df2<- setNames(data.frame(matrix(ncol = 5, nrow = 0)), c("type", "ind", "time", "coefhat", "tstat"))
+    dat2 <- filter(dat, location == paste0(countylist[i]))
+    dat3 <- ts(dat2$Target, start = year(min(dat$Year)), end = c(year(max(dat$Year)),12), frequency = 12)
+    outlier.county <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)
+    
+    df_sums2 <- data.frame(y = NA, yadj=NA)
+    if(!is.null(outlier.county$times)){
+      
+      df2<- outlier.county$outliers
+      df_sums2$y <- sum(outlier.county$y)
+      df_sums2$yadj <- sum(outlier.county$yadj)
+    } else {
+      df2 <- rbind(df2, data.frame(type = NA,
+                                   ind = NA,
+                                   time = NA,
+                                   coefhat = NA,
+                                   tstat = NA))
+      df_sums2 <- rbind(df_sums2, data.frame(y=NA, yadj=NA))
+      
+    }
+    
+    df2$location <- paste0(countylist[i])
+    df_sums2$location <- paste0(countylist[i])
+    df <- bind_rows(df, df2)
+    df_sums <- bind_rows(df_sums, df_sums2)
+  }
+  , error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
+  
+}
+
+df_sums2 <- df_sums %>%
[…]

### 2020-01-09 R/SCRIPTS/analysis-mortality.R

@@ -1,0 +1,95 @@
+###------Mort Loop-----
+## @knitr MortLoop
+
+dat <-
+  ###FERTILITY
+  bind_rows(read_tsv("./R/DATA-RAW/Fertility/Natality, 2003-2006,states.txt"),
+            read_tsv("./R/DATA-RAW/Fertility/Natality, 2007-2017,states.txt"))%>%
+  mutate(Month2 = str_pad(`Month Code`,2 , pad = "0"),
+         Month3 = as.yearmon(paste0(Year,"-", Month2)),
+         location = paste0("X", `State Code`)
+  ) %>%
+  dplyr::select(Year = Month3, location, Target = Births) %>%
+  na.omit %>%
+  arrange(location, Year)
+
+dat <- 
+  ###MORTALITY 
+  # read_tsv("DATA/Mortality/Underlying Cause of Death, 1999-2017,AL-CT.txt") %>%
+  read_tsv("DATA/Mortality/Underlying Cause of Death, 1999-2017,states.txt") %>%
+  mutate(Month2 = substr(`Month Code`,6,7),
+         Month3 = as.yearmon(paste0(Year,"-", Month2)),
+         # location = paste0("X", `County Code`)
+         location = paste0("X", `State Code`)
+  ) %>%
+  dplyr::select(Year = Month3, location, Target = Deaths) %>%
+  na.omit %>%
+  arrange(location, Year)
+
+countylist = unique(dat$location)
+
+df <- data.frame()
+df_sums <- data.frame()
+sigma <- 3.5
+
+for(i in 1:length(unique(countylist))){
+  tryCatch({
+    df2<- setNames(data.frame(matrix(ncol = 5, nrow = 0)), c("type", "ind", "time", "coefhat", "tstat"))
+    dat2 <- filter(dat, location == paste0(countylist[i]))
+    dat3 <- ts(dat2$Target, start = year(min(dat$Year)), end = c(year(max(dat$Year)),12), frequency = 12)
+    outlier.county <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)
+    
+    df_sums2 <- data.frame(y = NA, yadj=NA)
+    if(!is.null(outlier.county$times)){
+      
+      df2<- outlier.county$outliers
+      df_sums2$y <- sum(outlier.county$y)
+      df_sums2$yadj <- sum(outlier.county$yadj)
+    } else {
+      df2 <- rbind(df2, data.frame(type = NA,
[…]

### 2020-01-10 MANUSCRIPT/manuscript.tex
Update manuscript.tex

@@
 \begin{abstract}
+\noindent%
 The proliferation of data, modern computing advances, and powerful
 statistical algorithms unlock the potential search for hidden,
-previously un- or understudied demographic anomalies. Here, we
-investigate US state-level fertility and mortality time series since
+previously un- or understudied demographic anomalies. Here, we demonstrate how demographers can use causal inference techniques to identify anamolies by 
+investigating US state-level fertility and mortality time series since
 1999 to uncover the hidden baby booms/busts and mortality
@@
 results suggest the widespread detection of demographic anomalies. Our
-analysis does not examine the \emph{causes} of these anomalies and our
+analysis does not examine the \emph{causes} of these anomalies but our
 results point to important further research on the causes of anomalous
@@
 \noindent%
-{\it Keywords:} mortality, fertility, causal inference
+{\it Keywords:} mortality, fertility, causal inference, population trends
 \vfill
@@
 
+Causal inference, in its most common usage, is an attempt to undercover the underlying mechanism that results in changes in a phenomenon.
 Causal inference approaches in social sciences have a long history
@@
 expensive RCTs usually preclude their widespread adoption
-\citep{west2008ajph}. Using natural experiments, however, provides
-important and more widespread insights on the determinants of migration
+\citep{west2008ajph}. Natural experiments provides less certainty about causation and fewer opportunities to replicate work; natural experiments provide
+important and widespread insights such as the determinants of migration
 after Hurricane Katrina in 2006
@@
 Maria in Puerto Rico in 2016
-\citep{kishore2018mortality, santos2018use}. Both RCTs and natural
-experiments follow traditional, hypothesis testing, inductive scientific
+\citep{kishore2018mortality, santos2018use}.
+
+Both RCTs and natural
+experiments follow traditional hypothesis testing--inductive scientific
 paradigms where a question is first posed and scientists generate or
@@
 might go unnoticed.
+Data scientists are rethinking these approaches and have re-envisioned causal inference as a machine learning technique that allows big data researchers to move from correlations that have been established with good accuracy to making strong conclusions about the underlying mechanisms that produce these correlations
+\citep{Pearl2018}.
 

### 2020-01-10 MANUSCRIPT/manuscript.Rmd
More edits/updates.

@@
   The proliferation of data, modern computing advances, and powerful statistical algorithms unlock the potential search for hidden, previously un- or understudied demographic anomalies.
-Here, we investigate US state-level fertility and mortality time series since 1999 to uncover the hidden baby booms/busts and mortality plagues?/non-plauges?.
+Here, we demonstrate how demographers can use causal inference techniques to identify anamolies by investigating US state-level fertility and mortality time series since 1999 to uncover the hidden baby booms/busts and mortality plagues?/non-plauges?.
 We find 38 states exhibited at least one mortality anomaly, totalling more than 318k anomalous deaths and an additional 6 states exhibited at least one mortality non-plague?
@@
 read_chunk('../R/SCRIPTS/000-Libraries.R')
+read_chunk('../R/SCRIPTS/001-DataLoad.R')
 ```
@@
 ```{r libraries, include=FALSE}
+```
+
+```{r DataLoad, include=FALSE}
 ```
@@
 
-Causal inference approaches in social sciences have a long history [@grimmer2015ppsp].
+Causal inference, in its most common usage, is an attempt to uncover the underlying mechanism that results in changes in a phenomenon and have a long history approaches in social sciences [@grimmer2015ppsp].
 Often, the identification of a casual mechanism requires either a randomized control trial (RCT) or expert knowledge applied to a natural experiment.
 For population-level analysis, expensive RCTs usually preclude their widespread adoption [@west2008ajph].
-Using natural experiments, however, provides important and more widespread insights on the determinants of migration after Hurricane Katrina in 2006 [@fussellRecoveryMigrationCity2014;@horiDisplacementDynamicsSouthern2009], mini baby booms after electrical blackouts [@fetzer2018jpe], and the highly publicized estimates of excess mortality after Hurricane Maria in Puerto Rico in 2016 [@kishore2018mortality;@santos2018use].
+Using natural experiments provide less certainty about causation and fewer opportunities to replicate work but they provide important and widespread insights such as the determinants of migration, however, provides important and more widespread insights on the determinants of migration after Hurricane Katrina in 2006 [@fussellRecoveryMigrationCity2014;@horiDisplacementDynamicsSouthern2009], mini baby booms after electrical blackouts [@fetzer2018jpe], and the highly publicized estimates of excess mortality after Hurricane Maria in Puerto Rico in 2016 [@kishore2018mortality;@santos2018use].
 Both RCTs and natural experiments follow traditional, hypothesis testing, inductive scientific paradigms where a question is first posed and scientists generate or find data to answer the question.
 Such approaches, while extensively utilized and time-tested, are likely to miss important phenomena that might go unnoticed.
+
+Both RCTs and natural experiments follow traditional hypothesis testing--inductive scientific paradigms where a question is first posed and scientists generate or find data to answer the question.
+Such approaches, while extensively utilized and time-tested, are likely to miss important phenomena that might go unnoticed.
+Data scientists are rethinking these approaches and have re-envisioned causal inference as a machine learning technique that allows big data researchers to move from correlations that have been established with good accuracy to making strong conclusions about the underlying mechanisms that produce these correlations
+[@pearl2018book].
 
@@
 Based on Chien and Liu’s [-@chen1993joint] recommendation, we set a critical value of 3.5 which generally minimizes the possibility of Type I errors or false positive outliers.
-
-Outliers are then reported using a simple t-statistic.
+Outliers are reported using a simple t-statistic.
 
@@
 ## Mortality 
-
-We begin with mortality for New York State (Figure 1).
+We begin with mortality for New York State (\autoref{fig:mortnewyork}).
 We identify seven anomalies in the mortality time series for New York, all with t-statistics in excess of 3.91, making these anomalies highly significant.
 The algorithm correctly identifies September 2001 as an additive outlier (2001:09 t=6.40) where mortality in that month was 1,628 higher than anticipated.
-Clearly, this mortality event is likely caused by the September 11 tragedy and the detection of this mortality event provides confidence in our detection of other anomalies.
+This mortality event is likely caused by the September 11 tragedy and the detection of this mortality event provides confidence in our detection of other anomalies.
+
+```{r MortalityNewYork, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly Detection for New York state mortality, 1999-2016.} The top panel contains the original time series (light gray) and the corrected, counter-factual time series in the absence of anomalies (blue).
+The red dots on the top panel correspond to the onset of detected anomalies.
+The bottom panel contains the magnitude and type of the outlier.
+We detect all three types of outliers in New York mortality data during period 1999-2016.
+We detect additive outliers (AO) in January 2000, September 2001, January 2005, and January 2013; temporary change (TC) outliers in February 2007 and January 2015; and a level shift (LS) starting in February 2004.
+Outliers range in significance from a low t-statistic of 3.91 in the TC 2015:01 to a high t-statistic of 6.40 in the AO outlier in 2001:09.
+\\label{fig:mortnewyork}")}
+
+# df2<- setNames(data.frame(matrix(ncol = 5, nrow = 0)), c("type", "ind", "time", "coefhat", "tstat"))
+dat2 <- filter(dat_mort, location == "X36")
+dat3 <- ts(dat2$Target, start = year(min(dat_mort$Year)), end = c(year(max(dat_mort$Year)),12), frequency = 12)
+outlier.county <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)
+
+plot(outlier.county,
+     cex.axis=0.8) +
+title(main = "New York - Monthly Mortality",
+      adj = 1)
+```
+
+In **\autoref{fig:mortnewyork}**, notice the strong level shift (LS) that occurs in February 2004 (2004:02 t=5.59) which prevented almost 870 deaths per month.
+This shift totals more than 144,000 averted deaths compared to the counter-factual time series and is the single largest mortality protective anomaly among all states.
+This translates to 4.5% fewer deaths than expected over the time period.
+What is driving this mortality protection?
+What policies did NY put into place that might have contributed to this considerable mortality reduction?
+What environmental conditions may have changed?
+These are the kinds of questions that arise from our analyses.
+
+Contrast the mortality protection in New York with the enhanced mortality in New Hampshire (**\autoref{fig:mortnewhamp}**).
+Here we detect two highly significant level shifts (LS) in the monthly mortality data, first in April 2010 and again in November 2014 (2010:04 t=3.51; 2014:11 t=6.68).
+These anomalies suggest New Hampshire experienced 9,700 more deaths (+14% more than expected) in a seven-year period beginning in early 2010.
+This is the single largest percentage mortality increase/decrease we detected among all states.
+Not coincidentally, NH has the second highest opioid-related mortality in the US [@beetham2019access] and it is reasonable that we detect this epidemic in our results.
+
+```{r MortalityNewHamp, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly Detection for New Hampshire state mortality, 1999-2016.} Here we detect two outliers, both level shifts in outliers (LS) in April 2010 and again in November 2014.
+These anomalies suggest New Hampshire experienced approximately 9,700 more deaths than expected since 2010 or 14% more deaths in the state over just seven years.
+\\label{fig:mortnewhamp}")}
+
+# df2<- setNames(data.frame(matrix(ncol = 5, nrow = 0)), c("type", "ind", "time", "coefhat", "tstat"))
+dat2 <- filter(dat_mort, location == "X33")
+dat3 <- ts(dat2$Target, start = year(min(dat_mort$Year)), end = c(year(max(dat_mort$Year)),12), frequency = 12)
+outlier.county <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)
+
+plot(outlier.county,
+     cex.axis=0.8) +
+title(main = "New Hampshire - Monthly Mortality",
+      adj = 1)
+```
+
+In the case of New York and New Hampshire, some of the demographic anomalies have plausible explanations.
+It seems likely that New York’s AO anomaly in September 2001 is caused by the 9/11 tragedy and the rise in New Hampshire’s mortality starting in 2010 could be linked to the opioid epidemic.
+However, we detect numerous seemingly unexplainable demographic anomalies in other states.
+Figure 3 shows two such unexplained anomalies.
 

### 2020-01-10 MANUSCRIPT/manuscript.tex
More edits/updates.

--- 2020-01-09 MANUSCRIPT/manuscript.tex
@@
 previously un- or understudied demographic anomalies. Here, we
-investigate US state-level fertility and mortality time series since
-1999 to uncover the hidden baby booms/busts and mortality
-plagues?/non-plauges?. We find 38 states exhibited at least one
-mortality anomaly, totalling more than 318k anomalous deaths and an
-additional 6 states exhibited at least one mortality non-plague?
+demonstrate how demographers can use causal inference techniques to
+identify anamolies by investigating US state-level fertility and
+mortality time series since 1999 to uncover the hidden baby booms/busts
+and mortality plagues?/non-plauges?. We find 38 states exhibited at
+least one mortality anomaly, totalling more than 318k anomalous deaths
+and an additional 6 states exhibited at least one mortality non-plague?
 totalling 164k protective deaths. 12 states exhibited baby busts,
@@
 
-Causal inference approaches in social sciences have a long history
-\citep{grimmer2015ppsp}. Often, the identification of a casual mechanism
-requires either a randomized control trial (RCT) or expert knowledge
-applied to a natural experiment. For population-level analysis,
-expensive RCTs usually preclude their widespread adoption
-\citep{west2008ajph}. Using natural experiments, however, provides
-important and more widespread insights on the determinants of migration
-after Hurricane Katrina in 2006
+Causal inference, in its most common usage, is an attempt to uncover the
+underlying mechanism that results in changes in a phenomenon and have a
+long history approaches in social sciences \citep{grimmer2015ppsp}.
+Often, the identification of a casual mechanism requires either a
+randomized control trial (RCT) or expert knowledge applied to a natural
+experiment. For population-level analysis, expensive RCTs usually
+preclude their widespread adoption \citep{west2008ajph}. Using natural
+experiments provide less certainty about causation and fewer
+opportunities to replicate work but they provide important and
+widespread insights such as the determinants of migration, however,
+provides important and more widespread insights on the determinants of
+migration after Hurricane Katrina in 2006
 \citep{fussellRecoveryMigrationCity2014, horiDisplacementDynamicsSouthern2009},
@@
 might go unnoticed.
+
+Both RCTs and natural experiments follow traditional hypothesis
+testing--inductive scientific paradigms where a question is first posed
+and scientists generate or find data to answer the question. Such
+approaches, while extensively utilized and time-tested, are likely to
+miss important phenomena that might go unnoticed. Data scientists are
+rethinking these approaches and have re-envisioned causal inference as a
+machine learning technique that allows big data researchers to move from
+correlations that have been established with good accuracy to making
+strong conclusions about the underlying mechanisms that produce these
+correlations \citep{pearl2018book}.
 
@@
 3.5 which generally minimizes the possibility of Type I errors or false
-positive outliers.
-
-Outliers are then reported using a simple t-statistic.
+positive outliers. Outliers are reported using a simple t-statistic.
 
@@
 state-level since 1999. A full listing of these anomalies can be found
-in the \textbf{Supplementary Materials}. Here, we highlight highly
-significant mortality and fertility events across all three types of
-outliers (Additive Outliers, Level Shift Outliers, and Temporary Change
-Outliers) where we have identified plausible explanations for these
-outliers.
+in the \textbf{Supplementary Materials}. We begin with a summary of the
+anomalies we detect and then we highlight highly significant mortality
+and fertility anomalies across all three types of outliers (Additive
+Outliers, Level Shift Outliers, and Temporary Change Outliers) with
+plausible explanations.
+
+\hypertarget{overall-anomalies}{%
+\subsection{Overall Anomalies}\label{overall-anomalies}}
 
@@
 
-We begin with mortality for New York State (Figure 1). We identify seven
-anomalies in the mortality time series for New York, all with
-t-statistics in excess of 3.91, making these anomalies highly
+We begin with mortality for New York State (\autoref{fig:mortnewyork}).
+We identify seven anomalies in the mortality time series for New York,
+all with t-statistics in excess of 3.91, making these anomalies highly
 significant. The algorithm correctly identifies September 2001 as an
 additive outlier (2001:09 t=6.40) where mortality in that month was
-1,628 higher than anticipated. Clearly, this mortality event is likely
-caused by the September 11 tragedy and the detection of this mortality
-event provides confidence in our detection of other anomalies.
+1,628 higher than anticipated. This mortality event is likely caused by
+the September 11 tragedy and the detection of this mortality event
+provides confidence in our detection of other anomalies.
+
+\begin{figure}
+\centering
+\includegraphics{manuscript_files/figure-latex/MortalityNewYork-1.pdf}
+\caption{\textbf{Anomaly Detection for New York state mortality, 1999-2016.}
+The top panel contains the original time series (light gray) and the
+corrected, counter-factual time series in the absence of anomalies
+(blue). The red dots on the top panel correspond to the onset of
+detected anomalies. The bottom panel contains the magnitude and type of
+the outlier. We detect all three types of outliers in New York mortality
+data during period 1999-2016. We detect additive outliers (AO) in
+January 2000, September 2001, January 2005, and January 2013; temporary
+change (TC) outliers in February 2007 and January 2015; and a level
+shift (LS) starting in February 2004. Outliers range in significance
+from a low t-statistic of 3.91 in the TC 2015:01 to a high t-statistic
+of 6.40 in the AO outlier in 2001:09. \label{fig:mortnewyork}}
+\end{figure}
+
+\begin{verbatim}
+## numeric(0)
+\end{verbatim}
+
+In \textbf{\autoref{fig:mortnewyork}}, notice the strong level shift
+(LS) that occurs in February 2004 (2004:02 t=5.59) which prevented
+almost 870 deaths per month. This shift totals more than 144,000 averted
+deaths compared to the counter-factual time series and is the single
+largest mortality protective anomaly among all states. This translates
+to 4.5\% fewer deaths than expected over the time period. What is
+driving this mortality protection? What policies did NY put into place
+that might have contributed to this considerable mortality reduction?
+What environmental conditions may have changed? These are the kinds of
+questions that arise from our analyses.
+
+Contrast the mortality protection in New York with the enhanced
+mortality in New Hampshire (\textbf{\autoref{fig:mortnewhamp}}). Here we
+detect two highly significant level shifts (LS) in the monthly mortality
+data, first in April 2010 and again in November 2014 (2010:04 t=3.51;
+2014:11 t=6.68). These anomalies suggest New Hampshire experienced 9,700
+more deaths (+14\% more than expected) in a seven-year period beginning
+in early 2010. This is the single largest percentage mortality
+increase/decrease we detected among all states. Not coincidentally, NH
+has the second highest opioid-related mortality in the US
+\citep{beetham2019access} and it is reasonable that we detect this
+epidemic in our results.
+
+\begin{figure}
+\centering
+\includegraphics{manuscript_files/figure-latex/MortalityNewHamp-1.pdf}
+\caption{\textbf{Anomaly Detection for New Hampshire state mortality, 1999-2016.}
+Here we detect two outliers, both level shifts in outliers (LS) in April
+2010 and again in November 2014. These anomalies suggest New Hampshire
+experienced approximately 9,700 more deaths than expected since 2010 or
+14\% more deaths in the state over just seven years.
+\label{fig:mortnewhamp}}
+\end{figure}
+
+\begin{verbatim}
+## numeric(0)
+\end{verbatim}
+
+In the case of New York and New Hampshire, some of the demographic
+anomalies have plausible explanations. It seems likely that New York's
+AO anomaly in September 2001 is caused by the 9/11 tragedy and the rise
+in New Hampshire's mortality starting in 2010 could be linked to the
+opioid epidemic. However, we detect numerous seemingly unexplainable
+demographic anomalies in other states. Figure 3 shows two such
+unexplained anomalies.
 

### 2020-01-10 R/SCRIPTS/000-Libraries.R

@@ -87,3 +87,5 @@ registerDoParallel(clusters) # Register Cluster
 getDoParWorkers() # Determine Number of Utilized Clusters
 getDoParName() #  Name of the Currently Registered Parallel Computing Backend
 getDoParVersion() #  Version of the Currently Registered Parallel Computing Backend
+
+sigma <- 3.5

### 2020-01-10 R/SCRIPTS/001-AnomaliesDetection.R

@@ -20,28 +20,49 @@ df_sums <- data.frame()
 sigma <- 3.5
 
 for(i in 1:length(unique(countylist))){
+
   tryCatch({
     df2<- setNames(data.frame(matrix(ncol = 5, nrow = 0)), c("type", "ind", "time", "coefhat", "tstat"))
     dat2 <- filter(dat, location == paste0(countylist[i]))
     dat3 <- ts(dat2$Target, start = year(min(dat$Year)), end = c(year(max(dat$Year)),12), frequency = 12)
     outlier.county <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)
     
+    # df_sums2 <- data.frame(type = NA, tot = NA, abstot = NA, location = NA)
     df_sums2 <- data.frame(y = NA, yadj=NA)
     if(!is.null(outlier.county$times)){
       
       df2<- outlier.county$outliers
+      # df_sums2 <- data.frame(y=as.matrix(outlier.county$y), 
+      #                 yadj = as.matrix(outlier.county$yadj),
+      #                 month = as.Date(time(outlier.county$y))) %>%
+      #   mutate(ind = as.numeric(rownames(.)),
+      #          diff = y - yadj) %>%
+      #   left_join(., outlier.county$outliers) %>%
+      #   filter(diff != 0) %>%
+      #   fill(type) 
+        # group_by(type) %>%
+        # dplyr::summarise(tot = sum(diff),
+        #                  abstot = sum(abs(diff)))
+      
+      # df_sums2$location <- paste0(countylist[i])
       df_sums2$y <- sum(outlier.county$y)
       df_sums2$yadj <- sum(outlier.county$yadj)
+      
     } else {
       df2 <- rbind(df2, data.frame(type = NA,
                                    ind = NA,
                                    time = NA,
                                    coefhat = NA,
                                    tstat = NA))
+      # df_sums2 <- rbind(df_sums2)
       df_sums2 <- rbind(df_sums2, data.frame(y=NA, yadj=NA))
       
     }
     
+    # df2$location <- paste0(countylist[i])
+    # df_sums2$location <- paste0(countylist[i])
+    # df <- bind_rows(df, df2)
+    # df_sums <- bind_rows(df_sums, df_sums2)
     df2$location <- paste0(countylist[i])
     df_sums2$location <- paste0(countylist[i])
[…]

### 2020-01-10 R/SCRIPTS/001-DataLoad.R

@@ -1,0 +1,40 @@
+###------Data Load -----
+## @knitr DataLoad
+
+dat_fert <-
+  ###FERTILITY
+  bind_rows(read_tsv("../R/DATA-RAW/Fertility/Natality, 2003-2006,states.txt"),
+            read_tsv("../R/DATA-RAW/Fertility/Natality, 2007-2017,states.txt"))%>%
+  mutate(Month2 = str_pad(`Month Code`,2 , pad = "0"),
+         Month3 = as.yearmon(paste0(Year,"-", Month2)),
+         location = paste0("X", `State Code`)
+  ) %>%
+  dplyr::select(Year = Month3, location, Target = Births) %>%
+  na.omit %>%
+  arrange(location, Year)
+
+dat_mort <- 
+  ###MORTALITY 
+  # read_tsv("DATA/Mortality/Underlying Cause of Death, 1999-2017,AL-CT.txt") %>%
+  read_tsv("../R/DATA-RAW/Mortality/Underlying Cause of Death, 1999-2017,states.txt") %>%
+  mutate(Month2 = substr(`Month Code`,6,7),
+         Month3 = as.yearmon(paste0(Year,"-", Month2)),
+         # location = paste0("X", `County Code`)
+         location = paste0("X", `State Code`)
+  ) %>%
+  dplyr::select(Year = Month3, location, Target = Deaths) %>%
+  na.omit %>%
+  arrange(location, Year)
+
+
+
+# df2<- setNames(data.frame(matrix(ncol = 5, nrow = 0)), c("type", "ind", "time", "coefhat", "tstat"))
+# dat2 <- filter(dat_mort, location == "X36")
+# dat3 <- ts(dat2$Target, start = year(min(dat_mort$Year)), end = c(year(max(dat_mort$Year)),12), frequency = 12)
+# outlier.county <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)
+# 
+# plot(outlier.county,
+#      args.lines.yadj = list(col = "black"),
+#      args.lines.y = list(col = "gray80")) +
+# title(main = "New York - Monthly Deaths")
+

### 2020-01-14 MANUSCRIPT/manuscript.Rmd
Working through figures and associated text for the Main Document.

@@
 read_chunk('../R/SCRIPTS/001-DataLoad.R')
+read_chunk('../R/SCRIPTS/002-StatesAnalysis.R')
 ```
@@
 ```{r DataLoad, include=FALSE}
+```
+
+```{r StatesAnalysis, include=FALSE}
 ```
@@
 In situations with large data sets, an abductive approach is far superior to deductive hypothesis testing, as p-values with an extremely large number of cases are far from revealing [@head2015extent;@nuzzo2014scientific] and we do not want to make purely inductive inferences from data of questionable generalizability [@ruggles2014big].
-Owing to the long history of big datasets in demographic research (one of the original "big data" resources [@ruggles2014big]), the rich demographic data available in the United States make the potential revelation of interesting and important demographic phenomena not only possible, but extremely plausible – even if identification of the phenomena occurs without identifying the underlying cause.
+Owing to the long history of big data sets in demographic research (one of the original "big data" resources [@ruggles2014big]), the rich demographic data available in the United States make the potential revelation of interesting and important demographic phenomena not only possible, but extremely plausible – even if identification of the phenomena occurs without identifying the underlying cause.
 In other sciences, the identification of effects without causes has led to new theories of galactic migration of planets and other breakthroughs [@gomes2005n].
@@
 Of importance here is step 2, which relies on some critical value above which an outlier at time point m is considered spurious.
-Based on Chien and Liu’s [-@chen1993joint] recommendation, we set a critical value of 3.5 which generally minimizes the possibility of Type I errors or false positive outliers.
+Based on Chen and Liu’s [-@chen1993joint] recommendation, we set a critical value of 3.5 which generally minimizes the possibility of Type I errors or false positive outliers.
 Outliers are reported using a simple t-statistic.
@@
 
-# df2<- setNames(data.frame(matrix(ncol = 5, nrow = 0)), c("type", "ind", "time", "coefhat", "tstat"))
-dat2 <- filter(dat_mort, location == "X36")
-dat3 <- ts(dat2$Target, start = year(min(dat_mort$Year)), end = c(year(max(dat_mort$Year)),12), frequency = 12)
-outlier.county <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)
-
-plot(outlier.county,
+plot(outlier.mort.ny,
      cex.axis=0.8) +
@@
 
-# df2<- setNames(data.frame(matrix(ncol = 5, nrow = 0)), c("type", "ind", "time", "coefhat", "tstat"))
-dat2 <- filter(dat_mort, location == "X33")
-dat3 <- ts(dat2$Target, start = year(min(dat_mort$Year)), end = c(year(max(dat_mort$Year)),12), frequency = 12)
-outlier.county <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)
-
-plot(outlier.county,
+plot(outlier.mort.nh,
      cex.axis=0.8) +
@@
 
+## Fertility
+
+We begin with a clear temporary change (TC) outlier for fertility in Louisiana (**\autoref{fig:fertla}**).
+Here we detect two TC outliers, nearly back-to-back in 2005 (2005:08 *t* = `r prettyNum(outlier.fert.la$outliers$tstat[1], digits = 3)`; and 2005:10 *t* = `r prettyNum(outlier.fert.la$outliers$tstat[2], digits = 3)`), coinciding with the destruction of Hurricane Katrina from the same year.
+The migration associated with Hurricane Katrina has garnered most of the attention of demographers [@fussellRecoveryMigrationCity2014;@horiDisplacementDynamicsSouthern2009], but the hurricane had a very clear impact on fertility behaviors too, creating a mini "baby bust" in Louisiana.
+We estimate `r prettyNum(abs(sum(outlier.fert.la$y - outlier.fert.la$yadj)), digits = 0, big.mark = ",")` fewer births in Louisiana compared to the counterfactual, likely attributable to the Hurricane.
+
+```{r FertilityLouisiana, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Louisiana state fertility, 2003-2018.} Here we detect two outliers, back-to-back, likely resulting from Hurricane Katrina August and October 2005, representing a decrease of more than 4,400 births due to the hurricane.
+\\label{fig:fertla}")}
+
+plot(outlier.fert.la,
+     cex.axis=0.8) +
+  title(main = "Louisiana - Monthly Fertility",
+        adj = 1)
+```
+
+We also highlight a level shift (LS) outlier for fertility in Connecticut.
+Here we detect a strong (*t*= `prettyNum(outlier.fert.conn$outliers$tstat, digits = 3)`; `r prettyNum(outlier.fert.conn$outliers$coefhat, digits = 0)` births/month) level shift toward lower fertility beginning in August 2009 (2009:08) that continues through the end of the period (**\autoref{fig:fertconn}**).
+This LS toward lower fertility reduced the # of births by `r prettyNum(sum(outlier.fert.conn$y - outlier.fert.conn$yadj), digits = 0, big.mark = ",")` or `r percent(sum(outlier.fert.conn$y - outlier.fert.conn$yadj) / sum(outlier.fert.conn$yadj[80:180]))` lower than the counterfactual time series.
+The stock market crash of 2008, where the Dow Jones Industrial Average had the largest single-day loss up to that point, occured just 11 months before we detect a shift toward lower fertility.
+We believe the trend toward lower fertility in August 2009 and the stock market crash in September 2008 are linked.
+
+```{r FertilityConnecticut, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Hawaiian state fertility, 2003-2018.} Here we detect one outlier, an additive outlier (AO) in May 2014 ( \textit{t}=4.96), representing a 14% increase in expected births in that month.
+\\label{fig:fertconn}")}
+
+plot(outlier.fert.conn,
+     cex.axis=0.8) +
+  title(main = "Connecticut - Monthly Fertility",
+        adj = 1)
+```
+
+## Interesting Anomalous Fertility/Mortality events
+In the examples above, we highlighted four fertility/mortality anomalies with plausible explanations.
+However, these four anomalies are but a fraction of anomalies we detect.
+Here, we highlight four additional anomalies without easily identifiable explanations.
+
+In **\autoref{fig:ferthawaii}** we identify a single additive anomaly in Hawaiian fertility in May 2014.
+This is a strong anomaly with a *t*-statistic of 4.96, 14% above the counter-factual time series.
+This single, anomalous month is also the second highest monthly births in the time series.
+We have no plausible explanation for this anomaly.
+We do not believe this is simply a data error as the other extreme values, September 2008 and February 2005 with the highest and lowest recorded fertility respectively, were not identified as anomalous events.
+Even if we were to assume this anomaly resulted from data entry error, it remains an *unaltered* data error in the Hawaiian monthly fertility data.
+
+```{r AnomalyHawaii, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Hawaiian state fertility, 2003-2018.} Here we detect one outlier, an additive outlier (AO) in May 2014 ( \textit{t}=4.96), representing a 14% increase in expected births in that month.
+\\label{fig:ferthawaii}")}
+
+plot(outlier.fert.hi,
+     cex.axis=0.8) +
+  title(main = "Hawaii - Monthly Fertility",
+        adj = 1)
+```
+
+We detect a very interesting, unexplainable mortality anomaly in North Carolina (**\autoref{fig:fertmiss}**).
+Beginning in February 2008 , North Carolina saw a level shift (LS) (2008:02; *t*-stat = `r prettyNum(outlier.mort.nc$outliers$tstat[3], digits = 3)`) of 800 fewer deaths per month until the end of the period.
+Granted, the decade after the LS still saw three mortality anomalies that resulted in additional, unexpected deaths in 2009, 2012, and 2017, the LS is still a remarkable mortality reduction of nearly 100k (`r prettyNum(sum(outlier.mort.nc$y[62:180] - outlier.mort.nc$yadj[62:180]), digits = 0, big.mark = ",")`) deaths over 10 years!
+While North Carolina saw the raw count of mortality remain constant or slightly reduce since 2008, the state grew from 9.3 million residents in 2008 to 10.4 million in 2018.
+In spite of its growing population, North Carolina has, rather miraculously, *reduced* its overall mortality risk.
+```{r AnomalyMississippi, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in North Carolina state mortality, 1999-2018.} Here we detect one outlier, an additive outlier (AO) in May 2014 ( \textit{t}=4.96), representing a 14% increase in expected births in that month.
+\\label{fig:fertmiss}")}
+
+plot(outlier.mort.nc,
+     cex.axis=0.8) +
+  title(main = "North Carolina - Monthly Mortality",
+        adj = 1)
+```
+
+Hawaii - AO
+X33 - LS
+X28
+X22
+X20
+X05
+
 \newpage

### 2020-01-14 MANUSCRIPT/manuscript.tex
Working through figures and associated text for the Main Document.

@@
 purely inductive inferences from data of questionable generalizability
-\citep{ruggles2014big}. Owing to the long history of big datasets in
+\citep{ruggles2014big}. Owing to the long history of big data sets in
 demographic research (one of the original ``big data'' resources
@@
 which relies on some critical value above which an outlier at time point
-m is considered spurious. Based on Chien and Liu's
+m is considered spurious. Based on Chen and Liu's
 \citeyearpar{chen1993joint} recommendation, we set a critical value of
@@
 
+\hypertarget{fertility}{%
+\subsection{Fertility}\label{fertility}}
+
+We begin with a clear temporary change (TC) outlier for fertility in
+Louisiana (\textbf{\autoref{fig:fertla}}). Here we detect two TC
+outliers, nearly back-to-back in 2005 (2005:08 \emph{t} = -6; and
+2005:10 \emph{t} = -5), coinciding with the destruction of Hurricane
+Katrina from the same year. The migration associated with Hurricane
+Katrina has garnered most of the attention of demographers
+\citep{fussellRecoveryMigrationCity2014, horiDisplacementDynamicsSouthern2009},
+but the hurricane had a very clear impact on fertility behaviors too,
+creating a mini ``baby bust'' in Louisiana. We estimate 4,411 fewer
+births in Louisiana compared to the counterfactual, likely attributable
+to the Hurricane.
+
+\begin{figure}
+\centering
+\includegraphics{manuscript_files/figure-latex/FertilityLouisiana-1.pdf}
+\caption{\textbf{Anomaly detection in Louisiana state fertility, 2003-2018.}
+Here we detect two outliers, back-to-back, likely resulting from
+Hurricane Katrina August and October 2005, representing a decrease of
+more than 4,400 births due to the hurricane. \label{fig:fertla}}
+\end{figure}
+
+\begin{verbatim}
+## numeric(0)
+\end{verbatim}
+
+We also highlight a level shift (LS) outlier for fertility in
+Connecticut. Here we detect a strong (\emph{t}=
+\texttt{prettyNum(outlier.fert.conn\$outliers\$tstat,\ digits\ =\ 3)};
+-285 births/month) level shift toward lower fertility beginning in
+August 2009 (2009:08) that continues through the end of the period
+(\textbf{\autoref{fig:fertconn}}). This LS toward lower fertility
+reduced the \# of births by -28,752 or -8.57\% lower than the
+counterfactual time series. The stock market crash of 2008, where the
+Dow Jones Industrial Average had the largest single-day loss up to that
+point, occured just 11 months before we detect a shift toward lower
+fertility. We believe the trend toward lower fertility in August 2009
+and the stock market crash in September 2008 are linked.
+
+\begin{figure}
+\centering
+\includegraphics{manuscript_files/figure-latex/FertilityConnecticut-1.pdf}
+\caption{\textbf{Anomaly detection in Hawaiian state fertility, 2003-2018.}
+Here we detect one outlier, an additive outlier (AO) in May 2014 (
+extit\{t\}=4.96), representing a 14\% increase in expected births in
+that month. \label{fig:fertconn}}
+\end{figure}
+
+\begin{verbatim}
+## numeric(0)
+\end{verbatim}
+
+\hypertarget{interesting-anomalous-fertilitymortality-events}{%
+\subsection{Interesting Anomalous Fertility/Mortality
+events}\label{interesting-anomalous-fertilitymortality-events}}
+
+In the examples above, we highlighted four fertility/mortality anomalies
+with plausible explanations. However, these four anomalies are but a
+fraction of anomalies we detect. Here, we highlight four additional
+anomalies without easily identifiable explanations.
+
+In \textbf{\autoref{fig:ferthawaii}} we identify a single additive
+anomaly in Hawaiian fertility in May 2014. This is a strong anomaly with
+a \emph{t}-statistic of 4.96, 14\% above the counter-factual time
+series. This single, anomalous month is also the second highest monthly
+births in the time series. We have no plausible explanation for this
+anomaly. We do not believe this is simply a data error as the other
+extreme values, September 2008 and February 2005 with the highest and
+lowest recorded fertility respectively, were not identified as anomalous
+events. Even if we were to assume this anomaly resulted from data entry
+error, it remains an \emph{unaltered} data error in the Hawaiian monthly
+fertility data.
+
+\begin{figure}
+\centering
+\includegraphics{manuscript_files/figure-latex/AnomalyHawaii-1.pdf}
+\caption{\textbf{Anomaly detection in Hawaiian state fertility, 2003-2018.}
+Here we detect one outlier, an additive outlier (AO) in May 2014 (
+extit\{t\}=4.96), representing a 14\% increase in expected births in
+that month. \label{fig:ferthawaii}}
+\end{figure}
+
+\begin{verbatim}
+## numeric(0)
+\end{verbatim}
+
+We detect a very interesting, unexplainable mortality anomaly in North
+Carolina (\textbf{\autoref{fig:fertmiss}}). Beginning in February 2008 ,
+North Carolina saw a level shift (LS) (2008:02; \emph{t}-stat = -4.76)
+of 800 fewer deaths per month until the end of the period. Granted, the
+decade after the LS still saw three mortality anomalies that resulted in
+additional, unexpected deaths in 2009, 2012, and 2017, the LS is still a
+remarkable mortality reduction of nearly 100k (-88,335) deaths over 10
+years! While North Carolina saw the raw count of mortality remain
+constant or slightly reduce since 2008, the state grew from 9.3 million
+residents in 2008 to 10.4 million in 2018. In spite of its growing
+population, North Carolina has, rather miraculously, \emph{reduced} its
+overall mortality risk.
+\includegraphics{manuscript_files/figure-latex/AnomalyMississippi-1.pdf}
+
+\begin{verbatim}
+## numeric(0)
+\end{verbatim}
+
+Hawaii - AO X33 - LS X28 X22 X20 X05
+
 \newpage

### 2020-01-14 R/SCRIPTS/000-Libraries.R

@@ -89,3 +89,4 @@ getDoParName() #  Name of the Currently Registered Parallel Computing Backend
 getDoParVersion() #  Version of the Currently Registered Parallel Computing Backend
 
 sigma <- 3.5
+set.seed(1)

### 2020-01-14 R/SCRIPTS/001-AnomaliesDetection.R

@@ -99,6 +99,7 @@ df_sums <- data.frame()
 sigma <- 3.5
 
 for(i in 1:length(unique(countylist))){
+  set.seed(1)
   tryCatch({
     df2<- setNames(data.frame(matrix(ncol = 5, nrow = 0)), c("type", "ind", "time", "coefhat", "tstat"))
     dat2 <- filter(dat, location == paste0(countylist[i]))

### 2020-01-14 R/SCRIPTS/002-StatesAnalysis.R

@@ -1,0 +1,43 @@
+###------States Analysis  -----
+## @knitr StatesAnalysis
+
+# Mortality - New York
+dat2 <- filter(dat_mort, location == "X36")
+dat3 <- ts(dat2$Target, start = year(min(dat_mort$Year)), end = c(year(max(dat_mort$Year)),12), frequency = 12)
+outlier.mort.ny <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)
+
+# Mortality - New Hampshire
+dat2 <- filter(dat_mort, location == "X33")
+dat3 <- ts(dat2$Target, start = year(min(dat_mort$Year)), end = c(year(max(dat_mort$Year)),12), frequency = 12)
+outlier.mort.nh <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)
+
+# Fertility - Connecticut
+dat2 <- filter(dat_fert, location == "X09")
+dat3 <- ts(dat2$Target, start = year(min(dat_fert$Year)), end = c(year(max(dat_fert$Year)),12), frequency = 12)
+outlier.fert.conn <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)
+
+# Fertility - Louisiana
+dat2 <- filter(dat_fert, location == "X22")
+dat3 <- ts(dat2$Target, start = year(min(dat_fert$Year)), end = c(year(max(dat_fert$Year)),12), frequency = 12)
+outlier.fert.la <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)
+
+
+# Fertility - Hawaii
+dat2 <- filter(dat_fert, location == "X15")
+dat3 <- ts(dat2$Target, start = year(min(dat_fert$Year)), end = c(year(max(dat_fert$Year)),12), frequency = 12)
+outlier.fert.hi <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)
+
+# Fertility - Mississippi
+dat2 <- filter(dat_fert, location == "X22")
+dat3 <- ts(dat2$Target, start = year(min(dat_fert$Year)), end = c(year(max(dat_fert$Year)),12), frequency = 12)
[…]

### 2020-01-15 MANUSCRIPT/manuscript.Rmd
Results section mainly written. Still need an overall table/figure.

@@
 
-```{r StatesAnalysis, include=FALSE}
-```
+<!-- # ```{r StatesAnalysis, include=FALSE} -->
+<!-- # ``` -->
 
@@
 We begin with a summary of the anomalies we detect and then we highlight highly significant mortality and fertility anomalies across all three types of outliers (Additive Outliers, Level Shift Outliers, and Temporary Change Outliers) with plausible explanations.
+Finally, we highlight two strong anomalies that bely explanation.
 
@@
 ## Mortality 
-We begin with mortality for New York State (\autoref{fig:mortnewyork}).
+We begin with mortality for New York State (**\autoref{fig:mortnewhamp}a**).
 We identify seven anomalies in the mortality time series for New York, all with t-statistics in excess of 3.91, making these anomalies highly significant.
@@
 
-```{r MortalityNewYork, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly Detection for New York state mortality, 1999-2016.} The top panel contains the original time series (light gray) and the corrected, counter-factual time series in the absence of anomalies (blue).
-The red dots on the top panel correspond to the onset of detected anomalies.
-The bottom panel contains the magnitude and type of the outlier.
-We detect all three types of outliers in New York mortality data during period 1999-2016.
-We detect additive outliers (AO) in January 2000, September 2001, January 2005, and January 2013; temporary change (TC) outliers in February 2007 and January 2015; and a level shift (LS) starting in February 2004.
-Outliers range in significance from a low t-statistic of 3.91 in the TC 2015:01 to a high t-statistic of 6.40 in the AO outlier in 2001:09.
-\\label{fig:mortnewyork}")}
-
-plot(outlier.mort.ny,
-     cex.axis=0.8) +
-title(main = "New York - Monthly Mortality",
-      adj = 1)
-```
-
-In **\autoref{fig:mortnewyork}**, notice the strong level shift (LS) that occurs in February 2004 (2004:02 t=5.59) which prevented almost 870 deaths per month.
+In **\autoref{fig:mortnewhamp}a**, notice the strong level shift (LS) that occurs in February 2004 (2004:02 t=5.59) which prevented almost 870 deaths per month.
 This shift totals more than 144,000 averted deaths compared to the counter-factual time series and is the single largest mortality protective anomaly among all states.
@@
 
-Contrast the mortality protection in New York with the enhanced mortality in New Hampshire (**\autoref{fig:mortnewhamp}**).
+Contrast the mortality protection in New York with the enhanced mortality in New Hampshire (**\autoref{fig:mortnewhamp}b**).
 Here we detect two highly significant level shifts (LS) in the monthly mortality data, first in April 2010 and again in November 2014 (2010:04 t=3.51; 2014:11 t=6.68).
@@
 
-```{r MortalityNewHamp, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly Detection for New Hampshire state mortality, 1999-2016.} Here we detect two outliers, both level shifts in outliers (LS) in April 2010 and again in November 2014.
-These anomalies suggest New Hampshire experienced approximately 9,700 more deaths than expected since 2010 or 14% more deaths in the state over just seven years.
+```{r MortalityNewHamp, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly Detection for New York (a) and New Hampshire (b) state mortality, 1999-2016.} The top part of each panel contains the original time series (light gray), the corrected, counter-factual time series in the absence of anomalies (blue), and the red dots correspond to the onset of detected anomalies.
+The bottom part of each panel contains the magnitude and type of the outlier in red.
+In New York (a), we detect additive outliers (AO) in January 2000, September 2001, January 2005, and January 2013; temporary change (TC) outliers in February 2007 and January 2015; and a level shift (LS) starting in February 2004.
+In New Hampshire (b), we detect two outliers, both level shift outliers (LS) in April 2010 and again in November 2014.
+These anomalies suggest New York experienced a significant mortality event in September 2001 and New Hampshire experienced approximately 9,700 more deaths than expected since 2010 or 14% more deaths in the state over just seven years.
 \\label{fig:mortnewhamp}")}
 
-plot(outlier.mort.nh,
-     cex.axis=0.8) +
-title(main = "New Hampshire - Monthly Mortality",
-      adj = 1)
+a <-  function() {
+plot(outlier.mort.ny, cex = 0.8)
+title("New York - Mortality", adj = 0.9, line =2)
+}
+
+b <-  function() {
+  plot(outlier.mort.nh, cex = 0.8)
+  title("New Hampshire - Mortality", adj = 0.82,line =2)
+}
+
+plot_grid(ggdraw(a) ,
+               ggdraw(b), 
+               ncol=1,
+               labels= "auto")
 ```
-
-In the case of New York and New Hampshire, some of the demographic anomalies have plausible explanations.
-It seems likely that New York’s AO anomaly in September 2001 is caused by the 9/11 tragedy and the rise in New Hampshire’s mortality starting in 2010 could be linked to the opioid epidemic.
-However, we detect numerous seemingly unexplainable demographic anomalies in other states.
-Figure 3 shows two such unexplained anomalies.
 
@@
 
-We begin with a clear temporary change (TC) outlier for fertility in Louisiana (**\autoref{fig:fertla}**).
+We begin with a clear temporary change (TC) outlier for fertility in Louisiana (**\autoref{fig:fertla}a**).
 Here we detect two TC outliers, nearly back-to-back in 2005 (2005:08 *t* = `r prettyNum(outlier.fert.la$outliers$tstat[1], digits = 3)`; and 2005:10 *t* = `r prettyNum(outlier.fert.la$outliers$tstat[2], digits = 3)`), coinciding with the destruction of Hurricane Katrina from the same year.
@@
 
-```{r FertilityLouisiana, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Louisiana state fertility, 2003-2018.} Here we detect two outliers, back-to-back, likely resulting from Hurricane Katrina August and October 2005, representing a decrease of more than 4,400 births due to the hurricane.
-\\label{fig:fertla}")}
-
-plot(outlier.fert.la,
-     cex.axis=0.8) +
-  title(main = "Louisiana - Monthly Fertility",
-        adj = 1)
-```
-
 We also highlight a level shift (LS) outlier for fertility in Connecticut.
-Here we detect a strong (*t*= `prettyNum(outlier.fert.conn$outliers$tstat, digits = 3)`; `r prettyNum(outlier.fert.conn$outliers$coefhat, digits = 0)` births/month) level shift toward lower fertility beginning in August 2009 (2009:08) that continues through the end of the period (**\autoref{fig:fertconn}**).
+Here we detect a strong (*t*= `r prettyNum(outlier.fert.conn$outliers$tstat, digits = 3)`; `r prettyNum(outlier.fert.conn$outliers$coefhat, digits = 0)` births/month) level shift toward lower fertility beginning in August 2009 (2009:08) that continues through the end of the period (**\autoref{fig:fertla}b**).
 This LS toward lower fertility reduced the # of births by `r prettyNum(sum(outlier.fert.conn$y - outlier.fert.conn$yadj), digits = 0, big.mark = ",")` or `r percent(sum(outlier.fert.conn$y - outlier.fert.conn$yadj) / sum(outlier.fert.conn$yadj[80:180]))` lower than the counterfactual time series.
@@
 
-```{r FertilityConnecticut, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Hawaiian state fertility, 2003-2018.} Here we detect one outlier, an additive outlier (AO) in May 2014 ( \textit{t}=4.96), representing a 14% increase in expected births in that month.
-\\label{fig:fertconn}")}
+```{r FertilityLouisiana, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Louisiana (a) and Connecticut (b) state fertility, 2003-2018.} In (a), we detect two outliers, back-to-back, likely resulting from Hurricane Katrina in August and October 2005, representing a decrease of more than 4,400 births due to the hurricane.
+In (b), we detect one outlier, a level shift outlier (LS) in August 2009.
+We believe this reduction is attributable to the stock market crash 11 months earlier.
+\\label{fig:fertla}")}
 
-plot(outlier.fert.conn,
-     cex.axis=0.8) +
-  title(main = "Connecticut - Monthly Fertility",
-        adj = 1)
+a <-  function() {
+  plot(outlier.fert.la, cex.axis=0.8)
+  title(main = "Louisiana - Fertility", adj = 0.9)
+}
+
+b <- function(){
+  plot(outlier.fert.conn, cex.axis=0.8) +
+  title(main = "Connecticut - Fertility", adj = 0.88)
+}
+
+plot_grid(ggdraw(a) ,
+               ggdraw(b), 
+               ncol=1,
+               labels= "auto")
+
 ```
@@
 In the examples above, we highlighted four fertility/mortality anomalies with plausible explanations.
-However, these four anomalies are but a fraction of anomalies we detect.
-Here, we highlight four additional anomalies without easily identifiable explanations.
+In the case of New York and New Hampshire, the mortality anomalies have plausible explanations.
+It seems likely that New York’s AO anomaly in September 2001 is caused by the 9/11 tragedy and the rise in New Hampshire’s mortality starting in 2010 could be linked to the opioid epidemic.
+Similarly, Louisiana's TC anomalies seem linked to Hurricanes Katrina and Rita while the LS anomaly in Connecticut's fertility appears linked to the Great Recession.
+However, we detect numerous seemingly unexplainable demographic anomalies in other states.
+**\autoref{fig:ferthawaii}** shows two such unexplained anomalies.
 
-In **\autoref{fig:ferthawaii}** we identify a single additive anomaly in Hawaiian fertility in May 2014.
-This is a strong anomaly with a *t*-statistic of 4.96, 14% above the counter-factual time series.
+In **\autoref{fig:ferthawaii}a** we identify a single additive anomaly in Hawaiian fertility in May 2014.
+This is a strong anomaly with a *t*-statistic of `r prettyNum(outlier.fert.hi$outliers$tstat, digits = 3)`, 14% above the counter-factual time series.
 This single, anomalous month is also the second highest monthly births in the time series.
@@
 
-```{r AnomalyHawaii, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Hawaiian state fertility, 2003-2018.} Here we detect one outlier, an additive outlier (AO) in May 2014 ( \textit{t}=4.96), representing a 14% increase in expected births in that month.
+This is contrast to **\autoref{fig:ferthawaii}b**, where we identify a strange mortality reduction in Ohio (*t*-stat: `r prettyNum(outlier.mort.oh$outliers$tstat[7], digits = 3)`).
+This LS is more than 1,100 deaths per month less than the counterfactual time series, suggesting had nearly 40,000 fewer deaths since February 2015 than expected.
+This is the single-largest LS among all states.
+We could not identify the potential policies Ohio might have put into place to provide such a strong mortality protection.
+
+```{r AnomalyHawaii, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Hawaii state fertility (a) and Ohio state mortality (b).} Here we detect one outlier, an additive outlier (AO) in May 2014 in Hawaii (a), representing a large, unexplainable 14% increase in expected births in that month.
+We also detect a strong level shift (LS) reduction (b), representing a large, unexplainable reduction of nearly 40,000 deaths in Ohio.
 \\label{fig:ferthawaii}")}
 
-plot(outlier.fert.hi,
-     cex.axis=0.8) +
-  title(main = "Hawaii - Monthly Fertility",
-        adj = 1)
+a <-  function() {
+plot(outlier.fert.hi, cex.axis=0.8)
+  title(main = "Hawaii - Fertility", adj = 0.88)
+}
+
+b <-  function() {
+plot(outlier.mort.oh, cex.axis=0.8)
+  title(main = "Ohio - Mortality", adj = 0.9)
+}
+
+plot_grid(ggdraw(a) ,
+               ggdraw(b), 
+               ncol=1,
+               labels= "auto")
 ```
 
-We detect a very interesting, unexplainable mortality anomaly in North Carolina (**\autoref{fig:fertmiss}**).
-Beginning in February 2008 , North Carolina saw a level shift (LS) (2008:02; *t*-stat = `r prettyNum(outlier.mort.nc$outliers$tstat[3], digits = 3)`) of 800 fewer deaths per month until the end of the period.
-Granted, the decade after the LS still saw three mortality anomalies that resulted in additional, unexpected deaths in 2009, 2012, and 2017, the LS is still a remarkable mortality reduction of nearly 100k (`r prettyNum(sum(outlier.mort.nc$y[62:180] - outlier.mort.nc$yadj[62:180]), digits = 0, big.mark = ",")`) deaths over 10 years!
-While North Carolina saw the raw count of mortality remain constant or slightly reduce since 2008, the state grew from 9.3 million residents in 2008 to 10.4 million in 2018.
-In spite of its growing population, North Carolina has, rather miraculously, *reduced* its overall mortality risk.
-```{r AnomalyMississippi, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in North Carolina state mortality, 1999-2018.} Here we detect one outlier, an additive outlier (AO) in May 2014 ( \textit{t}=4.96), representing a 14% increase in expected births in that month.
-\\label{fig:fertmiss}")}
-
-plot(outlier.mort.nc,
-     cex.axis=0.8) +
-  title(main = "North Carolina - Monthly Mortality",
-        adj = 1)
-```
-
-Hawaii - AO
-X33 - LS
-X28
-X22
-X20
-X05
-
 \newpage

### 2020-01-15 MANUSCRIPT/manuscript.tex
Results section mainly written. Still need an overall table/figure.

@@
 Outliers, Level Shift Outliers, and Temporary Change Outliers) with
-plausible explanations.
+plausible explanations. Finally, we highlight two strong anomalies that
+bely explanation.
 
@@
 
-We begin with mortality for New York State (\autoref{fig:mortnewyork}).
-We identify seven anomalies in the mortality time series for New York,
-all with t-statistics in excess of 3.91, making these anomalies highly
-significant. The algorithm correctly identifies September 2001 as an
-additive outlier (2001:09 t=6.40) where mortality in that month was
-1,628 higher than anticipated. This mortality event is likely caused by
-the September 11 tragedy and the detection of this mortality event
-provides confidence in our detection of other anomalies.
+We begin with mortality for New York State
+(\textbf{\autoref{fig:mortnewhamp}a}). We identify seven anomalies in
+the mortality time series for New York, all with t-statistics in excess
+of 3.91, making these anomalies highly significant. The algorithm
+correctly identifies September 2001 as an additive outlier (2001:09
+t=6.40) where mortality in that month was 1,628 higher than anticipated.
+This mortality event is likely caused by the September 11 tragedy and
+the detection of this mortality event provides confidence in our
+detection of other anomalies.
 
-\begin{figure}
-\centering
-\includegraphics{manuscript_files/figure-latex/MortalityNewYork-1.pdf}
-\caption{\textbf{Anomaly Detection for New York state mortality, 1999-2016.}
-The top panel contains the original time series (light gray) and the
-corrected, counter-factual time series in the absence of anomalies
-(blue). The red dots on the top panel correspond to the onset of
-detected anomalies. The bottom panel contains the magnitude and type of
-the outlier. We detect all three types of outliers in New York mortality
-data during period 1999-2016. We detect additive outliers (AO) in
-January 2000, September 2001, January 2005, and January 2013; temporary
-change (TC) outliers in February 2007 and January 2015; and a level
-shift (LS) starting in February 2004. Outliers range in significance
-from a low t-statistic of 3.91 in the TC 2015:01 to a high t-statistic
-of 6.40 in the AO outlier in 2001:09. \label{fig:mortnewyork}}
-\end{figure}
-
-\begin{verbatim}
-## numeric(0)
-\end{verbatim}
-
-In \textbf{\autoref{fig:mortnewyork}}, notice the strong level shift
+In \textbf{\autoref{fig:mortnewhamp}a}, notice the strong level shift
 (LS) that occurs in February 2004 (2004:02 t=5.59) which prevented
@@
 Contrast the mortality protection in New York with the enhanced
-mortality in New Hampshire (\textbf{\autoref{fig:mortnewhamp}}). Here we
-detect two highly significant level shifts (LS) in the monthly mortality
-data, first in April 2010 and again in November 2014 (2010:04 t=3.51;
-2014:11 t=6.68). These anomalies suggest New Hampshire experienced 9,700
-more deaths (+14\% more than expected) in a seven-year period beginning
-in early 2010. This is the single largest percentage mortality
-increase/decrease we detected among all states. Not coincidentally, NH
-has the second highest opioid-related mortality in the US
-\citep{beetham2019access} and it is reasonable that we detect this
-epidemic in our results.
+mortality in New Hampshire (\textbf{\autoref{fig:mortnewhamp}b}). Here
+we detect two highly significant level shifts (LS) in the monthly
+mortality data, first in April 2010 and again in November 2014 (2010:04
+t=3.51; 2014:11 t=6.68). These anomalies suggest New Hampshire
+experienced 9,700 more deaths (+14\% more than expected) in a seven-year
+period beginning in early 2010. This is the single largest percentage
+mortality increase/decrease we detected among all states. Not
+coincidentally, NH has the second highest opioid-related mortality in
+the US \citep{beetham2019access} and it is reasonable that we detect
+this epidemic in our results.
 
@@
 \includegraphics{manuscript_files/figure-latex/MortalityNewHamp-1.pdf}
-\caption{\textbf{Anomaly Detection for New Hampshire state mortality, 1999-2016.}
-Here we detect two outliers, both level shifts in outliers (LS) in April
-2010 and again in November 2014. These anomalies suggest New Hampshire
+\caption{\textbf{Anomaly Detection for New York (a) and New Hampshire (b) state mortality, 1999-2016.}
+The top part of each panel contains the original time series (light
+gray), the corrected, counter-factual time series in the absence of
+anomalies (blue), and the red dots correspond to the onset of detected
+anomalies. The bottom part of each panel contains the magnitude and type
+of the outlier in red. In New York (a), we detect additive outliers (AO)
+in January 2000, September 2001, January 2005, and January 2013;
+temporary change (TC) outliers in February 2007 and January 2015; and a
+level shift (LS) starting in February 2004. In New Hampshire (b), we
+detect two outliers, both level shift outliers (LS) in April 2010 and
+again in November 2014. These anomalies suggest New York experienced a
+significant mortality event in September 2001 and New Hampshire
 experienced approximately 9,700 more deaths than expected since 2010 or
@@
 
-\begin{verbatim}
-## numeric(0)
-\end{verbatim}
-
-In the case of New York and New Hampshire, some of the demographic
-anomalies have plausible explanations. It seems likely that New York's
-AO anomaly in September 2001 is caused by the 9/11 tragedy and the rise
-in New Hampshire's mortality starting in 2010 could be linked to the
-opioid epidemic. However, we detect numerous seemingly unexplainable
-demographic anomalies in other states. Figure 3 shows two such
-unexplained anomalies.
-
 \hypertarget{fertility}{%
@@
 We begin with a clear temporary change (TC) outlier for fertility in
-Louisiana (\textbf{\autoref{fig:fertla}}). Here we detect two TC
+Louisiana (\textbf{\autoref{fig:fertla}a}). Here we detect two TC
 outliers, nearly back-to-back in 2005 (2005:08 \emph{t} = -6; and
@@
 
+We also highlight a level shift (LS) outlier for fertility in
+Connecticut. Here we detect a strong (\emph{t}= -3.71; -285
+births/month) level shift toward lower fertility beginning in August
+2009 (2009:08) that continues through the end of the period
+(\textbf{\autoref{fig:fertla}b}). This LS toward lower fertility reduced
+the \# of births by -28,752 or -8.57\% lower than the counterfactual
+time series. The stock market crash of 2008, where the Dow Jones
+Industrial Average had the largest single-day loss up to that point,
+occured just 11 months before we detect a shift toward lower fertility.
+We believe the trend toward lower fertility in August 2009 and the stock
+market crash in September 2008 are linked.
+
 \begin{figure}
@@
 \includegraphics{manuscript_files/figure-latex/FertilityLouisiana-1.pdf}
-\caption{\textbf{Anomaly detection in Louisiana state fertility, 2003-2018.}
-Here we detect two outliers, back-to-back, likely resulting from
-Hurricane Katrina August and October 2005, representing a decrease of
-more than 4,400 births due to the hurricane. \label{fig:fertla}}
+\caption{\textbf{Anomaly detection in Louisiana (a) and Connecticut (b) state fertility, 2003-2018.}
+In (a), we detect two outliers, back-to-back, likely resulting from
+Hurricane Katrina in August and October 2005, representing a decrease of
+more than 4,400 births due to the hurricane. In (b), we detect one
+outlier, a level shift outlier (LS) in August 2009. We believe this
+reduction is attributable to the stock market crash 11 months earlier.
+\label{fig:fertla}}
 \end{figure}
-
-\begin{verbatim}
-## numeric(0)
-\end{verbatim}
-
-We also highlight a level shift (LS) outlier for fertility in
-Connecticut. Here we detect a strong (\emph{t}=
-\texttt{prettyNum(outlier.fert.conn\$outliers\$tstat,\ digits\ =\ 3)};
--285 births/month) level shift toward lower fertility beginning in
-August 2009 (2009:08) that continues through the end of the period
-(\textbf{\autoref{fig:fertconn}}). This LS toward lower fertility
-reduced the \# of births by -28,752 or -8.57\% lower than the
-counterfactual time series. The stock market crash of 2008, where the
-Dow Jones Industrial Average had the largest single-day loss up to that
-point, occured just 11 months before we detect a shift toward lower
-fertility. We believe the trend toward lower fertility in August 2009
-and the stock market crash in September 2008 are linked.
-
-\begin{figure}
-\centering
-\includegraphics{manuscript_files/figure-latex/FertilityConnecticut-1.pdf}
-\caption{\textbf{Anomaly detection in Hawaiian state fertility, 2003-2018.}
-Here we detect one outlier, an additive outlier (AO) in May 2014 (
-extit\{t\}=4.96), representing a 14\% increase in expected births in
-that month. \label{fig:fertconn}}
-\end{figure}
-
-\begin{verbatim}
-## numeric(0)
-\end{verbatim}
 
@@
 In the examples above, we highlighted four fertility/mortality anomalies
-with plausible explanations. However, these four anomalies are but a
-fraction of anomalies we detect. Here, we highlight four additional
-anomalies without easily identifiable explanations.
+with plausible explanations. In the case of New York and New Hampshire,
+the mortality anomalies have plausible explanations. It seems likely
+that New York's AO anomaly in September 2001 is caused by the 9/11
+tragedy and the rise in New Hampshire's mortality starting in 2010 could
+be linked to the opioid epidemic. Similarly, Louisiana's TC anomalies
+seem linked to Hurricanes Katrina and Rita while the LS anomaly in
+Connecticut's fertility appears linked to the Great Recession. However,
+we detect numerous seemingly unexplainable demographic anomalies in
+other states. \textbf{\autoref{fig:ferthawaii}} shows two such
+unexplained anomalies.
 
-In \textbf{\autoref{fig:ferthawaii}} we identify a single additive
+In \textbf{\autoref{fig:ferthawaii}a} we identify a single additive
 anomaly in Hawaiian fertility in May 2014. This is a strong anomaly with
@@
 
+This is contrast to \textbf{\autoref{fig:ferthawaii}b}, where we
+identify a strange mortality reduction in Ohio (\emph{t}-stat: -4.27).
+This LS is more than 1,100 deaths per month less than the counterfactual
+time series, suggesting had nearly 40,000 fewer deaths since February
+2015 than expected. This is the single-largest LS among all states. We
+could not identify the potential policies Ohio might have put into place
+to provide such a strong mortality protection.
+
 \begin{figure}
@@
 \includegraphics{manuscript_files/figure-latex/AnomalyHawaii-1.pdf}
-\caption{\textbf{Anomaly detection in Hawaiian state fertility, 2003-2018.}
-Here we detect one outlier, an additive outlier (AO) in May 2014 (
-extit\{t\}=4.96), representing a 14\% increase in expected births in
-that month. \label{fig:ferthawaii}}
+\caption{\textbf{Anomaly detection in Hawaii state fertility (a) and Ohio state mortality (b).}
+Here we detect one outlier, an additive outlier (AO) in May 2014 in
+Hawaii (a), representing a large, unexplainable 14\% increase in
+expected births in that month. We also detect a strong level shift (LS)
+reduction (b), representing a large, unexplainable reduction of nearly
+40,000 deaths in Ohio. \label{fig:ferthawaii}}
 \end{figure}
-
-\begin{verbatim}
-## numeric(0)
-\end{verbatim}
-
-We detect a very interesting, unexplainable mortality anomaly in North
-Carolina (\textbf{\autoref{fig:fertmiss}}). Beginning in February 2008 ,
-North Carolina saw a level shift (LS) (2008:02; \emph{t}-stat = -4.76)
-of 800 fewer deaths per month until the end of the period. Granted, the
-decade after the LS still saw three mortality anomalies that resulted in
-additional, unexpected deaths in 2009, 2012, and 2017, the LS is still a
-remarkable mortality reduction of nearly 100k (-88,335) deaths over 10
-years! While North Carolina saw the raw count of mortality remain
-constant or slightly reduce since 2008, the state grew from 9.3 million
-residents in 2008 to 10.4 million in 2018. In spite of its growing
-population, North Carolina has, rather miraculously, \emph{reduced} its
-overall mortality risk.
-\includegraphics{manuscript_files/figure-latex/AnomalyMississippi-1.pdf}
-
-\begin{verbatim}
-## numeric(0)
-\end{verbatim}
-
-Hawaii - AO X33 - LS X28 X22 X20 X05
 

### 2020-01-15 R/SCRIPTS/001-AnomaliesDetection.R

@@ -16,68 +16,31 @@ dat <-
 countylist = unique(dat$location)
 
 df <- data.frame()
-df_sums <- data.frame()
+# df_sums <- data.frame()
 sigma <- 3.5
 
-for(i in 1:length(unique(countylist))){
-
+for(this.state in unique(countylist)){
+  set.seed(1)
+  print(this.state)
   tryCatch({
     df2<- setNames(data.frame(matrix(ncol = 5, nrow = 0)), c("type", "ind", "time", "coefhat", "tstat"))
-    dat2 <- filter(dat, location == paste0(countylist[i]))
+    state <- this.state
+    dat2 <- filter(dat, location == state)
     dat3 <- ts(dat2$Target, start = year(min(dat$Year)), end = c(year(max(dat$Year)),12), frequency = 12)
     outlier.county <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)
-    
-    # df_sums2 <- data.frame(type = NA, tot = NA, abstot = NA, location = NA)
-    df_sums2 <- data.frame(y = NA, yadj=NA)
-    if(!is.null(outlier.county$times)){
-      
-      df2<- outlier.county$outliers
-      # df_sums2 <- data.frame(y=as.matrix(outlier.county$y), 
-      #                 yadj = as.matrix(outlier.county$yadj),
-      #                 month = as.Date(time(outlier.county$y))) %>%
-      #   mutate(ind = as.numeric(rownames(.)),
-      #          diff = y - yadj) %>%
-      #   left_join(., outlier.county$outliers) %>%
-      #   filter(diff != 0) %>%
-      #   fill(type) 
-        # group_by(type) %>%
-        # dplyr::summarise(tot = sum(diff),
-        #                  abstot = sum(abs(diff)))
-      
-      # df_sums2$location <- paste0(countylist[i])
-      df_sums2$y <- sum(outlier.county$y)
-      df_sums2$yadj <- sum(outlier.county$yadj)
-      
-    } else {
-      df2 <- rbind(df2, data.frame(type = NA,
-                                   ind = NA,
-                                   time = NA,
-                                   coefhat = NA,
-                                   tstat = NA))
-      # df_sums2 <- rbind(df_sums2)
-      df_sums2 <- rbind(df_sums2, data.frame(y=NA, yadj=NA))
-      
-    }
-    
[…]

### 2020-01-15 R/SCRIPTS/001-DataLoad.R

@@ -27,14 +27,11 @@ dat_mort <-
   arrange(location, Year)
 
 
+outlier.mort.ny <- read_rds("../R/DATA-PROCESSED/Plots/mort_ny")
+outlier.mort.nh <- read_rds("../R/DATA-PROCESSED/Plots/mort_nh")
+outlier.fert.conn <- read_rds("../R/DATA-PROCESSED/Plots/fert_conn")
+outlier.fert.la <- read_rds("../R/DATA-PROCESSED/Plots/fert_la")
+outlier.fert.hi <- read_rds("../R/DATA-PROCESSED/Plots/fert_hi")
+outlier.mort.oh <- read_rds("../R/DATA-PROCESSED/Plots/mort_oh")
 
-# df2<- setNames(data.frame(matrix(ncol = 5, nrow = 0)), c("type", "ind", "time", "coefhat", "tstat"))
-# dat2 <- filter(dat_mort, location == "X36")
-# dat3 <- ts(dat2$Target, start = year(min(dat_mort$Year)), end = c(year(max(dat_mort$Year)),12), frequency = 12)
-# outlier.county <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)
-# 
-# plot(outlier.county,
-#      args.lines.yadj = list(col = "black"),
-#      args.lines.y = list(col = "gray80")) +
-# title(main = "New York - Monthly Deaths")
-
+# overall_mortality <- read_rds("../R/DATA-PROCESSED/")

### 2020-01-15 R/SCRIPTS/002-StatesAnalysis.R

@@ -1,43 +1,70 @@
 ###------States Analysis  -----
 ## @knitr StatesAnalysis
 
+dat_fert <-
+  ###FERTILITY
+  bind_rows(read_tsv("./R/DATA-RAW/Fertility/Natality, 2003-2006,states.txt"),
+            read_tsv("./R/DATA-RAW/Fertility/Natality, 2007-2017,states.txt"))%>%
+  mutate(Month2 = str_pad(`Month Code`,2 , pad = "0"),
+         Month3 = as.yearmon(paste0(Year,"-", Month2)),
+         location = paste0("X", `State Code`)
+  ) %>%
+  dplyr::select(Year = Month3, location, Target = Births) %>%
+  na.omit %>%
+  arrange(location, Year)
+
+dat_mort <- 
+  ###MORTALITY 
+  read_tsv("./R/DATA-RAW/Mortality/Underlying Cause of Death, 1999-2017,states.txt") %>%
+  mutate(Month2 = substr(`Month Code`,6,7),
+         Month3 = as.yearmon(paste0(Year,"-", Month2)),
+         # location = paste0("X", `County Code`)
+         location = paste0("X", `State Code`)
+  ) %>%
+  dplyr::select(Year = Month3, location, Target = Deaths) %>%
+  na.omit %>%
+  arrange(location, Year)
+
+set.seed(1)
 # Mortality - New York
 dat2 <- filter(dat_mort, location == "X36")
 dat3 <- ts(dat2$Target, start = year(min(dat_mort$Year)), end = c(year(max(dat_mort$Year)),12), frequency = 12)
 outlier.mort.ny <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)
+write_rds(outlier.mort.ny, "./R/DATA-PROCESSED/Plots/mort_ny")
 
 # Mortality - New Hampshire
 dat2 <- filter(dat_mort, location == "X33")
 dat3 <- ts(dat2$Target, start = year(min(dat_mort$Year)), end = c(year(max(dat_mort$Year)),12), frequency = 12)
 outlier.mort.nh <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)
+write_rds(outlier.mort.nh, "./R/DATA-PROCESSED/Plots/mort_nh")
 
 # Fertility - Connecticut
[…]

### 2020-01-16 MANUSCRIPT/manuscript.Rmd
Editting the readme

@@
 
+**\autoref{sumtable}** reports the overall stuff.
+
+```{r, echo = FALSE}
+
+ anom_mort <- read_rds("../R/DATA-PROCESSED/mortality_anomalies") %>%
+   na.omit %>%
+   mutate(component = "Mortality")
+
+ anom_fert <- read_rds("../R/DATA-PROCESSED/fertility_anomalies") %>%
+   na.omit %>%
+   mutate(component = "Fertility")
+
+ anom <- rbind(anom_mort, anom_fert)
+
+ types <- anom %>%
+   group_by(type, component) %>%
+   dplyr::summarise(n = n(),
+                    states =  n_distinct(location)) %>%
+   ungroup() %>%
+   gather(test, stat_, n:states) %>%
+   mutate(variable = paste0(component, "_", test)) %>%
+   ungroup() %>%
+   dplyr::select(-component, -test) %>%
+   spread(variable, stat_)
+
+ kable(types,
+       "latex",
+       booktabs = T,
+       col.names = c( "Type","n","States","n","States"),
+       caption = "\\textbf{Summary by Anomaly Type and Demographic Component.} n refers to the number of anomolies for each type and demogrpahic component, States refers to the number of unique states exhibiting that specific Type-demogrpahic component.
+\\label{sumtable}") %>%
+   kable_styling(position = "center") %>%
+   add_header_above(c(" "=1, "Fertility" = 2, "Mortality" = 2))
+
+```
+
 ## Mortality 
@@
 
-```{r MortalityNewHamp, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly Detection for New York (a) and New Hampshire (b) state mortality, 1999-2016.} The top part of each panel contains the original time series (light gray), the corrected, counter-factual time series in the absence of anomalies (blue), and the red dots correspond to the onset of detected anomalies.
+```{r MortalityAnomalies, echo= FALSE, message = FALSE, warning = FALSE,dev=c('pdf','png'), fig.height = 5, fig.cap= paste("\\textbf{Anomaly Detection for New York (a) and New Hampshire (b) state mortality, 1999-2016.} The top part of each panel contains the original time series (light gray), the corrected, counter-factual time series in the absence of anomalies (blue), and the red dots correspond to the onset of detected anomalies.
 The bottom part of each panel contains the magnitude and type of the outlier in red.
@@
 
-```{r FertilityLouisiana, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Louisiana (a) and Connecticut (b) state fertility, 2003-2018.} In (a), we detect two outliers, back-to-back, likely resulting from Hurricane Katrina in August and October 2005, representing a decrease of more than 4,400 births due to the hurricane.
+```{r FertilityAnomalies, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Louisiana (a) and Connecticut (b) state fertility, 2003-2018.} In (a), we detect two outliers, back-to-back, likely resulting from Hurricane Katrina in August and October 2005, representing a decrease of more than 4,400 births due to the hurricane.
 In (b), we detect one outlier, a level shift outlier (LS) in August 2009.
@@
                labels= "auto")
-
 ```
@@
 
-```{r AnomalyHawaii, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Hawaii state fertility (a) and Ohio state mortality (b).} Here we detect one outlier, an additive outlier (AO) in May 2014 in Hawaii (a), representing a large, unexplainable 14% increase in expected births in that month.
+```{r TrueAnomalies, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Hawaii state fertility (a) and Ohio state mortality (b).} Here we detect one outlier, an additive outlier (AO) in May 2014 in Hawaii (a), representing a large, unexplainable 14% increase in expected births in that month.
 We also detect a strong level shift (LS) reduction (b), representing a large, unexplainable reduction of nearly 40,000 deaths in Ohio.

### 2020-01-16 MANUSCRIPT/manuscript.tex
Editting the readme

@@
 
+\textbf{\autoref{sumtable}} reports the overall stuff.
+
+\begin{table}
+
+\caption{\label{tab:unnamed-chunk-1}\textbf{Summary by Anomaly Type and Demographic Component.} n refers to the number of anomolies for each type and demogrpahic component, States refers to the number of unique states exhibiting that specific Type-demogrpahic component.
+\label{sumtable}}
+\centering
+\begin{tabular}[t]{lrrrr}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Fertility} & \multicolumn{2}{c}{Mortality} \\
+\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
+Type & n & States & n & States\\
+\midrule
+AO & 9 & 9 & 85 & 38\\
+LS & 6 & 4 & 22 & 17\\
+TC & 7 & 6 & 49 & 28\\
+\bottomrule
+\end{tabular}
+\end{table}
+
 \hypertarget{mortality}{%
@@
 \centering
-\includegraphics{manuscript_files/figure-latex/MortalityNewHamp-1.pdf}
+\includegraphics{manuscript_files/figure-latex/MortalityAnomalies-1.pdf}
 \caption{\textbf{Anomaly Detection for New York (a) and New Hampshire (b) state mortality, 1999-2016.}
@@
 \centering
-\includegraphics{manuscript_files/figure-latex/FertilityLouisiana-1.pdf}
+\includegraphics{manuscript_files/figure-latex/FertilityAnomalies-1.pdf}
 \caption{\textbf{Anomaly detection in Louisiana (a) and Connecticut (b) state fertility, 2003-2018.}
@@
 \centering
-\includegraphics{manuscript_files/figure-latex/AnomalyHawaii-1.pdf}
+\includegraphics{manuscript_files/figure-latex/TrueAnomalies-1.pdf}
 \caption{\textbf{Anomaly detection in Hawaii state fertility (a) and Ohio state mortality (b).}

### 2020-01-16 R/SCRIPTS/001-AnomaliesDetection.R

@@ -16,7 +16,7 @@ dat <-
 countylist = unique(dat$location)
 
 df <- data.frame()
-# df_sums <- data.frame()
+df_sums <- data.frame()
 sigma <- 3.5
 
 for(this.state in unique(countylist)){
@@ -24,6 +24,7 @@ for(this.state in unique(countylist)){
   print(this.state)
   tryCatch({
     df2<- setNames(data.frame(matrix(ncol = 5, nrow = 0)), c("type", "ind", "time", "coefhat", "tstat"))
+    
     state <- this.state
     dat2 <- filter(dat, location == state)
     dat3 <- ts(dat2$Target, start = year(min(dat$Year)), end = c(year(max(dat$Year)),12), frequency = 12)
@@ -31,6 +32,12 @@ for(this.state in unique(countylist)){
     df2<- outlier.county$outliers
     df2$location <- state
     df <- bind_rows(df, df2)
+    
+    df_sums2 <- setNames(data.frame(matrix(ncol=3, nrow=1)), c("y", "yadj", "loc"))
+    df_sums2$ydiff <- sum(outlier.county$effects)
+    df_sums2$yadj <- sum(outlier.county$yadj)
+    df_sums2$loc <- this.state
+    
   } , error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
   
 }

### 2020-01-17 MANUSCRIPT/manuscript.Rmd
Display items finished.
To do:
- Perhaps an example in the Methods section?
- A little more text in the Results section
- A conclusion
- Rewrite the Abstract

@@
 read_chunk('../R/SCRIPTS/002-StatesAnalysis.R')
+read_chunk('../R/SCRIPTS/003-SummaryStats.R')
 ```
@@
 
-<!-- # ```{r StatesAnalysis, include=FALSE} -->
-<!-- # ``` -->
+```{r SumStats, include=FALSE}
+```
 
@@
 Based on Chen and Liu’s [-@chen1993joint] recommendation, we set a critical value of 3.5 which generally minimizes the possibility of Type I errors or false positive outliers.
-Outliers are reported using a simple t-statistic.
+
+We report outliers using a simple t-statistic and report the size of the outlier, or its impact on the time series, by subtracting $\hat{y}$, the adjusted time series without outliers removed, from $y$, the original, unadjusted time series.
 
@@
 
-**\autoref{sumtable}** reports the overall stuff.
-
 ```{r, echo = FALSE}
 
+sums2 <- sums %>%
+  dplyr::select(percent_anom = `\\% of Total`, everything())
+kable(sums,
+       "latex",
+       booktabs = T,
+       format.args = list(big.mark = ","),
+       escape = FALSE,
+       caption = "\\textbf{Summary by Demographic Component.}  refers to the number of anomolies for each type and demographic component, States refers to the number of unique states exhibiting that specific Type-demographic component.
+\\label{sumtable}") %>%
+   kable_styling(position = "center")
+
+```
+
+**\autoref{sumtable}** report the overall number of anomalies we detect of each type for births and deaths and some summary statistics across all anomaly types.
+We find considerably more mortality anomalies (n = `r paste(sums[2,2])`) than fertility anomalies (n = `r paste(sums[1,2])`).
+This is likely because mortality is more easily subject to instantaneous change compared with fertility.
+Changes in fertility should theoretically manifest approximately 9-months after change.
+
+We also find more anomalous deaths (`r paste(prettyNum(sums[2,6], digits = 0, big.mark = ","))`) than anomalous births (`r paste(prettyNum(sums[1,6], digits = 0, big.mark = ","))`).
+Fertility anomalies overwhelmingly tended toward lower fertility, with only a few fertility anomalies yielding more births.
+Conversely, mortality anomalies tended to be more evenly split between positive and negative anomalies, but tended toward more deaths rather than less deaths.
+These anomalous deaths and births account for `r sums2$percent_anom[2]`% and `r sums2$percent_anom[1]`% over the entire time series, respectively.
+
+**\autoref{fig:summap}** maps these results.
+Three states exhibited neither mortality nor fertility anomalies: Alaska, North Dakota, and South Dakota.
+Another nine states exhibited just a single anomaly (DE, DC, ID, IL, KS, NV, NM, UT, and WY).
+Both New York and Massachusetts exhibited the most anomalies with nine.
+
+```{r AnomalyMap, echo= FALSE, message = FALSE, warning = FALSE, fig.height=7, fig.cap= paste("\\textbf{Summary of Anomalies by type, demographic component, and State.} We observe much fewer fertility anomalies than mortality anomalies.
+\\label{fig:summap}")}
+
+data(state_laea)
 anom_mort <- read_rds("../R/DATA-PROCESSED/mortality_anomalies") %>%
@@
 
- anom <- rbind(anom_mort, anom_fert)
+anom_mort2 <- rbind(anom_mort, anom_fert) %>%
+  group_by(type, location, component) %>%
+  dplyr::summarise(Anomalies =  n()) %>%
+  mutate(ident = paste0(component, " - ", type),
+         GEOID = substr(location, 2,3)) %>%
+  ungroup() %>%
+  dplyr::select(GEOID, Anomalies, ident) %>%
+  spread(ident, Anomalies) %>%
+    mutate_all(funs(replace_na(.,0)))
 
- types <- anom %>%
-   group_by(type, component) %>%
-   dplyr::summarise(n = n(),
-                    states =  n_distinct(location)) %>%
-   ungroup() %>%
-   gather(test, stat_, n:states) %>%
-   mutate(variable = paste0(component, "_", test)) %>%
-   ungroup() %>%
-   dplyr::select(-component, -test) %>%
-   spread(variable, stat_)
+states <- left_join(state_laea, anom_mort2) %>%
+  gather(ident, Anomalies, `Fertility - AO`:`Mortality - TC`) %>%
+  mutate_all(funs(replace_na(.,0))) %>%
+  mutate(Anomalies = factor(Anomalies))
 
- kable(types,
-       "latex",
-       booktabs = T,
-       col.names = c( "Type","n","States","n","States"),
-       caption = "\\textbf{Summary by Anomaly Type and Demographic Component.} n refers to the number of anomolies for each type and demogrpahic component, States refers to the number of unique states exhibiting that specific Type-demogrpahic component.
-\\label{sumtable}") %>%
-   kable_styling(position = "center") %>%
-   add_header_above(c(" "=1, "Fertility" = 2, "Mortality" = 2))
+states$ident2 <- factor(
+  states$ident,
+  levels = c("Mortality - AO", "Fertility - AO", 
+             "Mortality - TC", "Fertility - TC", 
+             "Mortality - LS", "Fertility - LS")
+)
 
+colorpal <- c("#ffffff", "#ffffcc", "#a1dab4", "#41b6c4", "#2c7fb8", "#253494")
+
+  ggplot(data=states) +
+  geom_sf(aes(fill = Anomalies),  lwd = 1/2) +
+  theme_bw() +
+  theme(axis.title=element_blank(),
+        axis.text=element_blank(),
+        axis.ticks=element_blank(),
+        legend.position="bottom"
+  )+
+  guides(fill=guide_legend(nrow=1,byrow=TRUE)) +
+  facet_wrap(~ident2, ncol=2) +
+    scale_fill_manual(name = "# of Anomalies", values = colorpal) +
+
+  NULL
 ```

### 2020-01-17 MANUSCRIPT/manuscript.tex
Display items finished.
To do:
- Perhaps an example in the Methods section?
- A little more text in the Results section
- A conclusion
- Rewrite the Abstract

--- 2020-01-15 MANUSCRIPT/manuscript.tex
@@
 3.5 which generally minimizes the possibility of Type I errors or false
-positive outliers. Outliers are reported using a simple t-statistic.
+positive outliers.
+
+We report outliers using a simple t-statistic and report the size of the
+outlier, or its impact on the time series, by subtracting \(\hat{y}\),
+the adjusted time series without outliers removed, from \(y\), the
+original, unadjusted time series.
 
@@
 \subsection{Overall Anomalies}\label{overall-anomalies}}
+
+\begin{table}
+
+\caption{\label{tab:unnamed-chunk-1}\textbf{Summary by Demographic Component.}  refers to the number of anomolies for each type and demographic component, States refers to the number of unique states exhibiting that specific Type-demographic component.
+\label{sumtable}}
+\centering
+\begin{tabular}[t]{lrllrrr}
+\toprule
+Component & Anomalies & $y$ & $\hat{y}$ & $y-\hat{y}$ & $|y-\hat{y}|$ & \% of Total\\
+\midrule
+Fertility & 22 & 25.48M & 25.68M & -202,217 & 226,476 & 0.889\\
+Mortality & 156 & 43.57M & 43.52M & 48,535 & 627,364 & 1.440\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\textbf{\autoref{sumtable}} report the overall number of anomalies we
+detect of each type for births and deaths and some summary statistics
+across all anomaly types. We find considerably more mortality anomalies
+(n = 156) than fertility anomalies (n = 22). This is likely because
+mortality is more easily subject to instantaneous change compared with
+fertility. Changes in fertility should theoretically manifest
+approximately 9-months after change.
+
+We also find more anomalous deaths (627,364) than anomalous births
+(226,476). Fertility anomalies overwhelmingly tended toward lower
+fertility, with only a few fertility anomalies yielding more births.
+Conversely, mortality anomalies tended to be more evenly split between
+positive and negative anomalies, but tended toward more deaths rather
+than less deaths. These anomalous deaths and births account for 1.44\%
+and 0.889\% over the entire time series, respectively.
+
+\textbf{\autoref{fig:summap}} maps these results. Three states exhibited
+neither mortality nor fertility anomalies: Alaska, North Dakota, and
+South Dakota. Another nine states exhibited just a single anomaly (DE,
+DC, ID, IL, KS, NV, NM, UT, and WY). Both New York and Massachusetts
+exhibited the most anomalies with nine.
+
+\begin{figure}
+\centering
+\includegraphics{manuscript_files/figure-latex/AnomalyMap-1.pdf}
+\caption{\textbf{Summary of Anomalies by type, demographic component, and State.}
+We observe much fewer fertility anomalies than mortality anomalies.
+\label{fig:summap}}
+\end{figure}
 

### 2020-01-17 R/SCRIPTS/000-Libraries.R

@@ -47,6 +47,7 @@ pkgs <- c(
   "tmaptools",     # Cartographic tools
   "tigris",        # US shapefiles
   "censusapi",     # Census Data
+  "tidycensus",
   "sp",            # Spatial Objects
   "grid",          # Plot Grids
   "kableExtra",    # Pretty Tables
@@ -62,7 +63,8 @@ pkgs <- c(
   "CausalImpact",
   "bsts",
   "lubridate",
-  "tsoutliers"
+  "tsoutliers",
+  "numform"
 )
 
 # Install missing packages

### 2020-01-17 R/SCRIPTS/001-AnomaliesDetection.R

@@ -34,10 +34,13 @@ for(this.state in unique(countylist)){
     df <- bind_rows(df, df2)
     
     df_sums2 <- setNames(data.frame(matrix(ncol=3, nrow=1)), c("y", "yadj", "loc"))
+    df_sums2$y <- sum(outlier.county$y)
     df_sums2$ydiff <- sum(outlier.county$effects)
+    df_sums2$ydiffabs <- sum(abs(outlier.county$effects))
     df_sums2$yadj <- sum(outlier.county$yadj)
     df_sums2$loc <- this.state
     
+    df_sums <- bind_rows(df_sums, df_sums2)
   } , error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
   
 }
@@ -47,7 +50,7 @@ for(this.state in unique(countylist)){
 #          excessmortper = (excessmort/y)*100)
 
 write_rds(df, "./R/DATA-PROCESSED/fertility_anomalies")
-# write_rds(df_sums2, "./R/DATA-PROCESSED/fertility_anomalies_sums")
+write_rds(df_sums, "./R/DATA-PROCESSED/fertility_anomalies_sums")
 
 dat <- 
   ###MORTALITY 
@@ -65,6 +68,8 @@ dat <-
 countylist = unique(dat$location)
 
 df <- data.frame()
+df_sums <- data.frame()
+
 sigma <- 3.5
 
 for(this.state in unique(countylist)){
@@ -79,11 +84,21 @@ for(this.state in unique(countylist)){
     df2<- outlier.county$outliers
     df2$location <- state
     df <- bind_rows(df, df2)
+    
+    
+    df_sums2 <- setNames(data.frame(matrix(ncol=3, nrow=1)), c("y", "yadj", "loc"))
+    df_sums2$y <- sum(outlier.county$y)
+    df_sums2$ydiff <- sum(outlier.county$effects)
+    df_sums2$ydiffabs <- sum(abs(outlier.county$effects))
+    df_sums2$yadj <- sum(outlier.county$yadj)
+    df_sums2$loc <- this.state
+    df_sums <- bind_rows(df_sums, df_sums2)
+    
   } , error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
   
 }
 write_rds(df, "./R/DATA-PROCESSED/mortality_anomalies")
-# write_rds(df_sums, "./R/DATA-PROCESSED/mortality_anomalies_sum")
[…]

### 2020-01-20 MANUSCRIPT/manuscript.Rmd
Update manuscript.Rmd
Change to abstract and title

@@
 ---
-title: "Effects without a Cause: The Search for Demographic Anomalies"
+title: "Causal Inference in Population Trends: A Technique for finding Demographic Anomalies in Big Data"
 
@@
 - causal inference
+- abductive reasoning
+- big data
 
 abstract: |
-  The proliferation of data, modern computing advances, and powerful statistical algorithms unlock the potential search for hidden, previously un- or understudied demographic anomalies.
-Here, we demonstrate how demographers can use causal inference techniques to identify anamolies by investigating US state-level fertility and mortality time series since 1999 to uncover the hidden baby booms/busts and mortality plagues?/non-plauges?.
-We find 38 states exhibited at least one mortality anomaly, totalling more than 318k anomalous deaths and an additional 6 states exhibited at least one mortality non-plague?
-totalling 164k protective deaths.
-12 states exhibited baby busts, totalling more than 240k missing births and an additional 11 states exhibited baby booms totalling more than 134k additional births.
-These results suggest the widespread detection of demographic anomalies.
-Our analysis does not examine the *causes* of these anomalies and our results point to important further research on the causes of anomalous demographic behavior.
+The proliferation of big data, wider access to advanced computing platforms, and the development of powerful statistical algorithms can uncover hidden anomalies in demographic data that have previously been dismissed as noise.
+Here, we demonstrate how demographers can use causal inference techniques and abductive reasoning to identify fertility and mortality anamolies in state-level data trends by investigating demographic time series data since 1999.
+In demonstrating our technique, we uncover real spikes and dips in fertility and mortality over time, distinguished from regular trend variations.
+We show that 38 states exhibited at least one mortality anomaly over time, totalling more than 318,000 additional deaths, while an additional six states exhibited periods of decline resuling in 164,000 fewer deaths.
+Twelve states exhibited real, marked periods of declines in fertility, totalling more than 240,000 fewer births than expected, and eleven states had periods of excess births, resulting in an addition 134,000 births.
+Our results show how researchers can detect demographic anomalies in trend data and use those anomalies as a beginning point for further investigation into the causes.
 

### 2020-01-20 MANUSCRIPT/manuscript.Rmd
Update manuscript.Rmd
Some changes to results section

--- 2020-01-17 MANUSCRIPT/manuscript.Rmd
@@
 We find considerably more mortality anomalies (n = `r paste(sums[2,2])`) than fertility anomalies (n = `r paste(sums[1,2])`).
-This is likely because mortality is more easily subject to instantaneous change compared with fertility.
-Changes in fertility should theoretically manifest approximately 9-months after change.
+Given that anomalies are always related to events [@Song2018ACM], the finding of more mortality than fertility anomalies is not surprising.
+Mortality is more easily subject to instantaneous change (like a natural disaster) compared with fertility, and fertility is linked to human decision-making more strongly than mortality [@Stein2014Fertility].
+Previous studies document that some events, like the 2005 Oklahoma City bombings, resulted in both fertility and mortality changes [@Rogers2005OKBombing], but changes in fertility manifest approximately 9-months after an event, and the relationship between a fertility-inducing event and behavior change (such as a prolonged blizzard or power blackout) is less strong than the relationship between a mortality-inducing event (such as an earthquake) and death.
+For example, Evans and his colleagues [@Evans2008Hurricanebirth] show that catastrophic weather events can lead to increased fertility, but some events reduce fertility.
+Furthermore, the effect is seen primarily in those who already have one child, with childless people unlikely to have a child due to a weather event.
 
 We also find more anomalous deaths (`r paste(prettyNum(sums[2,6], digits = 0, big.mark = ","))`) than anomalous births (`r paste(prettyNum(sums[1,6], digits = 0, big.mark = ","))`).
-Fertility anomalies overwhelmingly tended toward lower fertility, with only a few fertility anomalies yielding more births.
-Conversely, mortality anomalies tended to be more evenly split between positive and negative anomalies, but tended toward more deaths rather than less deaths.
-These anomalous deaths and births account for `r sums2$percent_anom[2]`% and `r sums2$percent_anom[1]`% over the entire time series, respectively.
+Fertility anomalies overwhelmingly tend toward lower fertility, with only a few fertility anomalies yielding more births.
+This is consistent with research findings that link high severity weather events to significantly lower fertility [@Evans2008Hurricanebirth].
+Conversely, mortality anomalies tend to be more evenly split between positive and negative anomalies, but tend toward more deaths rather than less deaths.
+Again, this is not surprising, as the kinds of events that increase death (e.g., plant closings are associated with increased opioid death [@Venkataramani2019]) are more common than the events that decrease death (e.g., cannabis legalization is associated with decreased opioid deaths [@Livingston2017Cannabis]) These anomalous deaths and births account for `r sums2$percent_anom[2]`% and `r sums2$percent_anom[1]`% over the entire time series, respectively.
 

### 2020-01-20 MANUSCRIPT/manuscript.Rmd
Update manuscript.Rmd

@@
 We find considerably more mortality anomalies (n = `r paste(sums[2,2])`) than fertility anomalies (n = `r paste(sums[1,2])`).
-Given that anomalies are always related to events [@Song2018ACM], the finding of more mortality than fertility anomalies is not surprising.
-Mortality is more easily subject to instantaneous change (like a natural disaster) compared with fertility, and fertility is linked to human decision-making more strongly than mortality [@Stein2014Fertility].
-Previous studies document that some events, like the 2005 Oklahoma City bombings, resulted in both fertility and mortality changes [@Rogers2005OKBombing], but changes in fertility manifest approximately 9-months after an event, and the relationship between a fertility-inducing event and behavior change (such as a prolonged blizzard or power blackout) is less strong than the relationship between a mortality-inducing event (such as an earthquake) and death.
-For example, Evans and his colleagues [@Evans2008Hurricanebirth] show that catastrophic weather events can lead to increased fertility, but some events reduce fertility.
-Furthermore, the effect is seen primarily in those who already have one child, with childless people unlikely to have a child due to a weather event.
+Given that anomalies are always the produce of events [@Song2018ACM], the finding of more mortality than fertility anomalies is not surprising.
+Mortality is likely to spike in response to a catastrophic event (like an earthquake or terrorist attack) while the effects of a catastrophic event on fertility is less predictable.
+One reason for this is that fertility is linked to human decision-making more directly than mortality [@Stein2014Fertility] which results in more varied outcomes, so, for example, researchers find that catastrophic weather events can increase childbearing among those who already have a child, but not among the childless [@EvansHurricanebirth].
+Some events, like the 1995 Oklahoma City bombings, resulted in both fertility and mortality changes.
+However, changes in fertility manifest over a longer time horizon after an event, and the relationship between a fertility-inducing event and behavior change is less strong than the relationship between a mortality-inducing event (such as a terrorist incident) and death [@Rodgers2005OKBombing].
+Of course, not all events that impact fertility and mortality are catastrophic.
+Researchers have documented that more commonplace events such as massive layoff [@Venkataramani2019] or policy changes [@Livingston2017Cannabis] can also create anomalies in mortality patterns, but we have not found such evidence for fertility.
 
-We also find more anomalous deaths (`r paste(prettyNum(sums[2,6], digits = 0, big.mark = ","))`) than anomalous births (`r paste(prettyNum(sums[1,6], digits = 0, big.mark = ","))`).
+Consistent with more anomalies across the time lines, we find more anomalous deaths (`r paste(prettyNum(sums[2,6], digits = 0, big.mark = ","))`) than anomalous births (`r paste(prettyNum(sums[1,6], digits = 0, big.mark = ","))`).
 Fertility anomalies overwhelmingly tend toward lower fertility, with only a few fertility anomalies yielding more births.
@@
 ## Mortality 
-We begin with mortality for New York State (**\autoref{fig:mortnewhamp}a**).
-We identify seven anomalies in the mortality time series for New York, all with t-statistics in excess of 3.91, making these anomalies highly significant.
+Since the purpose of our paper is to demonstrate how demographers can make use of causal inference technique, we will not discuss all of our findings in this paper.
+Instead, we will use a few cases for illustration.
+For mortality, we will use New York State as an example (**\autoref{fig:mortnewhamp}a**).
+We identify seven anomalies in the mortality time series for New York, all with t-statistics in excess of 3.91, making these anomalies unlikely to be due to chance.
 The algorithm correctly identifies September 2001 as an additive outlier (2001:09 t=6.40) where mortality in that month was 1,628 higher than anticipated.
@@
 These are the kinds of questions that arise from our analyses.
+The purpose of our paper is not to answer these questions, but our findings underscore the need for more research that takes an abductive approach.
+By identifying anomalies through an inductive process, researchers can then look for underlying causes.
+Once those causes are identified, researchers can then use a deductive process to see if such events predict other (or future) anomalies.
 
-Contrast the mortality protection in New York with the enhanced mortality in New Hampshire (**\autoref{fig:mortnewhamp}b**).
-Here we detect two highly significant level shifts (LS) in the monthly mortality data, first in April 2010 and again in November 2014 (2010:04 t=3.51; 2014:11 t=6.68).
+To see the potential for combinging abductive reasoning with causal inference, contrast the mortality protection in New York with the enhanced mortality in New Hampshire (**\autoref{fig:mortnewhamp}b**).
+In New Hampshire we detect two significant level shifts (LS) in the monthly mortality data, first in April 2010 and again in November 2014 (2010:04 t=3.51; 2014:11 t=6.68).
 These anomalies suggest New Hampshire experienced 9,700 more deaths (+14% more than expected) in a seven-year period beginning in early 2010.
-This is the single largest percentage mortality increase/decrease we detected among all states.
-Not coincidentally, NH has the second highest opioid-related mortality in the US [@beetham2019access] and it is reasonable that we detect this epidemic in our results.
+These are events not experienced by neighborhing states during the same time period, and this is the single largest percentage mortality increase/decrease we detected among all states.
+Not coincidentally, NH has the second highest opioid-related mortality in the US [@beetham2019access] and it is likely that we detect this epidemic in our results.
+Isolating these anomalies and testing them against opioid sales data might yeild intriguing results.
 
@@
 
-We begin with a clear temporary change (TC) outlier for fertility in Louisiana (**\autoref{fig:fertla}a**).
+To illustrate how researchers can examine fertility anomalies, we begin with a clear temporary change (TC) outlier for fertility in Louisiana (**\autoref{fig:fertla}a**).
 Here we detect two TC outliers, nearly back-to-back in 2005 (2005:08 *t* = `r prettyNum(outlier.fert.la$outliers$tstat[1], digits = 3)`; and 2005:10 *t* = `r prettyNum(outlier.fert.la$outliers$tstat[2], digits = 3)`), coinciding with the destruction of Hurricane Katrina from the same year.
-The migration associated with Hurricane Katrina has garnered most of the attention of demographers [@fussellRecoveryMigrationCity2014;@horiDisplacementDynamicsSouthern2009], but the hurricane had a very clear impact on fertility behaviors too, creating a mini "baby bust" in Louisiana.
+The migration associated with Hurricane Katrina has garnered most of the attention of demographers [@fussellRecoveryMigrationCity2014;@horiDisplacementDynamicsSouthern2009], but the hurricane had a very clear impact on fertility behaviors too, creating a mini "baby bust" in Louisiana with the departure of many people in their childbearing years.
 We estimate `r prettyNum(abs(sum(outlier.fert.la$y - outlier.fert.la$yadj)), digits = 0, big.mark = ",")` fewer births in Louisiana compared to the counterfactual, likely attributable to the Hurricane.
@@
 The stock market crash of 2008, where the Dow Jones Industrial Average had the largest single-day loss up to that point, occured just 11 months before we detect a shift toward lower fertility.
-We believe the trend toward lower fertility in August 2009 and the stock market crash in September 2008 are linked.
+It is possible that the trend toward lower fertility in August 2009 and the stock market crash in September 2008 are linked, although, as stated previously, we are speculating only to illustrate how causal inference could work in an abductive research agenda.
 

### 2020-01-21 MANUSCRIPT/manuscript.Rmd
Fixed the bib file.

@@
 
-abstract: |
-The proliferation of big data, wider access to advanced computing platforms, and the development of powerful statistical algorithms can uncover hidden anomalies in demographic data that have previously been dismissed as noise.
+abstract: "The proliferation of big data, wider access to advanced computing platforms, and the development of powerful statistical algorithms can uncover hidden anomalies in demographic data that have previously been dismissed as noise.
 Here, we demonstrate how demographers can use causal inference techniques and abductive reasoning to identify fertility and mortality anamolies in state-level data trends by investigating demographic time series data since 1999.
@@
 Twelve states exhibited real, marked periods of declines in fertility, totalling more than 240,000 fewer births than expected, and eleven states had periods of excess births, resulting in an addition 134,000 births.
-Our results show how researchers can detect demographic anomalies in trend data and use those anomalies as a beginning point for further investigation into the causes.
+Our results show how researchers can detect demographic anomalies in trend data and use those anomalies as a beginning point for further investigation into the causes."
 
@@
 Mortality is likely to spike in response to a catastrophic event (like an earthquake or terrorist attack) while the effects of a catastrophic event on fertility is less predictable.
-One reason for this is that fertility is linked to human decision-making more directly than mortality [@Stein2014Fertility] which results in more varied outcomes, so, for example, researchers find that catastrophic weather events can increase childbearing among those who already have a child, but not among the childless [@EvansHurricanebirth].
+One reason for this is that fertility is linked to human decision-making more directly than mortality [@stein2014couples] which results in more varied outcomes, so, for example, researchers find that catastrophic weather events can increase childbearing among those who already have a child, but not among the childless [@Evans2008Hurricanebirth].
 Some events, like the 1995 Oklahoma City bombings, resulted in both fertility and mortality changes.

### 2020-01-21 MANUSCRIPT/manuscript.tex
Fixed the bib file.

@@
 {
-  \title{\bf Effects without a Cause: The Search for Demographic Anomalies}
+  \title{\bf Causal Inference in Population Trends: A Technique for finding
+Demographic Anomalies in Big Data}
 
@@
   \begin{center}
-    {\LARGE\bf Effects without a Cause: The Search for Demographic Anomalies}
+    {\LARGE\bf Causal Inference in Population Trends: A Technique for finding
+Demographic Anomalies in Big Data}
   \end{center}
@@
 \begin{abstract}
-The proliferation of data, modern computing advances, and powerful
-statistical algorithms unlock the potential search for hidden,
-previously un- or understudied demographic anomalies. Here, we
-demonstrate how demographers can use causal inference techniques to
-identify anamolies by investigating US state-level fertility and
-mortality time series since 1999 to uncover the hidden baby booms/busts
-and mortality plagues?/non-plauges?. We find 38 states exhibited at
-least one mortality anomaly, totalling more than 318k anomalous deaths
-and an additional 6 states exhibited at least one mortality non-plague?
-totalling 164k protective deaths. 12 states exhibited baby busts,
-totalling more than 240k missing births and an additional 11 states
-exhibited baby booms totalling more than 134k additional births. These
-results suggest the widespread detection of demographic anomalies. Our
-analysis does not examine the \emph{causes} of these anomalies and our
-results point to important further research on the causes of anomalous
-demographic behavior.\\
+The proliferation of big data, wider access to advanced computing
+platforms, and the development of powerful statistical algorithms can
+uncover hidden anomalies in demographic data that have previously been
+dismissed as noise. Here, we demonstrate how demographers can use causal
+inference techniques and abductive reasoning to identify fertility and
+mortality anamolies in state-level data trends by investigating
+demographic time series data since 1999. In demonstrating our technique,
+we uncover real spikes and dips in fertility and mortality over time,
+distinguished from regular trend variations. We show that 38 states
+exhibited at least one mortality anomaly over time, totalling more than
+318,000 additional deaths, while an additional six states exhibited
+periods of decline resuling in 164,000 fewer deaths. Twelve states
+exhibited real, marked periods of declines in fertility, totalling more
+than 240,000 fewer births than expected, and eleven states had periods
+of excess births, resulting in an addition 134,000 births. Our results
+show how researchers can detect demographic anomalies in trend data and
+use those anomalies as a beginning point for further investigation into
+the causes.
 \end{abstract}
@@
 \noindent%
-{\it Keywords:} mortality, fertility, causal inference
+{\it Keywords:} mortality, fertility, causal inference, abductive reasoning, big data
 \vfill
@@
 across all anomaly types. We find considerably more mortality anomalies
-(n = 156) than fertility anomalies (n = 22). This is likely because
-mortality is more easily subject to instantaneous change compared with
-fertility. Changes in fertility should theoretically manifest
-approximately 9-months after change.
+(n = 156) than fertility anomalies (n = 22). Given that anomalies are
+always the produce of events \citep{Song2018ACM}, the finding of more
+mortality than fertility anomalies is not surprising. Mortality is
+likely to spike in response to a catastrophic event (like an earthquake
+or terrorist attack) while the effects of a catastrophic event on
+fertility is less predictable. One reason for this is that fertility is
+linked to human decision-making more directly than mortality
+\citep{stein2014couples} which results in more varied outcomes, so, for
+example, researchers find that catastrophic weather events can increase
+childbearing among those who already have a child, but not among the
+childless \citep{Evans2008Hurricanebirth}. Some events, like the 1995
+Oklahoma City bombings, resulted in both fertility and mortality
+changes. However, changes in fertility manifest over a longer time
+horizon after an event, and the relationship between a
+fertility-inducing event and behavior change is less strong than the
+relationship between a mortality-inducing event (such as a terrorist
+incident) and death \citep{Rodgers2005OKBombing}. Of course, not all
+events that impact fertility and mortality are catastrophic. Researchers
+have documented that more commonplace events such as massive layoff
+\citep{Venkataramani2019} or policy changes
+\citep{Livingston2017Cannabis} can also create anomalies in mortality
+patterns, but we have not found such evidence for fertility.
 
-We also find more anomalous deaths (627,364) than anomalous births
-(226,476). Fertility anomalies overwhelmingly tended toward lower
-fertility, with only a few fertility anomalies yielding more births.
-Conversely, mortality anomalies tended to be more evenly split between
-positive and negative anomalies, but tended toward more deaths rather
-than less deaths. These anomalous deaths and births account for 1.44\%
-and 0.889\% over the entire time series, respectively.
+Consistent with more anomalies across the time lines, we find more
+anomalous deaths (627,364) than anomalous births (226,476). Fertility
+anomalies overwhelmingly tend toward lower fertility, with only a few
+fertility anomalies yielding more births. This is consistent with
+research findings that link high severity weather events to
+significantly lower fertility \citep{Evans2008Hurricanebirth}.
+Conversely, mortality anomalies tend to be more evenly split between
+positive and negative anomalies, but tend toward more deaths rather than
+less deaths. Again, this is not surprising, as the kinds of events that
+increase death (e.g., plant closings are associated with increased
+opioid death \citep{Venkataramani2019}) are more common than the events
+that decrease death (e.g., cannabis legalization is associated with
+decreased opioid deaths \citep{Livingston2017Cannabis}) These anomalous
+deaths and births account for 1.44\% and 0.889\% over the entire time
+series, respectively.
 
@@
 
-We begin with mortality for New York State
+Since the purpose of our paper is to demonstrate how demographers can
+make use of causal inference technique, we will not discuss all of our
+findings in this paper. Instead, we will use a few cases for
+illustration. For mortality, we will use New York State as an example
 (\textbf{\autoref{fig:mortnewhamp}a}). We identify seven anomalies in
 the mortality time series for New York, all with t-statistics in excess
-of 3.91, making these anomalies highly significant. The algorithm
-correctly identifies September 2001 as an additive outlier (2001:09
-t=6.40) where mortality in that month was 1,628 higher than anticipated.
-This mortality event is likely caused by the September 11 tragedy and
-the detection of this mortality event provides confidence in our
-detection of other anomalies.
+of 3.91, making these anomalies unlikely to be due to chance. The
+algorithm correctly identifies September 2001 as an additive outlier
+(2001:09 t=6.40) where mortality in that month was 1,628 higher than
+anticipated. This mortality event is likely caused by the September 11
+tragedy and the detection of this mortality event provides confidence in
+our detection of other anomalies.
 
@@
 What environmental conditions may have changed? These are the kinds of
-questions that arise from our analyses.
+questions that arise from our analyses. The purpose of our paper is not
+to answer these questions, but our findings underscore the need for more
+research that takes an abductive approach. By identifying anomalies
+through an inductive process, researchers can then look for underlying
+causes. Once those causes are identified, researchers can then use a
+deductive process to see if such events predict other (or future)
+anomalies.
 
-Contrast the mortality protection in New York with the enhanced
-mortality in New Hampshire (\textbf{\autoref{fig:mortnewhamp}b}). Here
-we detect two highly significant level shifts (LS) in the monthly
-mortality data, first in April 2010 and again in November 2014 (2010:04
-t=3.51; 2014:11 t=6.68). These anomalies suggest New Hampshire
-experienced 9,700 more deaths (+14\% more than expected) in a seven-year
-period beginning in early 2010. This is the single largest percentage
-mortality increase/decrease we detected among all states. Not
-coincidentally, NH has the second highest opioid-related mortality in
-the US \citep{beetham2019access} and it is reasonable that we detect
-this epidemic in our results.
+To see the potential for combinging abductive reasoning with causal
+inference, contrast the mortality protection in New York with the
+enhanced mortality in New Hampshire
+(\textbf{\autoref{fig:mortnewhamp}b}). In New Hampshire we detect two
+significant level shifts (LS) in the monthly mortality data, first in
+April 2010 and again in November 2014 (2010:04 t=3.51; 2014:11 t=6.68).
+These anomalies suggest New Hampshire experienced 9,700 more deaths
+(+14\% more than expected) in a seven-year period beginning in early
+2010. These are events not experienced by neighborhing states during the
+same time period, and this is the single largest percentage mortality
+increase/decrease we detected among all states. Not coincidentally, NH
+has the second highest opioid-related mortality in the US
+\citep{beetham2019access} and it is likely that we detect this epidemic
+in our results. Isolating these anomalies and testing them against
+opioid sales data might yeild intriguing results.
 
@@
 
-We begin with a clear temporary change (TC) outlier for fertility in
-Louisiana (\textbf{\autoref{fig:fertla}a}). Here we detect two TC
-outliers, nearly back-to-back in 2005 (2005:08 \emph{t} = -6; and
-2005:10 \emph{t} = -5), coinciding with the destruction of Hurricane
-Katrina from the same year. The migration associated with Hurricane
-Katrina has garnered most of the attention of demographers
+To illustrate how researchers can examine fertility anomalies, we begin
+with a clear temporary change (TC) outlier for fertility in Louisiana
+(\textbf{\autoref{fig:fertla}a}). Here we detect two TC outliers, nearly
+back-to-back in 2005 (2005:08 \emph{t} = -6; and 2005:10 \emph{t} = -5),
+coinciding with the destruction of Hurricane Katrina from the same year.
+The migration associated with Hurricane Katrina has garnered most of the
+attention of demographers
 \citep{fussellRecoveryMigrationCity2014, horiDisplacementDynamicsSouthern2009},
 but the hurricane had a very clear impact on fertility behaviors too,
-creating a mini ``baby bust'' in Louisiana. We estimate 4,411 fewer
-births in Louisiana compared to the counterfactual, likely attributable
-to the Hurricane.
+creating a mini ``baby bust'' in Louisiana with the departure of many
+people in their childbearing years. We estimate 4,411 fewer births in
+Louisiana compared to the counterfactual, likely attributable to the
+Hurricane.
 
@@
 occured just 11 months before we detect a shift toward lower fertility.
-We believe the trend toward lower fertility in August 2009 and the stock
-market crash in September 2008 are linked.
+It is possible that the trend toward lower fertility in August 2009 and
+the stock market crash in September 2008 are linked, although, as stated
+previously, we are speculating only to illustrate how causal inference
+could work in an abductive research agenda.
 

### 2020-01-24 MANUSCRIPT/manuscript.Rmd
All #'s should now be calculated in-text.

@@
 ---
-title: "Causal Inference in Population Trends: A Technique for finding Demographic Anomalies in Big Data"
+title: "Causal Inference in Population Trends: Searching for Demographic Anomalies in Big Data"
 
@@
 In demonstrating our technique, we uncover real spikes and dips in fertility and mortality over time, distinguished from regular trend variations.
-We show that 38 states exhibited at least one mortality anomaly over time, totalling more than 318,000 additional deaths, while an additional six states exhibited periods of decline resuling in 164,000 fewer deaths.
-Twelve states exhibited real, marked periods of declines in fertility, totalling more than 240,000 fewer births than expected, and eleven states had periods of excess births, resulting in an addition 134,000 births.
+We show that 36 states exhibited at least one mortality anomaly over time, totalling nearly 300,000 additional deaths, while an additional seven states exhibited periods of decline resuling in 250,000 fewer deaths.
+Ten states exhibited real, marked periods of declines in fertility, totalling more than 214,000 fewer births than expected, and 8 states had periods of excess births, resulting in an addition 12,000 births.
 Our results show how researchers can detect demographic anomalies in trend data and use those anomalies as a beginning point for further investigation into the causes."
@@
 ```{r SumStats, include=FALSE}
+```
+
+```{r, include = FALSE}
+
+## These calculations need to be checked. They're for the abstract.
+
+anom_mort <- read_rds("../R/DATA-PROCESSED/mortality_anomalies") %>%
+   na.omit %>%
+   mutate(component = "Mortality")
+
+ anom_fert <- read_rds("../R/DATA-PROCESSED/fertility_anomalies") %>%
+   na.omit %>%
+   mutate(component = "Fertility")
+
+abstract <- rbind(anom_mort, anom_fert) %>%
+   group_by(type, location, component) %>%
+   dplyr::summarise(Anomalies =  n()) %>%
+  ungroup() %>%
+  dplyr::select(location, component) %>%
+  group_by(component)
+
+nrow(distinct(abstract[which(abstract$component=="Mortality"),]))
+nrow(distinct(abstract[which(abstract$component=="Fertility"),]))
+
+abstractsums <- rbind(mortsum, fertsum)
+
+# Total # of states with mortality anomalies greater than 0
+nrow(abstractsums[which(abstractsums$component == "Mortality" & abstractsums$ydiff>0),])
+# Sums of Mortality anomalies greater than 0 
+sum(abstractsums$ydiff[which(abstractsums$component == "Mortality" & abstractsums$ydiff>0)])
+# Total # of states with mortality anomalies less than 0
+nrow(abstractsums[which(abstractsums$component == "Mortality" & abstractsums$ydiff<0),])
+# Sums of Mortality anomalies less than 0 
+sum(abstractsums$ydiff[which(abstractsums$component == "Mortality" & abstractsums$ydiff<0)])
+
+# Total # of states with mortality anomalies greater than 0
+nrow(abstractsums[which(abstractsums$component == "Fertility" & abstractsums$ydiff>0),])
+# Total # of states with 
+sum(abstractsums$ydiff[which(abstractsums$component == "Fertility" & abstractsums$ydiff>0)])
+nrow(abstractsums[which(abstractsums$component == "Fertility" & abstractsums$ydiff<0),])
+sum(abstractsums$ydiff[which(abstractsums$component == "Fertility" & abstractsums$ydiff<0)])
+
 ```
@@
 
-**\autoref{sumtable}** report the overall number of anomalies we detect of each type for births and deaths and some summary statistics across all anomaly types.
+**\autoref{sumtable}** report the overall number of anomalies we detect of each type for births and deaths and some summary statistics across all anomaly types and **\autoref{fig:summap}** maps these results.
 We find considerably more mortality anomalies (n = `r paste(sums[2,2])`) than fertility anomalies (n = `r paste(sums[1,2])`).
-Given that anomalies are always the produce of events [@Song2018ACM], the finding of more mortality than fertility anomalies is not surprising.
-Mortality is likely to spike in response to a catastrophic event (like an earthquake or terrorist attack) while the effects of a catastrophic event on fertility is less predictable.
+Given that anomalies are always the product of events [@Song2018ACM], finding more mortality than fertility anomalies is not surprising.
+Mortality is likely to spike in response to a catastrophic event (like an earthquake or terrorist attack) or due to a disease outbreak while the effects of a catastrophic event on fertility is less predictable.
 One reason for this is that fertility is linked to human decision-making more directly than mortality [@stein2014couples] which results in more varied outcomes, so, for example, researchers find that catastrophic weather events can increase childbearing among those who already have a child, but not among the childless [@Evans2008Hurricanebirth].
@@
 
-Consistent with more anomalies across the time lines, we find more anomalous deaths (`r paste(prettyNum(sums[2,6], digits = 0, big.mark = ","))`) than anomalous births (`r paste(prettyNum(sums[1,6], digits = 0, big.mark = ","))`).
+Consistent with more anomalies across the time series, we find more anomalous deaths (`r paste(prettyNum(sums[2,6], digits = 0, big.mark = ","))`) than anomalous births (`r paste(prettyNum(sums[1,6], digits = 0, big.mark = ","))`).
 Fertility anomalies overwhelmingly tend toward lower fertility, with only a few fertility anomalies yielding more births.
@@
 Conversely, mortality anomalies tend to be more evenly split between positive and negative anomalies, but tend toward more deaths rather than less deaths.
-Again, this is not surprising, as the kinds of events that increase death (e.g., plant closings are associated with increased opioid death [@Venkataramani2019]) are more common than the events that decrease death (e.g., cannabis legalization is associated with decreased opioid deaths [@Livingston2017Cannabis]) These anomalous deaths and births account for `r sums2$percent_anom[2]`% and `r sums2$percent_anom[1]`% over the entire time series, respectively.
+Again, this is not surprising, as the kinds of events that increase death (e.g., plant closings are associated with increased opioid death [@Venkataramani2019]) are more common than the events that decrease death (e.g., cannabis legalization is associated with decreased opioid deaths [@Livingston2017Cannabis]).
+These anomalous deaths and births account for `r sums2$percent_anom[2]`% and `r sums2$percent_anom[1]`% over the entire time series, respectively.
 
-**\autoref{fig:summap}** maps these results.
-Three states exhibited neither mortality nor fertility anomalies: Alaska, North Dakota, and South Dakota.
+As **\autoref{fig:summap}** shows, three states exhibited neither mortality nor fertility anomalies: Alaska, North Dakota, and South Dakota.
 Another nine states exhibited just a single anomaly (DE, DC, ID, IL, KS, NV, NM, UT, and WY).
@@
 ## Mortality 
-Since the purpose of our paper is to demonstrate how demographers can make use of causal inference technique, we will not discuss all of our findings in this paper.
+Since the purpose of our paper is to demonstrate how demographers can make use of causal inference technique, we will not discuss all of our findings here, in this paper.
 Instead, we will use a few cases for illustration.
 For mortality, we will use New York State as an example (**\autoref{fig:mortnewhamp}a**).
-We identify seven anomalies in the mortality time series for New York, all with t-statistics in excess of 3.91, making these anomalies unlikely to be due to chance.
-The algorithm correctly identifies September 2001 as an additive outlier (2001:09 t=6.40) where mortality in that month was 1,628 higher than anticipated.
+We identify seven anomalies in the mortality time series for New York, all with *t*-statistics in excess of 3.91, making these anomalies unlikely to be due to chance.
+The algorithm correctly identifies September 2001 as an additive outlier (2001:09 *t*= `r paste(prettyNum(outlier.mort.ny$outliers[2,5], digits = 4))`) where there were `r paste(prettyNum(outlier.mort.ny$outliers$coefhat[2], digits = 4, big.mark = ","))` more deaths in that month than anticipated.
 This mortality event is likely caused by the September 11 tragedy and the detection of this mortality event provides confidence in our detection of other anomalies.
 
-In **\autoref{fig:mortnewhamp}a**, notice the strong level shift (LS) that occurs in February 2004 (2004:02 t=5.59) which prevented almost 870 deaths per month.
-This shift totals more than 144,000 averted deaths compared to the counter-factual time series and is the single largest mortality protective anomaly among all states.
-This translates to 4.5% fewer deaths than expected over the time period.
+```{r, include = FALSE}
+
+ny_tstat <- prettyNum(outlier.mort.ny$outliers[3,5], digits = 4)
+ny_averteddeaths <- prettyNum(abs(outlier.mort.ny$outliers[3,4]), digits = 1)
+ny_averteddeathstot <- prettyNum(round(abs(outlier.mort.ny$outliers[3,4]*(228-outlier.mort.ny$outliers[3,2])),digits =-3), big.mark=",")
+ny_peraverted <- percent(abs(outlier.mort.ny$outliers[3,4]*(228-outlier.mort.ny$outliers[3,2]))/sum(outlier.mort.ny$y[62:228]))
+
+nh_tstat1 <- prettyNum(outlier.mort.nh$outliers[1,5], digits = 3)
+nh_tstat2 <- prettyNum(outlier.mort.nh$outliers[2,5], digits = 3)
+nh_totals <- prettyNum(sum(outlier.mort.nh$y - outlier.mort.nh$yadj), digits=1, big.mark=",")
+nh_per <- percent(1- sum(outlier.mort.nh$yadj[136:228])/ sum(outlier.mort.nh$y[136:228]))
+
+```
+
+In **\autoref{fig:mortnewhamp}a**, notice the strong level shift (LS) that occurs in February 2004 (2004:02 *t*= `r paste(ny_tstat)`) which prevented `r paste(ny_averteddeaths)` deaths per month.
+This shift totals more than `r paste(ny_averteddeathstot)` averted deaths compared to the counter-factual time series and is the single largest mortality protective anomaly among all states.
+This translates to `r ny_peraverted` fewer deaths than expected over the time period.
 What is driving this mortality protection?
@@
 
-To see the potential for combinging abductive reasoning with causal inference, contrast the mortality protection in New York with the enhanced mortality in New Hampshire (**\autoref{fig:mortnewhamp}b**).
-In New Hampshire we detect two significant level shifts (LS) in the monthly mortality data, first in April 2010 and again in November 2014 (2010:04 t=3.51; 2014:11 t=6.68).
-These anomalies suggest New Hampshire experienced 9,700 more deaths (+14% more than expected) in a seven-year period beginning in early 2010.
-These are events not experienced by neighborhing states during the same time period, and this is the single largest percentage mortality increase/decrease we detected among all states.
+To see the potential for combining abductive reasoning with causal inference, contrast the mortality protection in New York with the enhanced mortality in New Hampshire (**\autoref{fig:mortnewhamp}b**).
+In New Hampshire we detect two significant level shifts (LS) in the monthly mortality data, first in April 2010 and again in November 2014 (2010:04 *t*= `r nh_tstat1`; 2014:11 *t*= `r nh_tstat2`).
+These anomalies suggest New Hampshire experienced `r nh_totals` more deaths (+`r nh_per` more than expected) in a seven-year period beginning in early 2010.
+These are events not experienced by neighboring states during the same time period, and this is the single largest percentage mortality increase/decrease we detected among all states.
 Not coincidentally, NH has the second highest opioid-related mortality in the US [@beetham2019access] and it is likely that we detect this epidemic in our results.
-Isolating these anomalies and testing them against opioid sales data might yeild intriguing results.
+Isolating these anomalies and testing them against opioid sales data might yield intriguing results.
 
@@
 
+```{r, include = FALSE}
+la_tstat1 <- prettyNum(outlier.fert.la$outliers$tstat[1], digits = 4) 
+la_tstat2 <- prettyNum(outlier.fert.la$outliers$tstat[2], digits = 4)
+la_averbirths <- prettyNum(abs(sum(outlier.fert.la$y - outlier.fert.la$yadj)), digits = 0, big.mark = ",")
+
+conn_tstat <- prettyNum(outlier.fert.conn$outliers$tstat, digits = 3)
+conn_births <- prettyNum(outlier.fert.conn$outliers$coefhat, digits = 0)
+conn_tot <- prettyNum(sum(outlier.fert.conn$y - outlier.fert.conn$yadj), digits = 0, big.mark = ",")
+conn_per <- percent(abs(sum(outlier.fert.conn$y - outlier.fert.conn$yadj) / sum(outlier.fert.conn$yadj[80:180])))
+```
+
 To illustrate how researchers can examine fertility anomalies, we begin with a clear temporary change (TC) outlier for fertility in Louisiana (**\autoref{fig:fertla}a**).
-Here we detect two TC outliers, nearly back-to-back in 2005 (2005:08 *t* = `r prettyNum(outlier.fert.la$outliers$tstat[1], digits = 3)`; and 2005:10 *t* = `r prettyNum(outlier.fert.la$outliers$tstat[2], digits = 3)`), coinciding with the destruction of Hurricane Katrina from the same year.
+Here we detect two TC outliers, nearly back-to-back in 2005 (2005:08 *t* = `r la_tstat1`; and 2005:10 *t* = `r la_tstat2`), coinciding with the destruction of Hurricane Katrina from the same year.
 The migration associated with Hurricane Katrina has garnered most of the attention of demographers [@fussellRecoveryMigrationCity2014;@horiDisplacementDynamicsSouthern2009], but the hurricane had a very clear impact on fertility behaviors too, creating a mini "baby bust" in Louisiana with the departure of many people in their childbearing years.
-We estimate `r prettyNum(abs(sum(outlier.fert.la$y - outlier.fert.la$yadj)), digits = 0, big.mark = ",")` fewer births in Louisiana compared to the counterfactual, likely attributable to the Hurricane.
+We estimate `r la_averbirths` fewer births in Louisiana compared to the counterfactual, likely attributable to the Hurricane.
 
 We also highlight a level shift (LS) outlier for fertility in Connecticut.
-Here we detect a strong (*t*= `r prettyNum(outlier.fert.conn$outliers$tstat, digits = 3)`; `r prettyNum(outlier.fert.conn$outliers$coefhat, digits = 0)` births/month) level shift toward lower fertility beginning in August 2009 (2009:08) that continues through the end of the period (**\autoref{fig:fertla}b**).
-This LS toward lower fertility reduced the # of births by `r prettyNum(sum(outlier.fert.conn$y - outlier.fert.conn$yadj), digits = 0, big.mark = ",")` or `r percent(sum(outlier.fert.conn$y - outlier.fert.conn$yadj) / sum(outlier.fert.conn$yadj[80:180]))` lower than the counterfactual time series.
-The stock market crash of 2008, where the Dow Jones Industrial Average had the largest single-day loss up to that point, occured just 11 months before we detect a shift toward lower fertility.
+Here we detect a strong (*t*= `r conn_tstat` for `r conn_births` births/month) level shift toward lower fertility beginning in August 2009 (2009:08) that continues through the end of the period (**\autoref{fig:fertla}b**).
+This LS toward lower fertility reduced the # of births by `r conn_tot` or `r conn_per`) lower than the counterfactual time series.
+The stock market crash of 2008, where the Dow Jones Industrial Average had the largest single-day loss up to that point, occurred just 11 months before we detect a shift toward lower fertility.
 It is possible that the trend toward lower fertility in August 2009 and the stock market crash in September 2008 are linked, although, as stated previously, we are speculating only to illustrate how causal inference could work in an abductive research agenda.
@@
 
+```{r, include = FALSE}
+hi_tstat <- prettyNum(outlier.fert.hi$outliers$tstat, digits = 3)
+hi_per <- percent(1- outlier.fert.hi$y[137]/outlier.fert.hi$yadj[137])
+
+oh_tstat <- prettyNum(outlier.mort.oh$outliers$tstat[7], digits = 3)
+oh_deaths <- prettyNum(round(abs(outlier.mort.oh$outliers[7,4]), digits =0), big.mark=",")
+oh_tot <- prettyNum(abs(round(outlier.mort.oh$outliers[7,4]*(228-194), digits = -4)), big.mark=",")
+```
+
 In **\autoref{fig:ferthawaii}a** we identify a single additive anomaly in Hawaiian fertility in May 2014.
-This is a strong anomaly with a *t*-statistic of `r prettyNum(outlier.fert.hi$outliers$tstat, digits = 3)`, 14% above the counter-factual time series.
+This is a strong anomaly with a *t*-statistic of `r hi_tstat`, `r hi_per` above the counter-factual time series.
 This single, anomalous month is also the second highest monthly births in the time series.
@@
 
-This is contrast to **\autoref{fig:ferthawaii}b**, where we identify a strange mortality reduction in Ohio (*t*-stat: `r prettyNum(outlier.mort.oh$outliers$tstat[7], digits = 3)`).
-This LS is more than 1,100 deaths per month less than the counterfactual time series, suggesting had nearly 40,000 fewer deaths since February 2015 than expected.
+This is contrast to **\autoref{fig:ferthawaii}b**, where we identify a strange mortality reduction in Ohio (*t*-stat: `r oh_tstat`).
+This LS is more than `r oh_deaths` deaths per month less than the counterfactual time series, suggesting had nearly `r oh_tot` fewer deaths since February 2015 than expected.
 This is the single-largest LS among all states.

### 2020-01-24 MANUSCRIPT/manuscript.tex
All #'s should now be calculated in-text.

@@
 {
-  \title{\bf Causal Inference in Population Trends: A Technique for finding
-Demographic Anomalies in Big Data}
+  \title{\bf Causal Inference in Population Trends: Searching for Demographic
+Anomalies in Big Data}
 
@@
   \begin{center}
-    {\LARGE\bf Causal Inference in Population Trends: A Technique for finding
-Demographic Anomalies in Big Data}
+    {\LARGE\bf Causal Inference in Population Trends: Searching for Demographic
+Anomalies in Big Data}
   \end{center}
@@
 we uncover real spikes and dips in fertility and mortality over time,
-distinguished from regular trend variations. We show that 38 states
-exhibited at least one mortality anomaly over time, totalling more than
-318,000 additional deaths, while an additional six states exhibited
-periods of decline resuling in 164,000 fewer deaths. Twelve states
+distinguished from regular trend variations. We show that 36 states
+exhibited at least one mortality anomaly over time, totalling nearly
+300,000 additional deaths, while an additional seven states exhibited
+periods of decline resuling in 250,000 fewer deaths. Ten states
 exhibited real, marked periods of declines in fertility, totalling more
-than 240,000 fewer births than expected, and eleven states had periods
-of excess births, resulting in an addition 134,000 births. Our results
-show how researchers can detect demographic anomalies in trend data and
-use those anomalies as a beginning point for further investigation into
-the causes.
+than 214,000 fewer births than expected, and 8 states had periods of
+excess births, resulting in an addition 12,000 births. Our results show
+how researchers can detect demographic anomalies in trend data and use
+those anomalies as a beginning point for further investigation into the
+causes.
 \end{abstract}
@@
 
-\caption{\label{tab:unnamed-chunk-1}\textbf{Summary by Demographic Component.}  refers to the number of anomolies for each type and demographic component, States refers to the number of unique states exhibiting that specific Type-demographic component.
+\caption{\label{tab:unnamed-chunk-2}\textbf{Summary by Demographic Component.}  refers to the number of anomolies for each type and demographic component, States refers to the number of unique states exhibiting that specific Type-demographic component.
 \label{sumtable}}
@@
 detect of each type for births and deaths and some summary statistics
-across all anomaly types. We find considerably more mortality anomalies
-(n = 156) than fertility anomalies (n = 22). Given that anomalies are
-always the produce of events \citep{Song2018ACM}, the finding of more
-mortality than fertility anomalies is not surprising. Mortality is
-likely to spike in response to a catastrophic event (like an earthquake
-or terrorist attack) while the effects of a catastrophic event on
-fertility is less predictable. One reason for this is that fertility is
-linked to human decision-making more directly than mortality
-\citep{stein2014couples} which results in more varied outcomes, so, for
-example, researchers find that catastrophic weather events can increase
-childbearing among those who already have a child, but not among the
-childless \citep{Evans2008Hurricanebirth}. Some events, like the 1995
-Oklahoma City bombings, resulted in both fertility and mortality
-changes. However, changes in fertility manifest over a longer time
-horizon after an event, and the relationship between a
+across all anomaly types and \textbf{\autoref{fig:summap}} maps these
+results. We find considerably more mortality anomalies (n = 156) than
+fertility anomalies (n = 22). Given that anomalies are always the
+product of events \citep{Song2018ACM}, finding more mortality than
+fertility anomalies is not surprising. Mortality is likely to spike in
+response to a catastrophic event (like an earthquake or terrorist
+attack) or due to a disease outbreak while the effects of a catastrophic
+event on fertility is less predictable. One reason for this is that
+fertility is linked to human decision-making more directly than
+mortality \citep{stein2014couples} which results in more varied
+outcomes, so, for example, researchers find that catastrophic weather
+events can increase childbearing among those who already have a child,
+but not among the childless \citep{Evans2008Hurricanebirth}. Some
+events, like the 1995 Oklahoma City bombings, resulted in both fertility
+and mortality changes. However, changes in fertility manifest over a
+longer time horizon after an event, and the relationship between a
 fertility-inducing event and behavior change is less strong than the
@@
 
-Consistent with more anomalies across the time lines, we find more
+Consistent with more anomalies across the time series, we find more
 anomalous deaths (627,364) than anomalous births (226,476). Fertility
@@
 that decrease death (e.g., cannabis legalization is associated with
-decreased opioid deaths \citep{Livingston2017Cannabis}) These anomalous
+decreased opioid deaths \citep{Livingston2017Cannabis}). These anomalous
 deaths and births account for 1.44\% and 0.889\% over the entire time
@@
 
-\textbf{\autoref{fig:summap}} maps these results. Three states exhibited
-neither mortality nor fertility anomalies: Alaska, North Dakota, and
-South Dakota. Another nine states exhibited just a single anomaly (DE,
-DC, ID, IL, KS, NV, NM, UT, and WY). Both New York and Massachusetts
-exhibited the most anomalies with nine.
+As \textbf{\autoref{fig:summap}} shows, three states exhibited neither
+mortality nor fertility anomalies: Alaska, North Dakota, and South
+Dakota. Another nine states exhibited just a single anomaly (DE, DC, ID,
+IL, KS, NV, NM, UT, and WY). Both New York and Massachusetts exhibited
+the most anomalies with nine.
 
@@
 make use of causal inference technique, we will not discuss all of our
-findings in this paper. Instead, we will use a few cases for
+findings here, in this paper. Instead, we will use a few cases for
 illustration. For mortality, we will use New York State as an example
 (\textbf{\autoref{fig:mortnewhamp}a}). We identify seven anomalies in
-the mortality time series for New York, all with t-statistics in excess
-of 3.91, making these anomalies unlikely to be due to chance. The
+the mortality time series for New York, all with \emph{t}-statistics in
+excess of 3.91, making these anomalies unlikely to be due to chance. The
 algorithm correctly identifies September 2001 as an additive outlier
-(2001:09 t=6.40) where mortality in that month was 1,628 higher than
-anticipated. This mortality event is likely caused by the September 11
-tragedy and the detection of this mortality event provides confidence in
-our detection of other anomalies.
+(2001:09 \emph{t}= 6.396) where there were 1,628 more deaths in that
+month than anticipated. This mortality event is likely caused by the
+September 11 tragedy and the detection of this mortality event provides
+confidence in our detection of other anomalies.
 
 In \textbf{\autoref{fig:mortnewhamp}a}, notice the strong level shift
-(LS) that occurs in February 2004 (2004:02 t=5.59) which prevented
-almost 870 deaths per month. This shift totals more than 144,000 averted
-deaths compared to the counter-factual time series and is the single
-largest mortality protective anomaly among all states. This translates
-to 4.5\% fewer deaths than expected over the time period. What is
-driving this mortality protection? What policies did NY put into place
-that might have contributed to this considerable mortality reduction?
-What environmental conditions may have changed? These are the kinds of
-questions that arise from our analyses. The purpose of our paper is not
-to answer these questions, but our findings underscore the need for more
-research that takes an abductive approach. By identifying anomalies
-through an inductive process, researchers can then look for underlying
-causes. Once those causes are identified, researchers can then use a
-deductive process to see if such events predict other (or future)
-anomalies.
+(LS) that occurs in February 2004 (2004:02 \emph{t}= -5.594) which
+prevented 869 deaths per month. This shift totals more than 144,000
+averted deaths compared to the counter-factual time series and is the
+single largest mortality protective anomaly among all states. This
+translates to 6.90\% fewer deaths than expected over the time period.
+What is driving this mortality protection? What policies did NY put into
+place that might have contributed to this considerable mortality
+reduction? What environmental conditions may have changed? These are the
+kinds of questions that arise from our analyses. The purpose of our
+paper is not to answer these questions, but our findings underscore the
+need for more research that takes an abductive approach. By identifying
+anomalies through an inductive process, researchers can then look for
+underlying causes. Once those causes are identified, researchers can
+then use a deductive process to see if such events predict other (or
+future) anomalies.
 
-To see the potential for combinging abductive reasoning with causal
+To see the potential for combining abductive reasoning with causal
 inference, contrast the mortality protection in New York with the
@@
 significant level shifts (LS) in the monthly mortality data, first in
-April 2010 and again in November 2014 (2010:04 t=3.51; 2014:11 t=6.68).
-These anomalies suggest New Hampshire experienced 9,700 more deaths
-(+14\% more than expected) in a seven-year period beginning in early
-2010. These are events not experienced by neighborhing states during the
-same time period, and this is the single largest percentage mortality
-increase/decrease we detected among all states. Not coincidentally, NH
-has the second highest opioid-related mortality in the US
-\citep{beetham2019access} and it is likely that we detect this epidemic
-in our results. Isolating these anomalies and testing them against
-opioid sales data might yeild intriguing results.
+April 2010 and again in November 2014 (2010:04 \emph{t}= 3.51; 2014:11
+\emph{t}= 6.68). These anomalies suggest New Hampshire experienced 8,159
+more deaths (+9.24\% more than expected) in a seven-year period
+beginning in early 2010. These are events not experienced by neighboring
+states during the same time period, and this is the single largest
+percentage mortality increase/decrease we detected among all states. Not
+coincidentally, NH has the second highest opioid-related mortality in
+the US \citep{beetham2019access} and it is likely that we detect this
+epidemic in our results. Isolating these anomalies and testing them
+against opioid sales data might yield intriguing results.
 
@@
 (\textbf{\autoref{fig:fertla}a}). Here we detect two TC outliers, nearly
-back-to-back in 2005 (2005:08 \emph{t} = -6; and 2005:10 \emph{t} = -5),
-coinciding with the destruction of Hurricane Katrina from the same year.
-The migration associated with Hurricane Katrina has garnered most of the
-attention of demographers
+back-to-back in 2005 (2005:08 \emph{t} = -6.002; and 2005:10 \emph{t} =
+-4.997), coinciding with the destruction of Hurricane Katrina from the
+same year. The migration associated with Hurricane Katrina has garnered
+most of the attention of demographers
 \citep{fussellRecoveryMigrationCity2014, horiDisplacementDynamicsSouthern2009},
@@
 We also highlight a level shift (LS) outlier for fertility in
-Connecticut. Here we detect a strong (\emph{t}= -3.71; -285
+Connecticut. Here we detect a strong (\emph{t}= -3.71 for -285
 births/month) level shift toward lower fertility beginning in August
@@
 (\textbf{\autoref{fig:fertla}b}). This LS toward lower fertility reduced
-the \# of births by -28,752 or -8.57\% lower than the counterfactual
+the \# of births by -28,752 or 8.57\%) lower than the counterfactual
 time series. The stock market crash of 2008, where the Dow Jones
 Industrial Average had the largest single-day loss up to that point,
-occured just 11 months before we detect a shift toward lower fertility.
+occurred just 11 months before we detect a shift toward lower fertility.
 It is possible that the trend toward lower fertility in August 2009 and
@@
 anomaly in Hawaiian fertility in May 2014. This is a strong anomaly with
-a \emph{t}-statistic of 4.96, 14\% above the counter-factual time
+a \emph{t}-statistic of 4.96, -14.9\% above the counter-factual time
 series. This single, anomalous month is also the second highest monthly
@@
 identify a strange mortality reduction in Ohio (\emph{t}-stat: -4.27).
-This LS is more than 1,100 deaths per month less than the counterfactual
+This LS is more than 1,157 deaths per month less than the counterfactual
 time series, suggesting had nearly 40,000 fewer deaths since February

### 2020-01-24 MANUSCRIPT/manuscript.Rmd
Typo corrections.

@@
 
-We report outliers using a simple t-statistic and report the size of the outlier, or its impact on the time series, by subtracting $\hat{y}$, the adjusted time series without outliers removed, from $y$, the original, unadjusted time series.
+We report outliers using a simple t-statistic and report the size of the outlier, or its impact on the time series, by subtracting $\hat{y}$, the adjusted time series with outliers removed, from $y$, the original, unadjusted time series.
 

### 2020-01-24 MANUSCRIPT/manuscript.tex
Typo corrections.

@@
 outlier, or its impact on the time series, by subtracting \(\hat{y}\),
-the adjusted time series without outliers removed, from \(y\), the
+the adjusted time series with outliers removed, from \(y\), the
 original, unadjusted time series.

### 2020-01-28 MANUSCRIPT/manuscript.Rmd
Added Stephanie's conclusion.

@@
 We find considerably more mortality anomalies (n = `r paste(sums[2,2])`) than fertility anomalies (n = `r paste(sums[1,2])`).
-Given that anomalies are always the product of events [@Song2018ACM], finding more mortality than fertility anomalies is not surprising.
+Given that anomalies are always the product of events [@song2018anomaly], finding more mortality than fertility anomalies is not surprising.
 Mortality is likely to spike in response to a catastrophic event (like an earthquake or terrorist attack) or due to a disease outbreak while the effects of a catastrophic event on fertility is less predictable.
@@
 
+# Conclusion
+Data scientists frequently claim that the big data revolution is a turning point in scientific discovery that will allow us to solve some of the world’s most pressing problems [@grimmer2015ppsp].
+Social scientists are skeptical of such claims, because they better understand the complexities of the social world and know from experience that data, alone, is not enough [@bohon2018demography;@grimmer2015ppsp].
+Nonetheless, the increased availability of data and (more importantly, we argue) the development of advanced techniques for analyzing these data will enable important discovery [@monroe2015no].
+
+One technique that population scientists underutilize is causal inference.
+Causal inference, in the simplest terms, is the discovery of effects in search of a cause using big data and advanced computing algorithms [@imai2008misunderstandings].
+This inductive approach is uncommon in quantitative social science where hypothesis testing is expected, and approaches are largely deductive.
+However, common statistical hypothesis testing is impractical with big data, as significant p values are guaranteed, and such approaches do not allow us to uncover all the information that big data has to offer [@monroe2015no].
+Here, we call for an abductive approach, where causal inference algorithms are applied to high quality data to uncover irregularities that are unlikely to be attributable to expected variations in trends (or noise) as a first step to then developing testable hypotheses about causes.
+Abductive approaches allow researchers to move from the inductive to the deductive and sometimes work back and forth in aid of scientific discovery.
+
+In this paper, we show how the tsoutlier package in R can be implemented to conduct statistical time series outlier detection, an inductive approach that aids in the creation of deductive reasoning.
+This algorithm is one of many causal inference approaches that are freely and commercially available (see @rcausalimpact).
+In our initial work, we experimented with other approaches, such as Google’s causalimpact algorithm, which uncovered the same patterns we briefly discuss in this paper.
+By making more use of causal inference techniques and abductive modeling approaches, we argue that social scientists will be able to better understand how events or policy implementation can impact important outcomes.
+For example, we could uncover—on a wide scale—how gun control policies may reduce or increase injuries from shootings or how marijuana legalization might impact opioid deaths.
+We could also uncover how increases in extreme weather events impact a range of behaviors such as home sales, bottled water purchases, and even fertility.
+
+In our demonstration, we show how the application of causal inference to state-level time series fertility and mortality data uncovers three types of demographic anomalies:  those that occur and disappear quickly, those that occur and decay over time, and those that occur and remain.
+Uncovering these anomalies in and of themselves is important.
+For example, the algorithm we deploy clearly shows the fertility effects of Hurricanes Katrina and Rita as well as the mortality effects of the World Trade Center collapse on September 11, 2001.
+The ability to identify and differentiate types of anomalies is even more important, as we can potentially see how some policies might impact outcomes permanently and some might have an effect that is short-lived.
+Differentiating these types gives us greater insight into short- and long-term solutions to social problems.
+We urge social scientists to begin to use causal inference algorithms and other big data techniques, and we hope that this demonstration will illustrate their usefulness to the social science enterprise.
+
 \newpage

### 2020-01-28 MANUSCRIPT/manuscript.tex
Added Stephanie's conclusion.

@@
 fertility anomalies (n = 22). Given that anomalies are always the
-product of events \citep{Song2018ACM}, finding more mortality than
+product of events \citep{song2018anomaly}, finding more mortality than
 fertility anomalies is not surprising. Mortality is likely to spike in
@@
 
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
+Data scientists frequently claim that the big data revolution is a
+turning point in scientific discovery that will allow us to solve some
+of the world's most pressing problems \citep{grimmer2015ppsp}. Social
+scientists are skeptical of such claims, because they better understand
+the complexities of the social world and know from experience that data,
+alone, is not enough \citep{bohon2018demography, grimmer2015ppsp}.
+Nonetheless, the increased availability of data and (more importantly,
+we argue) the development of advanced techniques for analyzing these
+data will enable important discovery \citep{monroe2015no}.
+
+One technique that population scientists underutilize is causal
+inference. Causal inference, in the simplest terms, is the discovery of
+effects in search of a cause using big data and advanced computing
+algorithms \citep{imai2008misunderstandings}. This inductive approach is
+uncommon in quantitative social science where hypothesis testing is
+expected, and approaches are largely deductive. However, common
+statistical hypothesis testing is impractical with big data, as
+significant p values are guaranteed, and such approaches do not allow us
+to uncover all the information that big data has to offer
+\citep{monroe2015no}. Here, we call for an abductive approach, where
+causal inference algorithms are applied to high quality data to uncover
+irregularities that are unlikely to be attributable to expected
+variations in trends (or noise) as a first step to then developing
+testable hypotheses about causes. Abductive approaches allow researchers
+to move from the inductive to the deductive and sometimes work back and
+forth in aid of scientific discovery.
+
+In this paper, we show how the tsoutlier package in R can be implemented
+to conduct statistical time series outlier detection, an inductive
+approach that aids in the creation of deductive reasoning. This
+algorithm is one of many causal inference approaches that are freely and
+commercially available (see \citet{rcausalimpact}). In our initial work,
+we experimented with other approaches, such as Google's causalimpact
+algorithm, which uncovered the same patterns we briefly discuss in this
+paper. By making more use of causal inference techniques and abductive
+modeling approaches, we argue that social scientists will be able to
+better understand how events or policy implementation can impact
+important outcomes. For example, we could uncover---on a wide
+scale---how gun control policies may reduce or increase injuries from
+shootings or how marijuana legalization might impact opioid deaths. We
+could also uncover how increases in extreme weather events impact a
+range of behaviors such as home sales, bottled water purchases, and even
+fertility.
+
+In our demonstration, we show how the application of causal inference to
+state-level time series fertility and mortality data uncovers three
+types of demographic anomalies: those that occur and disappear quickly,
+those that occur and decay over time, and those that occur and remain.
+Uncovering these anomalies in and of themselves is important. For
+example, the algorithm we deploy clearly shows the fertility effects of
+Hurricanes Katrina and Rita as well as the mortality effects of the
+World Trade Center collapse on September 11, 2001. The ability to
+identify and differentiate types of anomalies is even more important, as
+we can potentially see how some policies might impact outcomes
+permanently and some might have an effect that is short-lived.
+Differentiating these types gives us greater insight into short- and
+long-term solutions to social problems. We urge social scientists to
+begin to use causal inference algorithms and other big data techniques,
+and we hope that this demonstration will illustrate their usefulness to
+the social science enterprise.
+
 \newpage

### 2020-01-28 MANUSCRIPT/manuscript.Rmd
Newest edits. Full paper.
I made the example figure in excel because it was faster.

@@
 abstract: "The proliferation of big data, wider access to advanced computing platforms, and the development of powerful statistical algorithms can uncover hidden anomalies in demographic data that have previously been dismissed as noise.
-Here, we demonstrate how demographers can use causal inference techniques and abductive reasoning to identify fertility and mortality anamolies in state-level data trends by investigating demographic time series data since 1999.
+Here, we demonstrate how social scientists can use causal inference techniques and abductive reasoning to identify fertility and mortality anamolies in state-level data trends by investigating demographic time series data since 1999.
 In demonstrating our technique, we uncover real spikes and dips in fertility and mortality over time, distinguished from regular trend variations.
@@
 
-Causal inference, in its most common usage, is an attempt to uncover the underlying mechanism that results in changes in a phenomenon and have a long history approaches in social sciences [@grimmer2015ppsp].
+Causal inference, in its most common usage, is an attempt to uncover the underlying mechanisms that result in changes in a phenomenon and has a long history as an approach to inquiry in the social sciences [@grimmer2015ppsp].
 Often, the identification of a casual mechanism requires either a randomized control trial (RCT) or expert knowledge applied to a natural experiment.
 For population-level analysis, expensive RCTs usually preclude their widespread adoption [@west2008ajph].
-Using natural experiments provide less certainty about causation and fewer opportunities to replicate work but they provide important and widespread insights such as the determinants of migration, however, provides important and more widespread insights on the determinants of migration after Hurricane Katrina in 2006 [@fussellRecoveryMigrationCity2014;@horiDisplacementDynamicsSouthern2009], mini baby booms after electrical blackouts [@fetzer2018jpe], and the highly publicized estimates of excess mortality after Hurricane Maria in Puerto Rico in 2016 [@kishore2018mortality;@santos2018use].
-Both RCTs and natural experiments follow traditional, hypothesis testing, inductive scientific paradigms where a question is first posed and scientists generate or find data to answer the question.
-Such approaches, while extensively utilized and time-tested, are likely to miss important phenomena that might go unnoticed.
+Using natural experiments provide less certainty about causation and fewer opportunities to replicate work but they provide important and widespread insights.
+
+<!-- such as the determinants of migration, however, provides important and more widespread insights on the determinants of migration after Hurricane Katrina in 2006 [@fussellRecoveryMigrationCity2014;@horiDisplacementDynamicsSouthern2009], mini baby booms after electrical blackouts [@fetzer2018jpe], and the highly publicized estimates of excess mortality after Hurricane Maria in Puerto Rico in 2016 [@kishore2018mortality;@santos2018use].
+-->
 
@@
 
-In this paper, we use modern statistical outlier detection algorithms [@chen1993joint] on nearly twenty years of mortality and fertility data at the US state-level to identify anomalous demographic behavior.
+In this paper, we use modern statistical outlier detection algorithms [@chen1993joint] as a means to illustrate one way that social scientists can engage in discovery using big data techniques on nearly twenty years of mortality and fertility data at the US state-level to identify anomalous demographic behavior.
 In essence, we identify *effects without knowledge of the cause*.
@@
 We do not necessarily know the causes of these anomalies but identifying them allows scientists with more detailed knowledge of local population dynamics, state-level policy making, or macro-economics to explain these phenomena post-hoc.
-From explanations of phenomena that may have previously gone unnoticed, demographers may be able to better forecast populations and provide policy solutions for impending problems.
+From explanations of phenomena that may have previously gone unnoticed, social scientists may be able to better forecast populations and provide policy solutions for impending problems such as climate change.
 
@@
 
+![**"Toy" example of anomaly detection using the exchange rate between the Euro and the Swiss Franc.** The Swiss Franc was pegged to the Euro at 1.2 until January 15 2015 when the peg was removed and the currency was allowed to trade more freely.
+This is a very strong anomly in the time series (*t*-stat = -144.07) detectable using statistical methods and visually.
+The countefactural estimate with the outlier removed would simply have kept the currency exchange at 1.2 ($\hat{y}$).
+\label{explanationfigure}](./manuscript_files/figure-latex/Figure1.png)
+
 where $\pi(B)=\sum_{i=o}^{inf} \pi_iB^i$.
-The identification of outliers then involves a three step process to (1) identify all potential outliers, $t_j$ and $L_j (B)$, (2) joint estimates of model parameters and outlier effects are computed to identify potentially spurious outliers, and (3) the outliers and effects are re-estimated without spurious outliers.
-Of importance here is step 2, which relies on some critical value above which an outlier at time point m is considered spurious.
+The identification of outliers then involves a three step process.
+First, (1) the algorithm identifies all potential outliers, $t_j$ and $L_j (B)$.
+Next, (2) we compute joint estimates of model parameters and outlier effects to identify potentially spurious outliers.
+Finally, (3) the outliers and effects are re-estimated without spurious outliers.
+Of importance here is step 2, which relies on some critical value above which an outlier at time point $m$ is considered spurious.
 Based on Chen and Liu’s [-@chen1993joint] recommendation, we set a critical value of 3.5 which generally minimizes the possibility of Type I errors or false positive outliers.
 
+```{r, include = FALSE }
+set.seed(1)
+
+stockdat <- tribble(
+  ~Date, ~Value,
+'2014-12-27',	1.203880343,
+'2014-12-28',	1.202523908,
+'2014-12-29',	1.202714098,
+'2014-12-30',	1.201988512,
+'2014-12-31',	1.202535102,
+'2015-01-01',	1.201727726,
+'2015-01-02',	1.201794127,
+'2015-01-04',	1.20132752,
+'2015-01-05',	1.20100032,
+'2015-01-06',	1.20071649,
+'2015-01-07',	1.200762136,
+'2015-01-08',	1.200787933,
+'2015-01-09',	1.20079581,
+'2015-01-11',	1.200989867,
+'2015-01-12',	1.2006267,
+'2015-01-13',	1.20058359,
+'2015-01-14',	1.200605219,
+'2015-01-15',	0.99417984,
+'2015-01-16',	0.992258896,
+'2015-01-18',	0.992353824,
+'2015-01-19',	1.019880601,
+'2015-01-20',	1.00940244,
+'2015-01-21',	0.999117945,
+'2015-01-22',	0.988725725,
+'2015-01-23',	0.939377117,
+'2015-01-24',	0.987456601,
+'2015-01-25',	0.980699544,
+'2015-01-26',	1.013809279,
+'2015-01-27',	1.02203072,
+'2015-01-28',	1.0234588,
+'2015-01-29',	1.0455615
+)
+dat3 <- ts(stockdat$Value)
+stockdat_outlier <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)
+```
+
 We report outliers using a simple t-statistic and report the size of the outlier, or its impact on the time series, by subtracting $\hat{y}$, the adjusted time series with outliers removed, from $y$, the original, unadjusted time series.
+
+**\autoref{explanationfigure}** shows a toy example for anomaly detection in a time series using the classic example of the exchange rate between the Euro and the Swiss Franc (CHF).
+On January 15 2015, the Swiss National Bank removed the currency peg of 1.20 francs per Euro, exposing the Franc to the volatility of the currency market.
+The Franc immediately began trading a reduced rate compared to the Euro.
+What would have been the Euro/CHF exchange rate had the Swiss National Bank *not* removed the peg?
+A simple counter-factual estimate would be to keep the exchange rate at 1.20 ($\hat{y}$, sometimes called a "synthetic control" [@abadie2010synthetic]).
+
+In the Swiss Franc example, we have knowledge of the Swiss Bank's activities after the fact or ex-post-facto to to create the counter-factual time series $\hat{y}$ of 1.20.
+But is this anomaly detectable without knowledge of the Swiss Bank's activities?
+In other words, can we detect the reduction of the exchange rate on January 15 using only the time series?
+Absolutely.
+the tsoutliers package identifies January 15 2015 as an extremely strong level-shift outlier (*t*-stat = `r round( stockdat_outlier$outliers$tstat[1], digits = 2)`).
+The real world is hardly this simplified where a direct intervention is known and is testable.
 
@@
        escape = FALSE,
-       caption = "\\textbf{Summary by Demographic Component.}  refers to the number of anomolies for each type and demographic component, States refers to the number of unique states exhibiting that specific Type-demographic component.
+       caption = "\\textbf{Summary by Demographic Component.}  Here we can see there 22 Fertility and 156 Mortality anomalies among the state-level time series, totalling more than 200k anomalous births and 600k anomalous deaths.
 \\label{sumtable}") %>%
@@
 
-**\autoref{sumtable}** report the overall number of anomalies we detect of each type for births and deaths and some summary statistics across all anomaly types and **\autoref{fig:summap}** maps these results.
+**\autoref{sumtable}** reports the overall number of anomalies we detect of each type for births and deaths and some summary statistics across all anomaly types and **\autoref{fig:summap}** maps these results.
 We find considerably more mortality anomalies (n = `r paste(sums[2,2])`) than fertility anomalies (n = `r paste(sums[1,2])`).
@@
 Of course, not all events that impact fertility and mortality are catastrophic.
-Researchers have documented that more commonplace events such as massive layoff [@Venkataramani2019] or policy changes [@Livingston2017Cannabis] can also create anomalies in mortality patterns, but we have not found such evidence for fertility.
+Researchers have documented that more commonplace events such as massive layoffs [@Venkataramani2019] or policy changes [@Livingston2017Cannabis] can also create anomalies in mortality patterns, but we have not found such evidence for fertility.
 
@@
 This is consistent with research findings that link high severity weather events to significantly lower fertility [@Evans2008Hurricanebirth].
-Conversely, mortality anomalies tend to be more evenly split between positive and negative anomalies, but tend toward more deaths rather than less deaths.
+Conversely, mortality anomalies tend to be more evenly split between positive and negative anomalies, but tend toward more deaths rather than fewer deaths.
 Again, this is not surprising, as the kinds of events that increase death (e.g., plant closings are associated with increased opioid death [@Venkataramani2019]) are more common than the events that decrease death (e.g., cannabis legalization is associated with decreased opioid deaths [@Livingston2017Cannabis]).
@@
 ## Mortality 
-Since the purpose of our paper is to demonstrate how demographers can make use of causal inference technique, we will not discuss all of our findings here, in this paper.
+Since the purpose of our paper is to demonstrate how social scientists can make use of causal inference technique, we will not discuss all of our findings in this paper.
 Instead, we will use a few cases for illustration.
@@
 The algorithm correctly identifies September 2001 as an additive outlier (2001:09 *t*= `r paste(prettyNum(outlier.mort.ny$outliers[2,5], digits = 4))`) where there were `r paste(prettyNum(outlier.mort.ny$outliers$coefhat[2], digits = 4, big.mark = ","))` more deaths in that month than anticipated.
-This mortality event is likely caused by the September 11 tragedy and the detection of this mortality event provides confidence in our detection of other anomalies.
+This mortality event is likely caused by the September 11 terrorist attack on the World Trade Center that immediately killed 2,606 people and the detection of this mortality event provides confidence in our detection of other anomalies.
 
@@
 What policies did NY put into place that might have contributed to this considerable mortality reduction?
-What environmental conditions may have changed?
+What environmental or economic conditions may have changed may have changed?
 These are the kinds of questions that arise from our analyses.
@@
 conn_births <- prettyNum(outlier.fert.conn$outliers$coefhat, digits = 0)
-conn_tot <- prettyNum(sum(outlier.fert.conn$y - outlier.fert.conn$yadj), digits = 0, big.mark = ",")
+conn_tot <- prettyNum(abs(sum(outlier.fert.conn$y - outlier.fert.conn$yadj)), digits = 0, big.mark = ",")
 conn_per <- percent(abs(sum(outlier.fert.conn$y - outlier.fert.conn$yadj) / sum(outlier.fert.conn$yadj[80:180])))
@@
 Here we detect two TC outliers, nearly back-to-back in 2005 (2005:08 *t* = `r la_tstat1`; and 2005:10 *t* = `r la_tstat2`), coinciding with the destruction of Hurricane Katrina from the same year.
-The migration associated with Hurricane Katrina has garnered most of the attention of demographers [@fussellRecoveryMigrationCity2014;@horiDisplacementDynamicsSouthern2009], but the hurricane had a very clear impact on fertility behaviors too, creating a mini "baby bust" in Louisiana with the departure of many people in their childbearing years.
+The migration associated with Hurricane Katrina has garnered most of the attention of social scientists [@fussellRecoveryMigrationCity2014;@horiDisplacementDynamicsSouthern2009], but the hurricane had a very clear impact on fertility behaviors too, creating a mini "baby bust" in Louisiana with the departure of many people in their childbearing years.
 We estimate `r la_averbirths` fewer births in Louisiana compared to the counterfactual, likely attributable to the Hurricane.
@@
 Here we detect a strong (*t*= `r conn_tstat` for `r conn_births` births/month) level shift toward lower fertility beginning in August 2009 (2009:08) that continues through the end of the period (**\autoref{fig:fertla}b**).
-This LS toward lower fertility reduced the # of births by `r conn_tot` or `r conn_per`) lower than the counterfactual time series.
+This LS toward lower fertility reduced the number of births by `r conn_tot` or `r conn_per`) lower than the counterfactual time series.
 The stock market crash of 2008, where the Dow Jones Industrial Average had the largest single-day loss up to that point, occurred just 11 months before we detect a shift toward lower fertility.
 It is possible that the trend toward lower fertility in August 2009 and the stock market crash in September 2008 are linked, although, as stated previously, we are speculating only to illustrate how causal inference could work in an abductive research agenda.
+A potentially interesting future line of research would be to link fertility anomalies to the kinds of economic shocks that may lead to outmigration of those in their childbearing years.
 
@@
 Similarly, Louisiana's TC anomalies seem linked to Hurricanes Katrina and Rita while the LS anomaly in Connecticut's fertility appears linked to the Great Recession.
-However, we detect numerous seemingly unexplainable demographic anomalies in other states.
+However, we detect numerous other demographic anomalies in other states, on the causes of which we will not speculate.
 **\autoref{fig:ferthawaii}** shows two such unexplained anomalies.
@@
 hi_tstat <- prettyNum(outlier.fert.hi$outliers$tstat, digits = 3)
-hi_per <- percent(1- outlier.fert.hi$y[137]/outlier.fert.hi$yadj[137])
+hi_per <- percent(abs(1- outlier.fert.hi$y[137]/outlier.fert.hi$yadj[137]))
 

### 2020-01-28 MANUSCRIPT/manuscript.tex
Newest edits. Full paper.
I made the example figure in excel because it was faster.

@@
 uncover hidden anomalies in demographic data that have previously been
-dismissed as noise. Here, we demonstrate how demographers can use causal
-inference techniques and abductive reasoning to identify fertility and
-mortality anamolies in state-level data trends by investigating
-demographic time series data since 1999. In demonstrating our technique,
-we uncover real spikes and dips in fertility and mortality over time,
-distinguished from regular trend variations. We show that 36 states
-exhibited at least one mortality anomaly over time, totalling nearly
-300,000 additional deaths, while an additional seven states exhibited
-periods of decline resuling in 250,000 fewer deaths. Ten states
-exhibited real, marked periods of declines in fertility, totalling more
-than 214,000 fewer births than expected, and 8 states had periods of
-excess births, resulting in an addition 12,000 births. Our results show
-how researchers can detect demographic anomalies in trend data and use
-those anomalies as a beginning point for further investigation into the
-causes.
+dismissed as noise. Here, we demonstrate how social scientists can use
+causal inference techniques and abductive reasoning to identify
+fertility and mortality anamolies in state-level data trends by
+investigating demographic time series data since 1999. In demonstrating
+our technique, we uncover real spikes and dips in fertility and
+mortality over time, distinguished from regular trend variations. We
+show that 36 states exhibited at least one mortality anomaly over time,
+totalling nearly 300,000 additional deaths, while an additional seven
+states exhibited periods of decline resuling in 250,000 fewer deaths.
+Ten states exhibited real, marked periods of declines in fertility,
+totalling more than 214,000 fewer births than expected, and 8 states had
+periods of excess births, resulting in an addition 12,000 births. Our
+results show how researchers can detect demographic anomalies in trend
+data and use those anomalies as a beginning point for further
+investigation into the causes.
 \end{abstract}
@@
 Causal inference, in its most common usage, is an attempt to uncover the
-underlying mechanism that results in changes in a phenomenon and have a
-long history approaches in social sciences \citep{grimmer2015ppsp}.
-Often, the identification of a casual mechanism requires either a
-randomized control trial (RCT) or expert knowledge applied to a natural
-experiment. For population-level analysis, expensive RCTs usually
-preclude their widespread adoption \citep{west2008ajph}. Using natural
-experiments provide less certainty about causation and fewer
-opportunities to replicate work but they provide important and
-widespread insights such as the determinants of migration, however,
-provides important and more widespread insights on the determinants of
-migration after Hurricane Katrina in 2006
-\citep{fussellRecoveryMigrationCity2014, horiDisplacementDynamicsSouthern2009},
-mini baby booms after electrical blackouts \citep{fetzer2018jpe}, and
-the highly publicized estimates of excess mortality after Hurricane
-Maria in Puerto Rico in 2016
-\citep{kishore2018mortality, santos2018use}. Both RCTs and natural
-experiments follow traditional, hypothesis testing, inductive scientific
-paradigms where a question is first posed and scientists generate or
-find data to answer the question. Such approaches, while extensively
-utilized and time-tested, are likely to miss important phenomena that
-might go unnoticed.
+underlying mechanisms that result in changes in a phenomenon and has a
+long history as an approach to inquiry in the social sciences
+\citep{grimmer2015ppsp}. Often, the identification of a casual mechanism
+requires either a randomized control trial (RCT) or expert knowledge
+applied to a natural experiment. For population-level analysis,
+expensive RCTs usually preclude their widespread adoption
+\citep{west2008ajph}. Using natural experiments provide less certainty
+about causation and fewer opportunities to replicate work but they
+provide important and widespread insights.
 
@@
 In this paper, we use modern statistical outlier detection algorithms
-\citep{chen1993joint} on nearly twenty years of mortality and fertility
-data at the US state-level to identify anomalous demographic behavior.
-In essence, we identify \emph{effects without knowledge of the cause}.
-We ask two questions regarding demographic anomalies: What are the
-hidden baby booms/busts and mortality spikes/dips in the United States
-over the last twenty years? We do not necessarily know the causes of
-these anomalies but identifying them allows scientists with more
-detailed knowledge of local population dynamics, state-level policy
-making, or macro-economics to explain these phenomena post-hoc. From
-explanations of phenomena that may have previously gone unnoticed,
-demographers may be able to better forecast populations and provide
-policy solutions for impending problems.
+\citep{chen1993joint} as a means to illustrate one way that social
+scientists can engage in discovery using big data techniques on nearly
+twenty years of mortality and fertility data at the US state-level to
+identify anomalous demographic behavior. In essence, we identify
+\emph{effects without knowledge of the cause}. We ask two questions
+regarding demographic anomalies: What are the hidden baby booms/busts
+and mortality spikes/dips in the United States over the last twenty
+years? We do not necessarily know the causes of these anomalies but
+identifying them allows scientists with more detailed knowledge of local
+population dynamics, state-level policy making, or macro-economics to
+explain these phenomena post-hoc. From explanations of phenomena that
+may have previously gone unnoticed, social scientists may be able to
+better forecast populations and provide policy solutions for impending
+problems such as climate change.
 
@@
 
+\begin{figure}
+\centering
+\includegraphics{./manuscript_files/figure-latex/Figure1.png}
+\caption{\textbf{``Toy'' example of anomaly detection using the exchange
+rate between the Euro and the Swiss Franc.} The Swiss Franc was pegged
+to the Euro at 1.2 until January 15 2015 when the peg was removed and
+the currency was allowed to trade more freely. This is a very strong
+anomly in the time series (\emph{t}-stat = -144.07) detectable using
+statistical methods and visually. The countefactural estimate with the
+outlier removed would simply have kept the currency exchange at 1.2
+(\(\hat{y}\)). \label{explanationfigure}}
+\end{figure}
+
 where \(\pi(B)=\sum_{i=o}^{inf} \pi_iB^i\). The identification of
-outliers then involves a three step process to (1) identify all
-potential outliers, \(t_j\) and \(L_j (B)\), (2) joint estimates of
-model parameters and outlier effects are computed to identify
-potentially spurious outliers, and (3) the outliers and effects are
-re-estimated without spurious outliers. Of importance here is step 2,
-which relies on some critical value above which an outlier at time point
-m is considered spurious. Based on Chen and Liu's
+outliers then involves a three step process. First, (1) the algorithm
+identifies all potential outliers, \(t_j\) and \(L_j (B)\). Next, (2) we
+compute joint estimates of model parameters and outlier effects to
+identify potentially spurious outliers. Finally, (3) the outliers and
+effects are re-estimated without spurious outliers. Of importance here
+is step 2, which relies on some critical value above which an outlier at
+time point \(m\) is considered spurious. Based on Chen and Liu's
 \citeyearpar{chen1993joint} recommendation, we set a critical value of
@@
 original, unadjusted time series.
+
+\textbf{\autoref{explanationfigure}} shows a toy example for anomaly
+detection in a time series using the classic example of the exchange
+rate between the Euro and the Swiss Franc (CHF). On January 15 2015, the
+Swiss National Bank removed the currency peg of 1.20 francs per Euro,
+exposing the Franc to the volatility of the currency market. The Franc
+immediately began trading a reduced rate compared to the Euro. What
+would have been the Euro/CHF exchange rate had the Swiss National Bank
+\emph{not} removed the peg? A simple counter-factual estimate would be
+to keep the exchange rate at 1.20 (\(\hat{y}\), sometimes called a
+``synthetic control'' \citep{abadie2010synthetic}).
+
+In the Swiss Franc example, we have knowledge of the Swiss Bank's
+activities after the fact or ex-post-facto to to create the
+counter-factual time series \(\hat{y}\) of 1.20. But is this anomaly
+detectable without knowledge of the Swiss Bank's activities? In other
+words, can we detect the reduction of the exchange rate on January 15
+using only the time series? Absolutely. the tsoutliers package
+identifies January 15 2015 as an extremely strong level-shift outlier
+(\emph{t}-stat = -144.07). The real world is hardly this simplified
+where a direct intervention is known and is testable.
 
@@
 
-\caption{\label{tab:unnamed-chunk-2}\textbf{Summary by Demographic Component.}  refers to the number of anomolies for each type and demographic component, States refers to the number of unique states exhibiting that specific Type-demographic component.
+\caption{\label{tab:unnamed-chunk-3}\textbf{Summary by Demographic Component.}  Here we can see there 22 Fertility and 156 Mortality anomalies among the state-level time series, totalling more than 200k anomalous births and 600k anomalous deaths.
 \label{sumtable}}
@@
 
-\textbf{\autoref{sumtable}} report the overall number of anomalies we
+\textbf{\autoref{sumtable}} reports the overall number of anomalies we
 detect of each type for births and deaths and some summary statistics
@@
 events that impact fertility and mortality are catastrophic. Researchers
-have documented that more commonplace events such as massive layoff
+have documented that more commonplace events such as massive layoffs
 \citep{Venkataramani2019} or policy changes
@@
 positive and negative anomalies, but tend toward more deaths rather than
-less deaths. Again, this is not surprising, as the kinds of events that
+fewer deaths. Again, this is not surprising, as the kinds of events that
 increase death (e.g., plant closings are associated with increased
@@
 
-Since the purpose of our paper is to demonstrate how demographers can
-make use of causal inference technique, we will not discuss all of our
-findings here, in this paper. Instead, we will use a few cases for
+Since the purpose of our paper is to demonstrate how social scientists
+can make use of causal inference technique, we will not discuss all of
+our findings in this paper. Instead, we will use a few cases for
 illustration. For mortality, we will use New York State as an example
@@
 month than anticipated. This mortality event is likely caused by the
-September 11 tragedy and the detection of this mortality event provides
+September 11 terrorist attack on the World Trade Center that immediately
+killed 2,606 people and the detection of this mortality event provides
 confidence in our detection of other anomalies.
@@
 place that might have contributed to this considerable mortality
-reduction? What environmental conditions may have changed? These are the
-kinds of questions that arise from our analyses. The purpose of our
-paper is not to answer these questions, but our findings underscore the
-need for more research that takes an abductive approach. By identifying
-anomalies through an inductive process, researchers can then look for
-underlying causes. Once those causes are identified, researchers can
-then use a deductive process to see if such events predict other (or
-future) anomalies.
+reduction? What environmental or economic conditions may have changed
+may have changed? These are the kinds of questions that arise from our
+analyses. The purpose of our paper is not to answer these questions, but
+our findings underscore the need for more research that takes an
+abductive approach. By identifying anomalies through an inductive
+process, researchers can then look for underlying causes. Once those
+causes are identified, researchers can then use a deductive process to
+see if such events predict other (or future) anomalies.
 
@@
 same year. The migration associated with Hurricane Katrina has garnered
-most of the attention of demographers
+most of the attention of social scientists
 \citep{fussellRecoveryMigrationCity2014, horiDisplacementDynamicsSouthern2009},
@@
 (\textbf{\autoref{fig:fertla}b}). This LS toward lower fertility reduced
-the \# of births by -28,752 or 8.57\%) lower than the counterfactual
+the number of births by 28,752 or 8.57\%) lower than the counterfactual
 time series. The stock market crash of 2008, where the Dow Jones
@@
 previously, we are speculating only to illustrate how causal inference
-could work in an abductive research agenda.
+could work in an abductive research agenda. A potentially interesting
+future line of research would be to link fertility anomalies to the
+kinds of economic shocks that may lead to outmigration of those in their
+childbearing years.
 
@@
 Connecticut's fertility appears linked to the Great Recession. However,
-we detect numerous seemingly unexplainable demographic anomalies in
-other states. \textbf{\autoref{fig:ferthawaii}} shows two such
-unexplained anomalies.
+we detect numerous other demographic anomalies in other states, on the
+causes of which we will not speculate. \textbf{\autoref{fig:ferthawaii}}
+shows two such unexplained anomalies.
 
@@
 anomaly in Hawaiian fertility in May 2014. This is a strong anomaly with
-a \emph{t}-statistic of 4.96, -14.9\% above the counter-factual time
+a \emph{t}-statistic of 4.96, 14.9\% above the counter-factual time
 series. This single, anomalous month is also the second highest monthly

### 2020-01-29 MANUSCRIPT/manuscript.Rmd
After Stephanie's suggested edits. Some spell checking too.

@@
 abstract: "The proliferation of big data, wider access to advanced computing platforms, and the development of powerful statistical algorithms can uncover hidden anomalies in demographic data that have previously been dismissed as noise.
-Here, we demonstrate how social scientists can use causal inference techniques and abductive reasoning to identify fertility and mortality anamolies in state-level data trends by investigating demographic time series data since 1999.
+Here, we demonstrate how social scientists can use causal inference techniques and abductive reasoning to identify fertility and mortality anomalies in state-level data trends by investigating demographic time series data since 1999.
 In demonstrating our technique, we uncover real spikes and dips in fertility and mortality over time, distinguished from regular trend variations.
-We show that 36 states exhibited at least one mortality anomaly over time, totalling nearly 300,000 additional deaths, while an additional seven states exhibited periods of decline resuling in 250,000 fewer deaths.
-Ten states exhibited real, marked periods of declines in fertility, totalling more than 214,000 fewer births than expected, and 8 states had periods of excess births, resulting in an addition 12,000 births.
+We show that 36 states exhibited at least one mortality anomaly over time, totaling nearly 300,000 additional deaths, while an additional seven states exhibited periods of decline resulting in 250,000 fewer deaths.
+Ten states exhibited real, marked periods of declines in fertility, totaling more than 214,000 fewer births than expected, and 8 states had periods of excess births, resulting in an addition 12,000 births.
 Our results show how researchers can detect demographic anomalies in trend data and use those anomalies as a beginning point for further investigation into the causes."

### 2020-01-29 MANUSCRIPT/manuscript.tex
After Stephanie's suggested edits. Some spell checking too.

@@
 causal inference techniques and abductive reasoning to identify
-fertility and mortality anamolies in state-level data trends by
+fertility and mortality anomalies in state-level data trends by
 investigating demographic time series data since 1999. In demonstrating
@@
 show that 36 states exhibited at least one mortality anomaly over time,
-totalling nearly 300,000 additional deaths, while an additional seven
-states exhibited periods of decline resuling in 250,000 fewer deaths.
+totaling nearly 300,000 additional deaths, while an additional seven
+states exhibited periods of decline resulting in 250,000 fewer deaths.
 Ten states exhibited real, marked periods of declines in fertility,
-totalling more than 214,000 fewer births than expected, and 8 states had
+totaling more than 214,000 fewer births than expected, and 8 states had
 periods of excess births, resulting in an addition 12,000 births. Our

### 2020-02-24 MANUSCRIPT/manuscript.Rmd
Updated. V1.0 of the manuscript.

@@
 
-abstract: "The proliferation of big data, wider access to advanced computing platforms, and the development of powerful statistical algorithms can uncover hidden anomalies in demographic data that have previously been dismissed as noise.
+# abstract: "The proliferation of big data, wider access to advanced computing platforms, and the development of powerful statistical algorithms can uncover hidden anomalies in demographic data that have previously been dismissed as noise.
 Here, we demonstrate how social scientists can use causal inference techniques and abductive reasoning to identify fertility and mortality anomalies in state-level data trends by investigating demographic time series data since 1999.
@@
 Our results show how researchers can detect demographic anomalies in trend data and use those anomalies as a beginning point for further investigation into the causes."
+
+abstract: "The proliferation of big data, wider access to advanced computing platforms, and the development of powerful statistical algorithms can uncover hidden anomalies in social data, previously dismissed as noise.
+Here, we combine causal inference techniques and abductive reasoning to identify fertility and mortality anomalies on twenty years of demographic data in the United States.
+We uncover real, “hidden” baby booms/busts and mortality spikes/dips, distinguishable from regular trend variations.
+We identify more than 22 and 156 fertility and mortality anomalies, totaling more than 200k and 600k anomalous births and deaths, respectively.
+Notable detectable mortality anomalies include the September 11 2001 terrorist attack in New York, the emergence and acceleration of the opioid epidemic in New Hampshire, and the excess mortality due to Hurricane Maria in Puerto Rico.
+Notable fertility anomalies include the “missing births” in Louisiana after Hurricane Katrina and the reduction in fertility behavior after the September 2008 stock market crash in Connecticut, amongst others.
+Our combined causal inference and abductive reasoning approach can be readily adapted to find other, undiscovered social phenomena or to evaluate the efficacy of important public policies."
 
@@
 
-The proliferation of data, advances in high performance (“super”) computing, and the development of powerful statistical algorithms mark the era of "Big Data" or data science [@van2016data;@zikopoulos2011], with the potential for researchers to find hidden or understudied social phenomena [@bohon2018demography].
-While the availability of Big Data and high performance computing allows novel exploration of data through causal inference [@bohon2018demography;@rcausalimpact;@shiffrin2016drawing], relatively few studies utilize causal inference techniques in the study of demographic phenomena.
-However, understanding social phenomena using these advances reveals important insights into society [@angrist1989lifetime;@mas2009peers] and allows us to better monitor population trends [@nobles2019;@torche2015hidden].
+The search for causality is one of the primary motiviations behind social science [@smith2011commentary; @pearl2009causal].
+Causal Inference, the identification of causal mechanisms, attempts to uncover the underlying mechanisms that result in changes in a phenomenon and has a long history as an approach to inquiry in the social sciences [@grimmer2015ppsp].
+Often, the identification of a casual mechanism is accomplished via a randomized control trial (RCT) or expert knowledge applied to a natural experiment [@salganik2019bit].
 
-Causal inference, in its most common usage, is an attempt to uncover the underlying mechanisms that result in changes in a phenomenon and has a long history as an approach to inquiry in the social sciences [@grimmer2015ppsp].
-Often, the identification of a casual mechanism requires either a randomized control trial (RCT) or expert knowledge applied to a natural experiment.
-For population-level analysis, expensive RCTs usually preclude their widespread adoption [@west2008ajph].
-Using natural experiments provide less certainty about causation and fewer opportunities to replicate work but they provide important and widespread insights.
+Unfortunately, both RCTs and natural experiments have drawbacks.
+RCTs are often prehibitively expensive for many scientists, precluding their widespread adoption in the social sciences [@west2008ajph].
+RCTs are also sometimes impossible to deploy in the "real world".
+For instance, an RCT to study country-level austeriy measures would be difficult, if not impossible, to implement.
+Natural experiments are a cheaper alternative, but provide less certainty about causation, fewer opportunities to replicate work, and require expert knowledge of the natural experiment to be properly employed [@pearl2018book].
+
+<!-- In its most basic application, causal inference approaches accomplish this with two populations, a control group ($Y(0)$) and a treatment group ($Y(1)$) and simply subtracting the results of the control from the treatment ($effect = Y(1) - Y(0))$).
+-->
 
@@
 
-Both RCTs and natural experiments follow traditional hypothesis testing--inductive scientific paradigms where a question is first posed and scientists generate or find data to answer the question.
-Such approaches, while extensively utilized and time-tested, are likely to miss important phenomena that might go unnoticed.
-Data scientists are rethinking these approaches and have re-envisioned causal inference as a machine learning technique that allows big data researchers to move from correlations that have been established with good accuracy to making strong conclusions about the underlying mechanisms that produce these correlations
-[@pearl2018book].
+Additionally, both RCTs and natural experiments follow traditional hypothesis testing, the deductive scientific paradigm where a question is first posed and scientists generate or find data to answer the question.
+Such approaches, while extensively utilized and time-tested, are likely to miss important phenomena that might go unnoticed or suffer from extensive p-hacking [@head2015extent;@nuzzo2014scientific;@ruggles2014big].
+Abductive reasoning, sometimes described as “inferring cause from effect” [@Crowder2017], moves the scientific process from the inductive to the deductive, and sometimes back and forth, to reach conclusions [@bryant2014realm] The physical sciences use abductive reasoning more fully compared to the social sciences.
+The discovery of Neptune in 1846 based on unexplained perturbations of the orbit of Uranus [@popper2005logic], the discovery of galactic migration of planets [@gomes2005n], and the current generation of gravitational wave detectors such as the Laser Interferometer Gravitational-Wave Observatory (LIGO) [@harry2010advanced] demonstrate the power of the abductive approach in the physical sciences.
 
-Abductive modeling is the movement from the inductive to the deductive, and sometimes back and forth, to reach conclusions [@bryant2014realm] or “inferring cause from effect” [@Crowder2017].
+In a data-rich environment, abductive reasoning can be more easily applied to first find effects and then infer their cause ex-post.
+The modern proliferation of data, advances in high performance (“super”) computing, and the development of powerful statistical algorithms provide an environment ripe for wedding causal inference with abductive reasoning [@van2016data;@zikopoulos2011], with the potential for researchers to find hidden or understudied social phenomena [@bohon2018demography].
+While the availability of Big Data and high performance computing allows novel exploration of data through causal inference [@bohon2018demography;@rcausalimpact;@shiffrin2016drawing], relatively few studies utilize causal inference techniques in the study of demographic phenomena.
+However, understanding social phenomena using these advances reveals important insights into society [@angrist1989lifetime;@mas2009peers] and allows us to better monitor population trends [@nobles2019;@torche2015hidden].
+The rich demographic data available in the United States make the potential revelation of interesting and important demographic phenomena not only possible, but extremely plausible – even if identification of the phenomena occurs without identifying the underlying cause.
+
+<!-- Data scientists are rethinking these approaches and have re-envisioned causal inference as a machine learning technique that allows big data researchers to move from correlations that have been established with good accuracy to making strong conclusions about the underlying mechanisms that produce these correlations [@pearl2018book].
+The discovery of Neptune in 1846 In other sciences, the identification of effects without causes has led to new theories of galactic migration of planets and other breakthroughs [@gomes2005n].
+-->
+
+<!-- Abductive modeling is the movement from the inductive to the deductive, and sometimes back and forth, to reach conclusions [@bryant2014realm] or “inferring cause from effect” [@Crowder2017].
 In situations with large data sets, an abductive approach is far superior to deductive hypothesis testing, as p-values with an extremely large number of cases are far from revealing [@head2015extent;@nuzzo2014scientific] and we do not want to make purely inductive inferences from data of questionable generalizability [@ruggles2014big].
-Owing to the long history of big data sets in demographic research (one of the original "big data" resources [@ruggles2014big]), the rich demographic data available in the United States make the potential revelation of interesting and important demographic phenomena not only possible, but extremely plausible – even if identification of the phenomena occurs without identifying the underlying cause.
-In other sciences, the identification of effects without causes has led to new theories of galactic migration of planets and other breakthroughs [@gomes2005n].
+Owing to the long history of big data sets in demographic research (one of the original "big data" resources [@ruggles2014big]),  -->
 
-In this paper, we use modern statistical outlier detection algorithms [@chen1993joint] as a means to illustrate one way that social scientists can engage in discovery using big data techniques on nearly twenty years of mortality and fertility data at the US state-level to identify anomalous demographic behavior.
+In this paper, we combine a modern statistical outlier detection algorithm [@chen1993joint], abductive reasoning, and large demographic data sets on more than 20 years of mortality/fertility to identify anomalous demographic behavior.
 In essence, we identify *effects without knowledge of the cause*.
-We ask two questions regarding demographic anomalies: What are the hidden baby booms/busts and mortality spikes/dips in the United States over the last twenty years?
+We ask one fundamanetal question regarding demographic anomalies: What are the hidden baby booms/busts and mortality spikes/dips in the United States over the last twenty years?
 We do not necessarily know the causes of these anomalies but identifying them allows scientists with more detailed knowledge of local population dynamics, state-level policy making, or macro-economics to explain these phenomena post-hoc.
@@
 A full listing of these anomalies can be found in the **Supplementary Materials**.
-We begin with a summary of the anomalies we detect and then we highlight highly significant mortality and fertility anomalies across all three types of outliers (Additive Outliers, Level Shift Outliers, and Temporary Change Outliers) with plausible explanations.
-Finally, we highlight two strong anomalies that bely explanation.
+We begin by highlighting significant fertility and mortality anomaloies across all three types of outliers (Additive Outliers, Level Shift Outliers, and Temporary Change outliers) with plausible explanations.
+We then highlight two strong anomalies that bely explanation.
+Finally, we conclude with a summary of the anomalies we detect.
+
+## Fertility
+
+```{r, include = FALSE}
+la_tstat1 <- prettyNum(outlier.fert.la$outliers$tstat[1], digits = 4) 
+la_tstat2 <- prettyNum(outlier.fert.la$outliers$tstat[2], digits = 4)
+la_averbirths <- prettyNum(abs(sum(outlier.fert.la$y - outlier.fert.la$yadj)), digits = 0, big.mark = ",")
+
+conn_tstat <- prettyNum(outlier.fert.conn$outliers$tstat, digits = 3)
+conn_births <- prettyNum(outlier.fert.conn$outliers$coefhat, digits = 0)
+conn_tot <- prettyNum(abs(sum(outlier.fert.conn$y - outlier.fert.conn$yadj)), digits = 0, big.mark = ",")
+conn_per <- percent(abs(sum(outlier.fert.conn$y - outlier.fert.conn$yadj) / sum(outlier.fert.conn$yadj[80:180])))
+```
+
+We begin with a clear temporary change (TC) outlier for fertility in Louisiana (**\autoref{fig:fertla}a**).
+Here we detect two TC outliers, nearly back-to-back in 2005 (2005:08 *t* = `r la_tstat1`; and 2005:10 *t* = `r la_tstat2`), coinciding with the destruction of Hurricane Katrina from the same year.
+The migration associated with Hurricane Katrina has garnered most of the attention of social scientists [@fussellRecoveryMigrationCity2014;@horiDisplacementDynamicsSouthern2009], but the hurricane had a very clear impact on fertility behaviors too, creating a mini "baby bust" in Louisiana with the departure of many people in their childbearing years.
+We estimate `r la_averbirths` fewer births in Louisiana compared to the counterfactual, likely attributable to the Hurricane.
+
+We also highlight a level shift (LS) outlier for fertility in Connecticut.
+Here we detect a strong (*t*= `r conn_tstat` for `r conn_births` births/month) level shift toward lower fertility beginning in August 2009 (2009:08) that continues through the end of the period (**\autoref{fig:fertla}b**).
+This LS toward lower fertility reduced the number of births by `r conn_tot` or `r conn_per`) lower than the counterfactual time series.
+The stock market crash of 2008, where the Dow Jones Industrial Average had the largest single-day loss up to that point, occurred just 11 months before we detect a shift toward lower fertility.
+It is possible that the trend toward lower fertility in August 2009 and the stock market crash in September 2008 are linked, although, as stated previously, we are speculating only to illustrate how causal inference could work in an abductive research agenda.
+A potentially interesting future line of research would be to link fertility anomalies to the kinds of economic shocks that may lead to outmigration of those in their childbearing years.
+
+```{r FertilityAnomalies, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Louisiana (a) and Connecticut (b) state fertility, 2003-2018.} The top part of each panel contains the original time series (light gray), the corrected, counter-factual time series in the absence of anomalies (blue), and the red dots correspond to the onset of detected anomalies.
+The bottom part of each panel contains the magnitude and type of the outlier in red.
+In (a), we detect two outliers, back-to-back, likely resulting from Hurricane Katrina in August and October 2005, representing a decrease of more than 4,400 births due to the hurricane (t-statistics= -6.002 and -4.997).
+In (b), we detect one outlier, a level shift outlier (LS) in August 2009.
+We believe this reduction is attributable to the stock market crash 11 months earlier (t-statistic = -3.71).
+\\label{fig:fertla}")}
+
+a <-  function() {
+  plot(outlier.fert.la, cex.axis=0.8)
+  title(main = "Louisiana - Fertility", adj = 0.9)
+}
+
+b <- function(){
+  plot(outlier.fert.conn, cex.axis=0.8) +
+  title(main = "Connecticut - Fertility", adj = 0.88)
+}
+
+plot_grid(ggdraw(a) ,
+               ggdraw(b), 
+               ncol=1,
+               labels= "auto")
+```
+
+## Mortality 
+For mortality, we will use New York State as an example (**\autoref{fig:mortnewhamp}a**).
+We identify seven anomalies in the mortality time series for New York, all with *t*-statistics in excess of 3.91, making these anomalies unlikely to be due to chance.
+The algorithm correctly identifies September 2001 as an additive outlier (2001:09 *t*= `r paste(prettyNum(outlier.mort.ny$outliers[2,5], digits = 4))`) where there were `r paste(prettyNum(outlier.mort.ny$outliers$coefhat[2], digits = 4, big.mark = ","))` more deaths in that month than anticipated.
+This mortality event is likely caused by the September 11 terrorist attack on the World Trade Center that immediately killed 2,606 people and the detection of this mortality event provides confidence in our detection of other anomalies.
+
+```{r, include = FALSE}
+
+ny_tstat <- prettyNum(outlier.mort.ny$outliers[3,5], digits = 4)
+ny_averteddeaths <- prettyNum(abs(outlier.mort.ny$outliers[3,4]), digits = 1)
+ny_averteddeathstot <- prettyNum(round(abs(outlier.mort.ny$outliers[3,4]*(228-outlier.mort.ny$outliers[3,2])),digits =-3), big.mark=",")
+ny_peraverted <- percent(abs(outlier.mort.ny$outliers[3,4]*(228-outlier.mort.ny$outliers[3,2]))/sum(outlier.mort.ny$y[62:228]))
+
+nh_tstat1 <- prettyNum(outlier.mort.nh$outliers[1,5], digits = 3)
+nh_tstat2 <- prettyNum(outlier.mort.nh$outliers[2,5], digits = 3)
+nh_totals <- prettyNum(sum(outlier.mort.nh$y - outlier.mort.nh$yadj), digits=1, big.mark=",")
+nh_per <- percent(1- sum(outlier.mort.nh$yadj[136:228])/ sum(outlier.mort.nh$y[136:228]))
+
+```
+
+In **\autoref{fig:mortnewhamp}a**, notice the strong level shift (LS) that occurs in February 2004 (2004:02 *t*= `r paste(ny_tstat)`) which prevented `r paste(ny_averteddeaths)` deaths per month.
+This shift totals more than `r paste(ny_averteddeathstot)` averted deaths compared to the counter-factual time series and is the single largest mortality protective anomaly among all states.
+This translates to `r ny_peraverted` fewer deaths than expected over the time period.
+What is driving this mortality protection?
+What policies did NY put into place that might have contributed to this considerable mortality reduction?
+What environmental or economic conditions may have changed may have changed?
+These are the kinds of questions that arise from our analyses.
+The purpose of our paper is not to answer these questions, but our findings underscore the need for more research that takes an abductive approach.
+By identifying anomalies through an inductive process, researchers can then look for underlying causes.
+Once those causes are identified, researchers can then use a deductive process to see if such events predict other (or future) anomalies.
+
+To see the potential for combining abductive reasoning with causal inference, contrast the mortality protection in New York with the enhanced mortality in New Hampshire (**\autoref{fig:mortnewhamp}b**).
+In New Hampshire we detect two significant level shifts (LS) in the monthly mortality data, first in April 2010 and again in November 2014 (2010:04 *t*= `r nh_tstat1`; 2014:11 *t*= `r nh_tstat2`).
+These anomalies suggest New Hampshire experienced `r nh_totals` more deaths (+`r nh_per` more than expected) in a seven-year period beginning in early 2010.
+These are events not experienced by neighboring states during the same time period, and this is the single largest percentage mortality increase/decrease we detected among all states.
+Not coincidentally, NH has the second highest opioid-related mortality in the US [@beetham2019access] and it is likely that we detect this epidemic in our results.
+Isolating these anomalies and testing them against opioid sales data might yield intriguing results.
+
+```{r MortalityAnomalies, echo= FALSE, message = FALSE, warning = FALSE,dev=c('pdf','png'), fig.height = 5, fig.cap= paste("\\textbf{Anomaly Detection for New York (a) and New Hampshire (b) state mortality, 1999-2016.} In New York (a), we detect additive outliers (AO) in January 2000, September 2001, January 2005, and January 2013; temporary change (TC) outliers in February 2007 and January 2015; and a level shift (LS) starting in February 2004.
+In New Hampshire (b), we detect two outliers, both level shift outliers (LS) in April 2010 and again in November 2014.
+These anomalies suggest New York experienced a significant mortality event in September 2001 and New Hampshire experienced approximately 9,700 more deaths than expected since 2010 or 14% more deaths in the state over just seven years.
+\\label{fig:mortnewhamp}")}
+
+a <-  function() {
+plot(outlier.mort.ny, cex = 0.8)
+title("New York - Mortality", adj = 0.9, line =2)
+}
+
+b <-  function() {
+  plot(outlier.mort.nh, cex = 0.8)
+  title("New Hampshire - Mortality", adj = 0.82,line =2)
+}
+
+plot_grid(ggdraw(a) ,
+               ggdraw(b), 
+               ncol=1,
+               labels= "auto")
+```
+
+## Interesting Anomalous Fertility/Mortality events
+In the examples above, we highlighted four fertility/mortality anomalies with plausible explanations.
+In the case of New York and New Hampshire, the mortality anomalies have plausible explanations.
+It seems likely that New York’s AO anomaly in September 2001 is caused by the 9/11 tragedy and the rise in New Hampshire’s mortality starting in 2010 could be linked to the opioid epidemic.
+Similarly, Louisiana's TC anomalies seem linked to Hurricanes Katrina and Rita while the LS anomaly in Connecticut's fertility appears linked to the Great Recession.
+However, we detect numerous other demographic anomalies in other states, on the causes of which we will not speculate.
+**\autoref{fig:ferthawaii}** shows two such unexplained anomalies.
+
+```{r, include = FALSE}
+hi_tstat <- prettyNum(outlier.fert.hi$outliers$tstat, digits = 3)
+hi_per <- percent(abs(1- outlier.fert.hi$y[137]/outlier.fert.hi$yadj[137]))
+
+oh_tstat <- prettyNum(outlier.mort.oh$outliers$tstat[7], digits = 3)
+oh_deaths <- prettyNum(round(abs(outlier.mort.oh$outliers[7,4]), digits =0), big.mark=",")
+oh_tot <- prettyNum(abs(round(outlier.mort.oh$outliers[7,4]*(228-194), digits = -4)), big.mark=",")
+```
+
+In **\autoref{fig:ferthawaii}a** we identify a single additive anomaly in Hawaiian fertility in May 2014.
+This is a strong anomaly with a *t*-statistic of `r hi_tstat`, `r hi_per` above the counter-factual time series.
+This single, anomalous month is also the second highest monthly births in the time series.
+We have no plausible explanation for this anomaly.
+We do not believe this is simply a data error as the other extreme values, September 2008 and February 2005 with the highest and lowest recorded fertility respectively, were not identified as anomalous events.
+Even if we were to assume this anomaly resulted from data entry error, it remains an *unaltered* data error in the Hawaiian monthly fertility data.
+
+This is contrast to **\autoref{fig:ferthawaii}b**, where we identify a strange mortality reduction in Ohio (*t*-stat: `r oh_tstat`).
+This LS is more than `r oh_deaths` deaths per month less than the counterfactual time series, suggesting had nearly `r oh_tot` fewer deaths since February 2015 than expected.
+This is the single-largest LS among all states.
+We could not identify the potential policies Ohio might have put into place to provide such a strong mortality protection.
+
+```{r TrueAnomalies, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Hawaii state fertility (a) and Ohio state mortality (b).} Here we detect one outlier, an additive outlier (AO) in May 2014 in Hawaii (a), representing a large, unexplainable 14% increase in expected births in that month.
+We also detect a strong level shift (LS) reduction (b), representing a large, unexplainable reduction of nearly 40,000 deaths in Ohio.
+\\label{fig:ferthawaii}")}
+
+a <-  function() {
+plot(outlier.fert.hi, cex.axis=0.8)
+  title(main = "Hawaii - Fertility", adj = 0.88)
+}
+
+b <-  function() {
+plot(outlier.mort.oh, cex.axis=0.8)
+  title(main = "Ohio - Mortality", adj = 0.9)
+}
+
+plot_grid(ggdraw(a) ,
+               ggdraw(b), 
+               ncol=1,
+               labels= "auto")
+```
 
@@
 **\autoref{sumtable}** reports the overall number of anomalies we detect of each type for births and deaths and some summary statistics across all anomaly types and **\autoref{fig:summap}** maps these results.
+
 We find considerably more mortality anomalies (n = `r paste(sums[2,2])`) than fertility anomalies (n = `r paste(sums[1,2])`).
@@
 
-## Mortality 
-Since the purpose of our paper is to demonstrate how social scientists can make use of causal inference technique, we will not discuss all of our findings in this paper.
-Instead, we will use a few cases for illustration.
-For mortality, we will use New York State as an example (**\autoref{fig:mortnewhamp}a**).
-We identify seven anomalies in the mortality time series for New York, all with *t*-statistics in excess of 3.91, making these anomalies unlikely to be due to chance.
-The algorithm correctly identifies September 2001 as an additive outlier (2001:09 *t*= `r paste(prettyNum(outlier.mort.ny$outliers[2,5], digits = 4))`) where there were `r paste(prettyNum(outlier.mort.ny$outliers$coefhat[2], digits = 4, big.mark = ","))` more deaths in that month than anticipated.
-This mortality event is likely caused by the September 11 terrorist attack on the World Trade Center that immediately killed 2,606 people and the detection of this mortality event provides confidence in our detection of other anomalies.
-
-```{r, include = FALSE}
-
-ny_tstat <- prettyNum(outlier.mort.ny$outliers[3,5], digits = 4)
-ny_averteddeaths <- prettyNum(abs(outlier.mort.ny$outliers[3,4]), digits = 1)
-ny_averteddeathstot <- prettyNum(round(abs(outlier.mort.ny$outliers[3,4]*(228-outlier.mort.ny$outliers[3,2])),digits =-3), big.mark=",")
-ny_peraverted <- percent(abs(outlier.mort.ny$outliers[3,4]*(228-outlier.mort.ny$outliers[3,2]))/sum(outlier.mort.ny$y[62:228]))
-
-nh_tstat1 <- prettyNum(outlier.mort.nh$outliers[1,5], digits = 3)
-nh_tstat2 <- prettyNum(outlier.mort.nh$outliers[2,5], digits = 3)
-nh_totals <- prettyNum(sum(outlier.mort.nh$y - outlier.mort.nh$yadj), digits=1, big.mark=",")
-nh_per <- percent(1- sum(outlier.mort.nh$yadj[136:228])/ sum(outlier.mort.nh$y[136:228]))
-
-```
-
-In **\autoref{fig:mortnewhamp}a**, notice the strong level shift (LS) that occurs in February 2004 (2004:02 *t*= `r paste(ny_tstat)`) which prevented `r paste(ny_averteddeaths)` deaths per month.
-This shift totals more than `r paste(ny_averteddeathstot)` averted deaths compared to the counter-factual time series and is the single largest mortality protective anomaly among all states.
-This translates to `r ny_peraverted` fewer deaths than expected over the time period.
-What is driving this mortality protection?
-What policies did NY put into place that might have contributed to this considerable mortality reduction?
-What environmental or economic conditions may have changed may have changed?
-These are the kinds of questions that arise from our analyses.
-The purpose of our paper is not to answer these questions, but our findings underscore the need for more research that takes an abductive approach.
-By identifying anomalies through an inductive process, researchers can then look for underlying causes.
-Once those causes are identified, researchers can then use a deductive process to see if such events predict other (or future) anomalies.
-
-To see the potential for combining abductive reasoning with causal inference, contrast the mortality protection in New York with the enhanced mortality in New Hampshire (**\autoref{fig:mortnewhamp}b**).
-In New Hampshire we detect two significant level shifts (LS) in the monthly mortality data, first in April 2010 and again in November 2014 (2010:04 *t*= `r nh_tstat1`; 2014:11 *t*= `r nh_tstat2`).
-These anomalies suggest New Hampshire experienced `r nh_totals` more deaths (+`r nh_per` more than expected) in a seven-year period beginning in early 2010.
-These are events not experienced by neighboring states during the same time period, and this is the single largest percentage mortality increase/decrease we detected among all states.
-Not coincidentally, NH has the second highest opioid-related mortality in the US [@beetham2019access] and it is likely that we detect this epidemic in our results.
-Isolating these anomalies and testing them against opioid sales data might yield intriguing results.
-
-```{r MortalityAnomalies, echo= FALSE, message = FALSE, warning = FALSE,dev=c('pdf','png'), fig.height = 5, fig.cap= paste("\\textbf{Anomaly Detection for New York (a) and New Hampshire (b) state mortality, 1999-2016.} The top part of each panel contains the original time series (light gray), the corrected, counter-factual time series in the absence of anomalies (blue), and the red dots correspond to the onset of detected anomalies.
-The bottom part of each panel contains the magnitude and type of the outlier in red.
-In New York (a), we detect additive outliers (AO) in January 2000, September 2001, January 2005, and January 2013; temporary change (TC) outliers in February 2007 and January 2015; and a level shift (LS) starting in February 2004.
-In New Hampshire (b), we detect two outliers, both level shift outliers (LS) in April 2010 and again in November 2014.
-These anomalies suggest New York experienced a significant mortality event in September 2001 and New Hampshire experienced approximately 9,700 more deaths than expected since 2010 or 14% more deaths in the state over just seven years.
-\\label{fig:mortnewhamp}")}
-
-a <-  function() {
-plot(outlier.mort.ny, cex = 0.8)
-title("New York - Mortality", adj = 0.9, line =2)
-}
-
-b <-  function() {
-  plot(outlier.mort.nh, cex = 0.8)
-  title("New Hampshire - Mortality", adj = 0.82,line =2)
-}
-
-plot_grid(ggdraw(a) ,
-               ggdraw(b), 
-               ncol=1,
-               labels= "auto")
-```
-
-## Fertility
-
-```{r, include = FALSE}
-la_tstat1 <- prettyNum(outlier.fert.la$outliers$tstat[1], digits = 4) 
-la_tstat2 <- prettyNum(outlier.fert.la$outliers$tstat[2], digits = 4)
-la_averbirths <- prettyNum(abs(sum(outlier.fert.la$y - outlier.fert.la$yadj)), digits = 0, big.mark = ",")
-
-conn_tstat <- prettyNum(outlier.fert.conn$outliers$tstat, digits = 3)
-conn_births <- prettyNum(outlier.fert.conn$outliers$coefhat, digits = 0)
-conn_tot <- prettyNum(abs(sum(outlier.fert.conn$y - outlier.fert.conn$yadj)), digits = 0, big.mark = ",")
-conn_per <- percent(abs(sum(outlier.fert.conn$y - outlier.fert.conn$yadj) / sum(outlier.fert.conn$yadj[80:180])))
-```
-
-To illustrate how researchers can examine fertility anomalies, we begin with a clear temporary change (TC) outlier for fertility in Louisiana (**\autoref{fig:fertla}a**).
-Here we detect two TC outliers, nearly back-to-back in 2005 (2005:08 *t* = `r la_tstat1`; and 2005:10 *t* = `r la_tstat2`), coinciding with the destruction of Hurricane Katrina from the same year.
-The migration associated with Hurricane Katrina has garnered most of the attention of social scientists [@fussellRecoveryMigrationCity2014;@horiDisplacementDynamicsSouthern2009], but the hurricane had a very clear impact on fertility behaviors too, creating a mini "baby bust" in Louisiana with the departure of many people in their childbearing years.
-We estimate `r la_averbirths` fewer births in Louisiana compared to the counterfactual, likely attributable to the Hurricane.
-
-We also highlight a level shift (LS) outlier for fertility in Connecticut.
-Here we detect a strong (*t*= `r conn_tstat` for `r conn_births` births/month) level shift toward lower fertility beginning in August 2009 (2009:08) that continues through the end of the period (**\autoref{fig:fertla}b**).
-This LS toward lower fertility reduced the number of births by `r conn_tot` or `r conn_per`) lower than the counterfactual time series.
-The stock market crash of 2008, where the Dow Jones Industrial Average had the largest single-day loss up to that point, occurred just 11 months before we detect a shift toward lower fertility.
-It is possible that the trend toward lower fertility in August 2009 and the stock market crash in September 2008 are linked, although, as stated previously, we are speculating only to illustrate how causal inference could work in an abductive research agenda.
-A potentially interesting future line of research would be to link fertility anomalies to the kinds of economic shocks that may lead to outmigration of those in their childbearing years.
-
-```{r FertilityAnomalies, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Louisiana (a) and Connecticut (b) state fertility, 2003-2018.} In (a), we detect two outliers, back-to-back, likely resulting from Hurricane Katrina in August and October 2005, representing a decrease of more than 4,400 births due to the hurricane.
-In (b), we detect one outlier, a level shift outlier (LS) in August 2009.
-We believe this reduction is attributable to the stock market crash 11 months earlier.
-\\label{fig:fertla}")}
-
-a <-  function() {
-  plot(outlier.fert.la, cex.axis=0.8)
-  title(main = "Louisiana - Fertility", adj = 0.9)
-}
-
-b <- function(){
-  plot(outlier.fert.conn, cex.axis=0.8) +
-  title(main = "Connecticut - Fertility", adj = 0.88)
-}
-
-plot_grid(ggdraw(a) ,
-               ggdraw(b), 
-               ncol=1,
-               labels= "auto")
-```
-
-## Interesting Anomalous Fertility/Mortality events
-In the examples above, we highlighted four fertility/mortality anomalies with plausible explanations.
-In the case of New York and New Hampshire, the mortality anomalies have plausible explanations.
-It seems likely that New York’s AO anomaly in September 2001 is caused by the 9/11 tragedy and the rise in New Hampshire’s mortality starting in 2010 could be linked to the opioid epidemic.
-Similarly, Louisiana's TC anomalies seem linked to Hurricanes Katrina and Rita while the LS anomaly in Connecticut's fertility appears linked to the Great Recession.
-However, we detect numerous other demographic anomalies in other states, on the causes of which we will not speculate.
-**\autoref{fig:ferthawaii}** shows two such unexplained anomalies.
-
-```{r, include = FALSE}
-hi_tstat <- prettyNum(outlier.fert.hi$outliers$tstat, digits = 3)
-hi_per <- percent(abs(1- outlier.fert.hi$y[137]/outlier.fert.hi$yadj[137]))
-
-oh_tstat <- prettyNum(outlier.mort.oh$outliers$tstat[7], digits = 3)
-oh_deaths <- prettyNum(round(abs(outlier.mort.oh$outliers[7,4]), digits =0), big.mark=",")
-oh_tot <- prettyNum(abs(round(outlier.mort.oh$outliers[7,4]*(228-194), digits = -4)), big.mark=",")
-```
-
-In **\autoref{fig:ferthawaii}a** we identify a single additive anomaly in Hawaiian fertility in May 2014.
-This is a strong anomaly with a *t*-statistic of `r hi_tstat`, `r hi_per` above the counter-factual time series.
-This single, anomalous month is also the second highest monthly births in the time series.
-We have no plausible explanation for this anomaly.
-We do not believe this is simply a data error as the other extreme values, September 2008 and February 2005 with the highest and lowest recorded fertility respectively, were not identified as anomalous events.
-Even if we were to assume this anomaly resulted from data entry error, it remains an *unaltered* data error in the Hawaiian monthly fertility data.
-
-This is contrast to **\autoref{fig:ferthawaii}b**, where we identify a strange mortality reduction in Ohio (*t*-stat: `r oh_tstat`).
-This LS is more than `r oh_deaths` deaths per month less than the counterfactual time series, suggesting had nearly `r oh_tot` fewer deaths since February 2015 than expected.
-This is the single-largest LS among all states.
-We could not identify the potential policies Ohio might have put into place to provide such a strong mortality protection.
-
-```{r TrueAnomalies, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Hawaii state fertility (a) and Ohio state mortality (b).} Here we detect one outlier, an additive outlier (AO) in May 2014 in Hawaii (a), representing a large, unexplainable 14% increase in expected births in that month.
-We also detect a strong level shift (LS) reduction (b), representing a large, unexplainable reduction of nearly 40,000 deaths in Ohio.
-\\label{fig:ferthawaii}")}
-
-a <-  function() {
-plot(outlier.fert.hi, cex.axis=0.8)
-  title(main = "Hawaii - Fertility", adj = 0.88)
-}
-
-b <-  function() {
-plot(outlier.mort.oh, cex.axis=0.8)
-  title(main = "Ohio - Mortality", adj = 0.9)
-}
-
-plot_grid(ggdraw(a) ,
-               ggdraw(b), 
-               ncol=1,
-               labels= "auto")
-```
-
 # Conclusion
 Data scientists frequently claim that the big data revolution is a turning point in scientific discovery that will allow us to solve some of the world’s most pressing problems [@grimmer2015ppsp].
-Social scientists are skeptical of such claims, because they better understand the complexities of the social world and know from experience that data, alone, is not enough [@bohon2018demography;@grimmer2015ppsp].
+Social scientists are skeptical of such claims, because they better understand the complexities of the social world and know from experience that data alone is not enough to solve social problems [@bohon2018demography;@grimmer2015ppsp].
 Nonetheless, the increased availability of data and (more importantly, we argue) the development of advanced techniques for analyzing these data will enable important discovery [@monroe2015no].

### 2020-02-24 MANUSCRIPT/manuscript.tex
Updated. V1.0 of the manuscript.

@@
 \usepackage{makecell}
-\usepackage{xcolor}
-
 \usepackage{booktabs}
@@
 \usepackage{makecell}
+\usepackage{xcolor}
 
@@
 {
-  \title{\bf Causal Inference in Population Trends: Searching for Demographic
-Anomalies in Big Data}
+  \title{\bf Causal Inference in Population Trends: Searching for
+Demographic Anomalies in Big Data}
 
@@
   \begin{center}
-    {\LARGE\bf Causal Inference in Population Trends: Searching for Demographic
-Anomalies in Big Data}
+    {\LARGE\bf Causal Inference in Population Trends: Searching for
+Demographic Anomalies in Big Data}
   \end{center}
@@
 platforms, and the development of powerful statistical algorithms can
-uncover hidden anomalies in demographic data that have previously been
-dismissed as noise. Here, we demonstrate how social scientists can use
-causal inference techniques and abductive reasoning to identify
-fertility and mortality anomalies in state-level data trends by
-investigating demographic time series data since 1999. In demonstrating
-our technique, we uncover real spikes and dips in fertility and
-mortality over time, distinguished from regular trend variations. We
-show that 36 states exhibited at least one mortality anomaly over time,
-totaling nearly 300,000 additional deaths, while an additional seven
-states exhibited periods of decline resulting in 250,000 fewer deaths.
-Ten states exhibited real, marked periods of declines in fertility,
-totaling more than 214,000 fewer births than expected, and 8 states had
-periods of excess births, resulting in an addition 12,000 births. Our
-results show how researchers can detect demographic anomalies in trend
-data and use those anomalies as a beginning point for further
-investigation into the causes.
+uncover hidden anomalies in social data, previously dismissed as noise.
+Here, we combine causal inference techniques and abductive reasoning to
+identify fertility and mortality anomalies on twenty years of
+demographic data in the United States. We uncover real, ``hidden'' baby
+booms/busts and mortality spikes/dips, distinguishable from regular
+trend variations. We identify more than 22 and 156 fertility and
+mortality anomalies, totaling more than 200k and 600k anomalous births
+and deaths, respectively. Notable detectable mortality anomalies include
+the September 11 2001 terrorist attack in New York, the emergence and
+acceleration of the opioid epidemic in New Hampshire, and the excess
+mortality due to Hurricane Maria in Puerto Rico. Notable fertility
+anomalies include the ``missing births'' in Louisiana after Hurricane
+Katrina and the reduction in fertility behavior after the September 2008
+stock market crash in Connecticut, amongst others. Our combined causal
+inference and abductive reasoning approach can be readily adapted to
+find other, undiscovered social phenomena or to evaluate the efficacy of
+important public policies.
 \end{abstract}
@@
 \noindent%
-{\it Keywords:} mortality, fertility, causal inference, abductive reasoning, big data
+{\it Keywords:} mortality, fertility, causal inference, abductive
+reasoning, big data
 \vfill
@@
 
-The proliferation of data, advances in high performance (``super'')
-computing, and the development of powerful statistical algorithms mark
-the era of ``Big Data'' or data science
-\citep{van2016data, zikopoulos2011}, with the potential for researchers
-to find hidden or understudied social phenomena
+The search for causality is one of the primary motiviations behind
+social science \citep{smith2011commentary, pearl2009causal}. Causal
+Inference, the identification of causal mechanisms, attempts to uncover
+the underlying mechanisms that result in changes in a phenomenon and has
+a long history as an approach to inquiry in the social sciences
+\citep{grimmer2015ppsp}. Often, the identification of a casual mechanism
+is accomplished via a randomized control trial (RCT) or expert knowledge
+applied to a natural experiment \citep{salganik2019bit}.
+
+Unfortunately, both RCTs and natural experiments have drawbacks. RCTs
+are often prehibitively expensive for many scientists, precluding their
+widespread adoption in the social sciences \citep{west2008ajph}. RCTs
+are also sometimes impossible to deploy in the ``real world''. For
+instance, an RCT to study country-level austeriy measures would be
+difficult, if not impossible, to implement. Natural experiments are a
+cheaper alternative, but provide less certainty about causation, fewer
+opportunities to replicate work, and require expert knowledge of the
+natural experiment to be properly employed \citep{pearl2018book}.
+
+Additionally, both RCTs and natural experiments follow traditional
+hypothesis testing, the deductive scientific paradigm where a question
+is first posed and scientists generate or find data to answer the
+question. Such approaches, while extensively utilized and time-tested,
+are likely to miss important phenomena that might go unnoticed or suffer
+from extensive p-hacking
+\citep{head2015extent, nuzzo2014scientific, ruggles2014big}. Abductive
+reasoning, sometimes described as ``inferring cause from effect''
+\citep{Crowder2017}, moves the scientific process from the inductive to
+the deductive, and sometimes back and forth, to reach conclusions
+\citep{bryant2014realm} The physical sciences use abductive reasoning
+more fully compared to the social sciences. The discovery of Neptune in
+1846 based on unexplained perturbations of the orbit of Uranus
+\citep{popper2005logic}, the discovery of galactic migration of planets
+\citep{gomes2005n}, and the current generation of gravitational wave
+detectors such as the Laser Interferometer Gravitational-Wave
+Observatory (LIGO) \citep{harry2010advanced} demonstrate the power of
+the abductive approach in the physical sciences.
+
+In a data-rich environment, abductive reasoning can be more easily
+applied to first find effects and then infer their cause ex-post. The
+modern proliferation of data, advances in high performance (``super'')
+computing, and the development of powerful statistical algorithms
+provide an environment ripe for wedding causal inference with abductive
+reasoning \citep{van2016data, zikopoulos2011}, with the potential for
+researchers to find hidden or understudied social phenomena
 \citep{bohon2018demography}. While the availability of Big Data and high
@@
 \citep{angrist1989lifetime, mas2009peers} and allows us to better
-monitor population trends \citep{nobles2019, torche2015hidden}.
+monitor population trends \citep{nobles2019, torche2015hidden}. The rich
+demographic data available in the United States make the potential
+revelation of interesting and important demographic phenomena not only
+possible, but extremely plausible -- even if identification of the
+phenomena occurs without identifying the underlying cause.
 
-Causal inference, in its most common usage, is an attempt to uncover the
-underlying mechanisms that result in changes in a phenomenon and has a
-long history as an approach to inquiry in the social sciences
-\citep{grimmer2015ppsp}. Often, the identification of a casual mechanism
-requires either a randomized control trial (RCT) or expert knowledge
-applied to a natural experiment. For population-level analysis,
-expensive RCTs usually preclude their widespread adoption
-\citep{west2008ajph}. Using natural experiments provide less certainty
-about causation and fewer opportunities to replicate work but they
-provide important and widespread insights.
-
-Both RCTs and natural experiments follow traditional hypothesis
-testing--inductive scientific paradigms where a question is first posed
-and scientists generate or find data to answer the question. Such
-approaches, while extensively utilized and time-tested, are likely to
-miss important phenomena that might go unnoticed. Data scientists are
-rethinking these approaches and have re-envisioned causal inference as a
-machine learning technique that allows big data researchers to move from
-correlations that have been established with good accuracy to making
-strong conclusions about the underlying mechanisms that produce these
-correlations \citep{pearl2018book}.
-
-Abductive modeling is the movement from the inductive to the deductive,
-and sometimes back and forth, to reach conclusions
-\citep{bryant2014realm} or ``inferring cause from effect''
-\citep{Crowder2017}. In situations with large data sets, an abductive
-approach is far superior to deductive hypothesis testing, as p-values
-with an extremely large number of cases are far from revealing
-\citep{head2015extent, nuzzo2014scientific} and we do not want to make
-purely inductive inferences from data of questionable generalizability
-\citep{ruggles2014big}. Owing to the long history of big data sets in
-demographic research (one of the original ``big data'' resources
-\citep{ruggles2014big}), the rich demographic data available in the
-United States make the potential revelation of interesting and important
-demographic phenomena not only possible, but extremely plausible -- even
-if identification of the phenomena occurs without identifying the
-underlying cause. In other sciences, the identification of effects
-without causes has led to new theories of galactic migration of planets
-and other breakthroughs \citep{gomes2005n}.
-
-In this paper, we use modern statistical outlier detection algorithms
-\citep{chen1993joint} as a means to illustrate one way that social
-scientists can engage in discovery using big data techniques on nearly
-twenty years of mortality and fertility data at the US state-level to
+In this paper, we combine a modern statistical outlier detection
+algorithm \citep{chen1993joint}, abductive reasoning, and large
+demographic data sets on more than 20 years of mortality/fertility to
 identify anomalous demographic behavior. In essence, we identify
-\emph{effects without knowledge of the cause}. We ask two questions
-regarding demographic anomalies: What are the hidden baby booms/busts
-and mortality spikes/dips in the United States over the last twenty
-years? We do not necessarily know the causes of these anomalies but
-identifying them allows scientists with more detailed knowledge of local
-population dynamics, state-level policy making, or macro-economics to
-explain these phenomena post-hoc. From explanations of phenomena that
+\emph{effects without knowledge of the cause}. We ask one fundamanetal
+question regarding demographic anomalies: What are the hidden baby
+booms/busts and mortality spikes/dips in the United States over the last
+twenty years? We do not necessarily know the causes of these anomalies
+but identifying them allows scientists with more detailed knowledge of
+local population dynamics, state-level policy making, or macro-economics
+to explain these phenomena post-hoc. From explanations of phenomena that
 may have previously gone unnoticed, social scientists may be able to
@@
 state-level since 1999. A full listing of these anomalies can be found
-in the \textbf{Supplementary Materials}. We begin with a summary of the
-anomalies we detect and then we highlight highly significant mortality
-and fertility anomalies across all three types of outliers (Additive
-Outliers, Level Shift Outliers, and Temporary Change Outliers) with
-plausible explanations. Finally, we highlight two strong anomalies that
-bely explanation.
+in the \textbf{Supplementary Materials}. We begin by highlighting
+significant fertility and mortality anomaloies across all three types of
+outliers (Additive Outliers, Level Shift Outliers, and Temporary Change
+outliers) with plausible explanations. We then highlight two strong
+anomalies that bely explanation. Finally, we conclude with a summary of
+the anomalies we detect.
+
+\hypertarget{fertility}{%
+\subsection{Fertility}\label{fertility}}
+
+We begin with a clear temporary change (TC) outlier for fertility in
+Louisiana (\textbf{\autoref{fig:fertla}a}). Here we detect two TC
+outliers, nearly back-to-back in 2005 (2005:08 \emph{t} = -6.002; and
+2005:10 \emph{t} = -4.997), coinciding with the destruction of Hurricane
+Katrina from the same year. The migration associated with Hurricane
+Katrina has garnered most of the attention of social scientists
+\citep{fussellRecoveryMigrationCity2014, horiDisplacementDynamicsSouthern2009},
+but the hurricane had a very clear impact on fertility behaviors too,
+creating a mini ``baby bust'' in Louisiana with the departure of many
+people in their childbearing years. We estimate 4,411 fewer births in
+Louisiana compared to the counterfactual, likely attributable to the
+Hurricane.
+
+We also highlight a level shift (LS) outlier for fertility in
+Connecticut. Here we detect a strong (\emph{t}= -3.71 for -285
+births/month) level shift toward lower fertility beginning in August
+2009 (2009:08) that continues through the end of the period
+(\textbf{\autoref{fig:fertla}b}). This LS toward lower fertility reduced
+the number of births by 28,752 or 9\%) lower than the counterfactual
+time series. The stock market crash of 2008, where the Dow Jones
+Industrial Average had the largest single-day loss up to that point,
+occurred just 11 months before we detect a shift toward lower fertility.
+It is possible that the trend toward lower fertility in August 2009 and
+the stock market crash in September 2008 are linked, although, as stated
+previously, we are speculating only to illustrate how causal inference
+could work in an abductive research agenda. A potentially interesting
+future line of research would be to link fertility anomalies to the
+kinds of economic shocks that may lead to outmigration of those in their
+childbearing years.
+
+\begin{figure}
+\centering
+\includegraphics{manuscript_files/figure-latex/FertilityAnomalies-1.pdf}
+\caption{\textbf{Anomaly detection in Louisiana (a) and Connecticut (b) state fertility, 2003-2018.}
+The top part of each panel contains the original time series (light
+gray), the corrected, counter-factual time series in the absence of
+anomalies (blue), and the red dots correspond to the onset of detected
+anomalies. The bottom part of each panel contains the magnitude and type
+of the outlier in red. In (a), we detect two outliers, back-to-back,
+likely resulting from Hurricane Katrina in August and October 2005,
+representing a decrease of more than 4,400 births due to the hurricane
+(t-statistics= -6.002 and -4.997). In (b), we detect one outlier, a
+level shift outlier (LS) in August 2009. We believe this reduction is
+attributable to the stock market crash 11 months earlier (t-statistic =
+-3.71). \label{fig:fertla}}
+\end{figure}
+
+\hypertarget{mortality}{%
+\subsection{Mortality}\label{mortality}}
+
+For mortality, we will use New York State as an example
+(\textbf{\autoref{fig:mortnewhamp}a}). We identify seven anomalies in
+the mortality time series for New York, all with \emph{t}-statistics in
+excess of 3.91, making these anomalies unlikely to be due to chance. The
+algorithm correctly identifies September 2001 as an additive outlier
+(2001:09 \emph{t}= 6.396) where there were 1,628 more deaths in that
+month than anticipated. This mortality event is likely caused by the
+September 11 terrorist attack on the World Trade Center that immediately
+killed 2,606 people and the detection of this mortality event provides
+confidence in our detection of other anomalies.
+
+In \textbf{\autoref{fig:mortnewhamp}a}, notice the strong level shift
+(LS) that occurs in February 2004 (2004:02 \emph{t}= -5.594) which
+prevented 869 deaths per month. This shift totals more than 144,000
+averted deaths compared to the counter-factual time series and is the
+single largest mortality protective anomaly among all states. This
+translates to 7\% fewer deaths than expected over the time period. What
+is driving this mortality protection? What policies did NY put into
+place that might have contributed to this considerable mortality
+reduction? What environmental or economic conditions may have changed
+may have changed? These are the kinds of questions that arise from our
+analyses. The purpose of our paper is not to answer these questions, but
+our findings underscore the need for more research that takes an
+abductive approach. By identifying anomalies through an inductive
+process, researchers can then look for underlying causes. Once those
+causes are identified, researchers can then use a deductive process to
+see if such events predict other (or future) anomalies.
+
+To see the potential for combining abductive reasoning with causal
+inference, contrast the mortality protection in New York with the
+enhanced mortality in New Hampshire
+(\textbf{\autoref{fig:mortnewhamp}b}). In New Hampshire we detect two
+significant level shifts (LS) in the monthly mortality data, first in
+April 2010 and again in November 2014 (2010:04 \emph{t}= 3.51; 2014:11
+\emph{t}= 6.68). These anomalies suggest New Hampshire experienced 8,159
+more deaths (+9\% more than expected) in a seven-year period beginning
+in early 2010. These are events not experienced by neighboring states
+during the same time period, and this is the single largest percentage
+mortality increase/decrease we detected among all states. Not
+coincidentally, NH has the second highest opioid-related mortality in
+the US \citep{beetham2019access} and it is likely that we detect this
+epidemic in our results. Isolating these anomalies and testing them
+against opioid sales data might yield intriguing results.
+
+\begin{figure}
+\centering
+\includegraphics{manuscript_files/figure-latex/MortalityAnomalies-1.pdf}
+\caption{\textbf{Anomaly Detection for New York (a) and New Hampshire (b) state mortality, 1999-2016.}
+In New York (a), we detect additive outliers (AO) in January 2000,
+September 2001, January 2005, and January 2013; temporary change (TC)
+outliers in February 2007 and January 2015; and a level shift (LS)
+starting in February 2004. In New Hampshire (b), we detect two outliers,
+both level shift outliers (LS) in April 2010 and again in November 2014.
+These anomalies suggest New York experienced a significant mortality
+event in September 2001 and New Hampshire experienced approximately
+9,700 more deaths than expected since 2010 or 14\% more deaths in the
+state over just seven years. \label{fig:mortnewhamp}}
+\end{figure}
+
+\hypertarget{interesting-anomalous-fertilitymortality-events}{%
+\subsection{Interesting Anomalous Fertility/Mortality
+events}\label{interesting-anomalous-fertilitymortality-events}}
+
+In the examples above, we highlighted four fertility/mortality anomalies
+with plausible explanations. In the case of New York and New Hampshire,
+the mortality anomalies have plausible explanations. It seems likely
+that New York's AO anomaly in September 2001 is caused by the 9/11
+tragedy and the rise in New Hampshire's mortality starting in 2010 could
+be linked to the opioid epidemic. Similarly, Louisiana's TC anomalies
+seem linked to Hurricanes Katrina and Rita while the LS anomaly in
+Connecticut's fertility appears linked to the Great Recession. However,
+we detect numerous other demographic anomalies in other states, on the
+causes of which we will not speculate. \textbf{\autoref{fig:ferthawaii}}
+shows two such unexplained anomalies.
+
+In \textbf{\autoref{fig:ferthawaii}a} we identify a single additive
+anomaly in Hawaiian fertility in May 2014. This is a strong anomaly with
+a \emph{t}-statistic of 4.96, 15\% above the counter-factual time
+series. This single, anomalous month is also the second highest monthly
+births in the time series. We have no plausible explanation for this
+anomaly. We do not believe this is simply a data error as the other
+extreme values, September 2008 and February 2005 with the highest and
+lowest recorded fertility respectively, were not identified as anomalous
+events. Even if we were to assume this anomaly resulted from data entry
+error, it remains an \emph{unaltered} data error in the Hawaiian monthly
+fertility data.
+
+This is contrast to \textbf{\autoref{fig:ferthawaii}b}, where we
+identify a strange mortality reduction in Ohio (\emph{t}-stat: -4.27).
+This LS is more than 1,157 deaths per month less than the counterfactual
+time series, suggesting had nearly 40,000 fewer deaths since February
+2015 than expected. This is the single-largest LS among all states. We
+could not identify the potential policies Ohio might have put into place
+to provide such a strong mortality protection.
+
+\begin{figure}
+\centering
+\includegraphics{manuscript_files/figure-latex/TrueAnomalies-1.pdf}
+\caption{\textbf{Anomaly detection in Hawaii state fertility (a) and Ohio state mortality (b).}
+Here we detect one outlier, an additive outlier (AO) in May 2014 in
+Hawaii (a), representing a large, unexplainable 14\% increase in
+expected births in that month. We also detect a strong level shift (LS)
+reduction (b), representing a large, unexplainable reduction of nearly
+40,000 deaths in Ohio. \label{fig:ferthawaii}}
+\end{figure}
 
@@
 
-\caption{\label{tab:unnamed-chunk-3}\textbf{Summary by Demographic Component.}  Here we can see there 22 Fertility and 156 Mortality anomalies among the state-level time series, totalling more than 200k anomalous births and 600k anomalous deaths.
+\caption{\label{tab:unnamed-chunk-6}\textbf{Summary by Demographic Component.}  Here we can see there 22 Fertility and 156 Mortality anomalies among the state-level time series, totalling more than 200k anomalous births and 600k anomalous deaths.
 \label{sumtable}}
@@
 across all anomaly types and \textbf{\autoref{fig:summap}} maps these
-results. We find considerably more mortality anomalies (n = 156) than
-fertility anomalies (n = 22). Given that anomalies are always the
-product of events \citep{song2018anomaly}, finding more mortality than
-fertility anomalies is not surprising. Mortality is likely to spike in
-response to a catastrophic event (like an earthquake or terrorist
-attack) or due to a disease outbreak while the effects of a catastrophic
-event on fertility is less predictable. One reason for this is that
-fertility is linked to human decision-making more directly than
-mortality \citep{stein2014couples} which results in more varied
-outcomes, so, for example, researchers find that catastrophic weather
-events can increase childbearing among those who already have a child,
-but not among the childless \citep{Evans2008Hurricanebirth}. Some
-events, like the 1995 Oklahoma City bombings, resulted in both fertility
-and mortality changes. However, changes in fertility manifest over a
-longer time horizon after an event, and the relationship between a
+results.
+
+We find considerably more mortality anomalies (n = 156) than fertility
+anomalies (n = 22). Given that anomalies are always the product of
+events \citep{song2018anomaly}, finding more mortality than fertility
+anomalies is not surprising. Mortality is likely to spike in response to
+a catastrophic event (like an earthquake or terrorist attack) or due to
+a disease outbreak while the effects of a catastrophic event on
+fertility is less predictable. One reason for this is that fertility is
+linked to human decision-making more directly than mortality
+\citep{stein2014couples} which results in more varied outcomes, so, for
+example, researchers find that catastrophic weather events can increase
+childbearing among those who already have a child, but not among the
+childless \citep{Evans2008Hurricanebirth}. Some events, like the 1995
+Oklahoma City bombings, resulted in both fertility and mortality
+changes. However, changes in fertility manifest over a longer time
+horizon after an event, and the relationship between a
 fertility-inducing event and behavior change is less strong than the
@@
 
-\hypertarget{mortality}{%
-\subsection{Mortality}\label{mortality}}
-
-Since the purpose of our paper is to demonstrate how social scientists
-can make use of causal inference technique, we will not discuss all of
-our findings in this paper. Instead, we will use a few cases for
-illustration. For mortality, we will use New York State as an example
-(\textbf{\autoref{fig:mortnewhamp}a}). We identify seven anomalies in
-the mortality time series for New York, all with \emph{t}-statistics in
-excess of 3.91, making these anomalies unlikely to be due to chance. The
-algorithm correctly identifies September 2001 as an additive outlier
-(2001:09 \emph{t}= 6.396) where there were 1,628 more deaths in that
-month than anticipated. This mortality event is likely caused by the
-September 11 terrorist attack on the World Trade Center that immediately
-killed 2,606 people and the detection of this mortality event provides
-confidence in our detection of other anomalies.
-
-In \textbf{\autoref{fig:mortnewhamp}a}, notice the strong level shift
-(LS) that occurs in February 2004 (2004:02 \emph{t}= -5.594) which
-prevented 869 deaths per month. This shift totals more than 144,000
-averted deaths compared to the counter-factual time series and is the
-single largest mortality protective anomaly among all states. This
-translates to 6.90\% fewer deaths than expected over the time period.
-What is driving this mortality protection? What policies did NY put into
-place that might have contributed to this considerable mortality
-reduction? What environmental or economic conditions may have changed
-may have changed? These are the kinds of questions that arise from our
-analyses. The purpose of our paper is not to answer these questions, but
-our findings underscore the need for more research that takes an
-abductive approach. By identifying anomalies through an inductive
-process, researchers can then look for underlying causes. Once those
-causes are identified, researchers can then use a deductive process to
-see if such events predict other (or future) anomalies.
-
-To see the potential for combining abductive reasoning with causal
-inference, contrast the mortality protection in New York with the
-enhanced mortality in New Hampshire
-(\textbf{\autoref{fig:mortnewhamp}b}). In New Hampshire we detect two
-significant level shifts (LS) in the monthly mortality data, first in
-April 2010 and again in November 2014 (2010:04 \emph{t}= 3.51; 2014:11
-\emph{t}= 6.68). These anomalies suggest New Hampshire experienced 8,159
-more deaths (+9.24\% more than expected) in a seven-year period
-beginning in early 2010. These are events not experienced by neighboring
-states during the same time period, and this is the single largest
-percentage mortality increase/decrease we detected among all states. Not
-coincidentally, NH has the second highest opioid-related mortality in
-the US \citep{beetham2019access} and it is likely that we detect this
-epidemic in our results. Isolating these anomalies and testing them
-against opioid sales data might yield intriguing results.
-
-\begin{figure}
-\centering
-\includegraphics{manuscript_files/figure-latex/MortalityAnomalies-1.pdf}
-\caption{\textbf{Anomaly Detection for New York (a) and New Hampshire (b) state mortality, 1999-2016.}
-The top part of each panel contains the original time series (light
-gray), the corrected, counter-factual time series in the absence of
-anomalies (blue), and the red dots correspond to the onset of detected
-anomalies. The bottom part of each panel contains the magnitude and type
-of the outlier in red. In New York (a), we detect additive outliers (AO)
-in January 2000, September 2001, January 2005, and January 2013;
-temporary change (TC) outliers in February 2007 and January 2015; and a
-level shift (LS) starting in February 2004. In New Hampshire (b), we
-detect two outliers, both level shift outliers (LS) in April 2010 and
-again in November 2014. These anomalies suggest New York experienced a
-significant mortality event in September 2001 and New Hampshire
-experienced approximately 9,700 more deaths than expected since 2010 or
-14\% more deaths in the state over just seven years.
-\label{fig:mortnewhamp}}
-\end{figure}
-
-\hypertarget{fertility}{%
-\subsection{Fertility}\label{fertility}}
-
-To illustrate how researchers can examine fertility anomalies, we begin
-with a clear temporary change (TC) outlier for fertility in Louisiana
-(\textbf{\autoref{fig:fertla}a}). Here we detect two TC outliers, nearly
-back-to-back in 2005 (2005:08 \emph{t} = -6.002; and 2005:10 \emph{t} =
--4.997), coinciding with the destruction of Hurricane Katrina from the
-same year. The migration associated with Hurricane Katrina has garnered
-most of the attention of social scientists
-\citep{fussellRecoveryMigrationCity2014, horiDisplacementDynamicsSouthern2009},
-but the hurricane had a very clear impact on fertility behaviors too,
-creating a mini ``baby bust'' in Louisiana with the departure of many
-people in their childbearing years. We estimate 4,411 fewer births in
-Louisiana compared to the counterfactual, likely attributable to the
-Hurricane.
-
-We also highlight a level shift (LS) outlier for fertility in
-Connecticut. Here we detect a strong (\emph{t}= -3.71 for -285
-births/month) level shift toward lower fertility beginning in August
-2009 (2009:08) that continues through the end of the period
-(\textbf{\autoref{fig:fertla}b}). This LS toward lower fertility reduced
-the number of births by 28,752 or 8.57\%) lower than the counterfactual
-time series. The stock market crash of 2008, where the Dow Jones
-Industrial Average had the largest single-day loss up to that point,
-occurred just 11 months before we detect a shift toward lower fertility.
-It is possible that the trend toward lower fertility in August 2009 and
-the stock market crash in September 2008 are linked, although, as stated
-previously, we are speculating only to illustrate how causal inference
-could work in an abductive research agenda. A potentially interesting
-future line of research would be to link fertility anomalies to the
-kinds of economic shocks that may lead to outmigration of those in their
-childbearing years.
-
-\begin{figure}
-\centering
-\includegraphics{manuscript_files/figure-latex/FertilityAnomalies-1.pdf}
-\caption{\textbf{Anomaly detection in Louisiana (a) and Connecticut (b) state fertility, 2003-2018.}
-In (a), we detect two outliers, back-to-back, likely resulting from
-Hurricane Katrina in August and October 2005, representing a decrease of
-more than 4,400 births due to the hurricane. In (b), we detect one
-outlier, a level shift outlier (LS) in August 2009. We believe this
-reduction is attributable to the stock market crash 11 months earlier.
-\label{fig:fertla}}
-\end{figure}
-
-\hypertarget{interesting-anomalous-fertilitymortality-events}{%
-\subsection{Interesting Anomalous Fertility/Mortality
-events}\label{interesting-anomalous-fertilitymortality-events}}
-
-In the examples above, we highlighted four fertility/mortality anomalies
-with plausible explanations. In the case of New York and New Hampshire,
-the mortality anomalies have plausible explanations. It seems likely
-that New York's AO anomaly in September 2001 is caused by the 9/11
-tragedy and the rise in New Hampshire's mortality starting in 2010 could
-be linked to the opioid epidemic. Similarly, Louisiana's TC anomalies
-seem linked to Hurricanes Katrina and Rita while the LS anomaly in
-Connecticut's fertility appears linked to the Great Recession. However,
-we detect numerous other demographic anomalies in other states, on the
-causes of which we will not speculate. \textbf{\autoref{fig:ferthawaii}}
-shows two such unexplained anomalies.
-
-In \textbf{\autoref{fig:ferthawaii}a} we identify a single additive
-anomaly in Hawaiian fertility in May 2014. This is a strong anomaly with
-a \emph{t}-statistic of 4.96, 14.9\% above the counter-factual time
-series. This single, anomalous month is also the second highest monthly
-births in the time series. We have no plausible explanation for this
-anomaly. We do not believe this is simply a data error as the other
-extreme values, September 2008 and February 2005 with the highest and
-lowest recorded fertility respectively, were not identified as anomalous
-events. Even if we were to assume this anomaly resulted from data entry
-error, it remains an \emph{unaltered} data error in the Hawaiian monthly
-fertility data.
-
-This is contrast to \textbf{\autoref{fig:ferthawaii}b}, where we
-identify a strange mortality reduction in Ohio (\emph{t}-stat: -4.27).
-This LS is more than 1,157 deaths per month less than the counterfactual
-time series, suggesting had nearly 40,000 fewer deaths since February
-2015 than expected. This is the single-largest LS among all states. We
-could not identify the potential policies Ohio might have put into place
-to provide such a strong mortality protection.
-
-\begin{figure}
-\centering
-\includegraphics{manuscript_files/figure-latex/TrueAnomalies-1.pdf}
-\caption{\textbf{Anomaly detection in Hawaii state fertility (a) and Ohio state mortality (b).}
-Here we detect one outlier, an additive outlier (AO) in May 2014 in
-Hawaii (a), representing a large, unexplainable 14\% increase in
-expected births in that month. We also detect a strong level shift (LS)
-reduction (b), representing a large, unexplainable reduction of nearly
-40,000 deaths in Ohio. \label{fig:ferthawaii}}
-\end{figure}
-
 \hypertarget{conclusion}{%
@@
 scientists are skeptical of such claims, because they better understand
-the complexities of the social world and know from experience that data,
-alone, is not enough \citep{bohon2018demography, grimmer2015ppsp}.
-Nonetheless, the increased availability of data and (more importantly,
-we argue) the development of advanced techniques for analyzing these
-data will enable important discovery \citep{monroe2015no}.
+the complexities of the social world and know from experience that data
+alone is not enough to solve social problems
+\citep{bohon2018demography, grimmer2015ppsp}. Nonetheless, the increased
+availability of data and (more importantly, we argue) the development of
+advanced techniques for analyzing these data will enable important
+discovery \citep{monroe2015no}.
 

### 2020-02-24 R/SANDBOX/PR_mort.R

@@ -1,18 +1,29 @@
 
 # Reading in the mortality data
-PRdeaths <- readxl::read_xlsx("DATA/Deaths0918.xlsx", sheet =1)
+PRdeaths <- readxl::read_xlsx("./R/DATA-RAW/Deaths0918.xlsx", sheet =1)
+
+PRdeaths <- readxl::read_xlsx("./R/DATA-RAW/deahts_pr_4Hauer.xlsx", sheet =1) %>%
+  pivot_longer(-Year, names_to = "Month", values_to = "Deaths") %>%
+  mutate(Month3 = AsDate(paste0(Month, " ", Year))) %>%
+  filter(Year >= 1999) 
+
+z <- ts(PRdeaths$Deaths, 
+        start = year(min(PRdeaths$Month3)), 
+        end = c(year(max(PRdeaths$Month3)),12), frequency = 12)
 
 # Generating the TS outliers using just the `Deaths`
-outlier.morttots <- tsoutliers::tso(ts(PRdeaths$Deaths),  maxit.iloop=10)
+outlier.morttots <- tsoutliers::tso(ts(PRdeaths$Deaths, 
+                                       start = year(min(PRdeaths$Month3)), 
+                                       end = c(year(max(PRdeaths$Month3)),12), frequency = 12),  maxit.iloop=10)
 
 # Generating the TS outliers using the `Deaths` and the regressor variable `Population_Estimate`
-outlier.mortots.popest <- tsoutliers::tso(ts(PRdeaths$Deaths), xreg = PRdeaths$Population_Estimate,  maxit.iloop=10)
+# outlier.mortots.popest <- tsoutliers::tso(ts(PRdeaths$Deaths), xreg = PRdeaths$Population_Estimate,  maxit.iloop=10)
 
 # Plotting the figure for the total `Deaths`
 plot(outlier.morttots)
-
+outlier.morttots
 # Summing up the difference in the Mortality using only the `Deaths`
-sum(outlier.morttots$y - outlier.morttots$yadj)
+sum(outlier.morttots$y[225:240] - outlier.morttots$yadj[225:240])
 
 # Summing up the difference using the `Deaths` and the `Population_Estimate` regressor
 sum(outlier.mortots.popest$y - outlier.mortots.popest$yadj)

### 2020-02-24 R/SCRIPTS/000-Libraries.R

@@ -64,7 +64,8 @@ pkgs <- c(
   "bsts",
   "lubridate",
   "tsoutliers",
-  "numform"
+  "numform",
+  "statebins"
 )
 
 # Install missing packages

### 2020-08-03 MainDocument/MainDocument.Rmd
Updated.

--- 2020-02-24 MANUSCRIPT/manuscript.Rmd
@@
 
-bibliography: mybibfile
+bibliography: ../MainDocument/mybibfile.bib
+
 output: rticles::asa_article
+
 header-includes:
@@
 
-The search for causality is one of the primary motiviations behind social science [@smith2011commentary; @pearl2009causal].
+The search for causality is one of the primary motivations behind social science [@smith2011commentary; @pearl2009causal].
 Causal Inference, the identification of causal mechanisms, attempts to uncover the underlying mechanisms that result in changes in a phenomenon and has a long history as an approach to inquiry in the social sciences [@grimmer2015ppsp].
@@
 Unfortunately, both RCTs and natural experiments have drawbacks.
-RCTs are often prehibitively expensive for many scientists, precluding their widespread adoption in the social sciences [@west2008ajph].
+RCTs are often prohibitively expensive for many scientists, precluding their widespread adoption in the social sciences [@west2008ajph].
 RCTs are also sometimes impossible to deploy in the "real world".
-For instance, an RCT to study country-level austeriy measures would be difficult, if not impossible, to implement.
+For instance, an RCT to study country-level austerity measures would be difficult, if not impossible, to implement.
 Natural experiments are a cheaper alternative, but provide less certainty about causation, fewer opportunities to replicate work, and require expert knowledge of the natural experiment to be properly employed [@pearl2018book].
@@
 In essence, we identify *effects without knowledge of the cause*.
-We ask one fundamanetal question regarding demographic anomalies: What are the hidden baby booms/busts and mortality spikes/dips in the United States over the last twenty years?
+We ask one fundamental question regarding demographic anomalies: What are the hidden baby booms/busts and mortality spikes/dips in the United States over the last twenty years?
 We do not necessarily know the causes of these anomalies but identifying them allows scientists with more detailed knowledge of local population dynamics, state-level policy making, or macro-economics to explain these phenomena post-hoc.
@@
 We use a statistical time series outlier detection algorithm [@chen1993joint], implemented in the R programming language [@rcore] via the tsoutliers package [@tsoutliers2019].
-This algorithm iteratively uses ARIMA models to 1) produce a counter-factual time series to initially detect an outlier or anomaly, and 2) refit the ARIMA with the outliers removed.
+This algorithm iteratively uses ARIMA models to 1) identify potential outliers or anomalies and 2) refit the ARIMA with the outliers removed to produce a counter-factual time series.
 Here we briefly summarize and describe the method.
@@
 Often, the behavior of a time series can be described and summarized in ARIMA models.
-If a series of values, $y_t^*$, is subject to $m$ interventions or outliers at time points $t_1,t_2,…,t_m$, then $y_t^*$ can be defined as
+If a series of values, $y_t^*$, is subject to $m$ interventions or outliers at time points $t_1,t_2,…,t_m$ with weights $\omega$ then $y_t^*$ can be defined as
 
-$$y_t^* = \sum_{j=1}^{m} \omega_jL_j(B)I_t(t_j) + \frac{\theta(B)}{\phi(B)\alpha(B)}\alpha_t$$
+\begin{equation}
+\label{eq:arima}
+y_t^* = \sum_{j=1}^{m} \omega_jL_j(B)I_t(t_j) + \frac{\theta(B)}{\phi(B)\alpha(B)}\alpha_t 
+\end{equation}
 
@@
 We examine three types of outliers at time point $t_m$:  
-1) additive outliers (AO), defined as $L_j(B)=1$;
-2) level shift outliers (LS), defined as $L_j(B) = 1/(1-B)$; and
-2) temporary change outliers (TC), defined as $L_j(B) = 1/(1-\delta B)$.
+1. additive outliers (AO), defined as $L_j(B)=1$;  
+2. level shift outliers (LS), defined as $L_j(B) = 1/(1-B)$; and  
+3. temporary change outliers (TC), defined as $L_j(B) = 1/(1-\delta B)$ where $\delta$ is equal to 0.7.
 
@@
 
-An outlier is detected using a regression equation
+An outlier is detected with the estimated residuals using a regression equation
 
-$$ \pi(B)y_t^* \equiv \hat{e} = \sum_{j=1}^m \omega_j \pi(B)L_j(B)I_t(t_j) + \alpha_t $$
+\begin{equation}
+\label{eq:residuals}
+\pi(B)y_t^* \equiv \hat{e} = \sum_{j=1}^m \omega_j \pi(B)L_j(B)I_t(t_j) + \alpha_t
+\end{equation}
+where $\pi(B)=\sum_{i=o}^{inf} \pi_iB^i$.
+
+Equations \ref{eq:arima} and \ref{eq:residuals} allow for an automatic detection of anomalies iterated over a two-stage process.
+
+In stage 1, anomalies are located.
+First, an ARIMA model is fit to the time series using the `forecast` package in R [@Rforecast;@hymdman2008] where the best performing ARIMA model is selected based on the Bayesian information criterion (BIC).
+Next, the residuals from the forecast are checked for their significance using equation \ref{eq:residuals} where only anomalies above a critical *t*-static are considered "true" anomalies ($|\tau| \geq 3.5$).
+We choose this critical value based on Chen and Liu's [-@chen1993joint] recommendation to minimize Type I or false-positive anomalies.
+Finally, two additional rules are implemented: If multiple anomalies are detected at the same time point, only the most significant anomaly is selected and if anomalies of the same type at consecutive time periods are detected, only the anomaly with highest *t*-statistic is selected.
+
+In stage 2, anomalies are removed from the time series and a new ARIMA model is chosen and fit.
+The selection of the initial ARIMA model could have been affected by the presence of the anomalies, making some anomalies spuriously identified.
+To correct for this, a new ARIMA model is fit accounting for additional regression effects in equation \ref{eq:arima} from the list of candidate anomalies identified in stage 1, effectively removing the anomalies from the time series.
+Each anomaly is then reassessed under the new model and those anomalies that are no longer significant are removed.
+
+These two stages are then iterated until no additional anomalies are detected.
+
+## Robustness Check
+To determine the false-positive (Type I) and false-negative (Type II) error rates under a critical value of 3.5, we ran two simulations.
+For the false-positive error we simulated 1000 time series with 180 time steps without introducing any anomalies and ran the `tsoutliers` algorithm on them.
+Any anomalies we detect in these time series are spurious anomalies.
+For the false-negative error, we simulated 1000 time series with 180 time steps but introduced a single anomaly of the same magnitude at the same time point in each time series.
+If the `tsoutliers` algorithm fails to detect the anomaly, it would be considered a false-negative error.
+
+```{r , include = FALSE, cache = TRUE}
+
+set.seed(1)
+results_noutliers <- data.frame()
+# y2 <- matrix(nrow = 1000, ncol=180)
+for(i in 1:1000){
+  tryCatch({
+y <- arima.sim(model = list(ar = 0.7, ma = -0.4), n = 180)
+
+df2<- tso(y, cval = 3.5)$outliers
+df2$iteration <- i
+results_noutliers <- bind_rows(results_noutliers, df2)
+# print(i)
+  } , error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
+}
+
+# (a <- nrow(results_noutliers) / (180*1000))
+# 
+# a * 20000
+
+results_AO <- data.frame()
+# y2 <- matrix(nrow = 1000, ncol=180)
+for(i in 1:1000){
+  tryCatch({
+    y <- arima.sim(model = list(ar = 0.7, ma = -0.4), n = 180)
+    y[150] <- 4
+
+    df2<- tso(y, cval = 3.5)$outliers
+    df2$iteration <- i
+    results_AO <- bind_rows(results_AO, df2)
+    # print(i)
+  } , error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
+}
+```
+In our anomaly-less time series, we detect `r nrow(results_noutliers)` anomalies on the 180,000 candidate time steps for a false-positive error rate of `r prettyNum(nrow(results_noutliers) / 180000*100, digits = 3)`%.
+In our time series containing a single anomaly, we detect `r nrow(results_AO[which(results_AO$time == 150),])` anomalies out of the 1000 candidate time series, for a false-negative error rate of `r (1-(nrow(results_AO[which(results_AO$time == 150),]) / 1000)) * 100`%.
+This provides a statistical power of `r nrow(results_AO[which(results_AO$time == 150),]) / 1000`.
+These results give us confidence that the anomalies we detect in this paper are true anomalies.
 
@@
 The countefactural estimate with the outlier removed would simply have kept the currency exchange at 1.2 ($\hat{y}$).
-\label{explanationfigure}](./manuscript_files/figure-latex/Figure1.png)
+\label{explanationfigure}](../MainDocument/MainDocument_files/figure-latex/Figure1.png)
 
-where $\pi(B)=\sum_{i=o}^{inf} \pi_iB^i$.
-The identification of outliers then involves a three step process.
+ <!-- The identification of outliers then involves a three step process.
 First, (1) the algorithm identifies all potential outliers, $t_j$ and $L_j (B)$.
@@
 Based on Chen and Liu’s [-@chen1993joint] recommendation, we set a critical value of 3.5 which generally minimizes the possibility of Type I errors or false positive outliers.
+-->
 
@@
 
-We report outliers using a simple t-statistic and report the size of the outlier, or its impact on the time series, by subtracting $\hat{y}$, the adjusted time series with outliers removed, from $y$, the original, unadjusted time series.
+<!-- We report outliers using a simple t-statistic and report the size of the outlier, or its impact on the time series, by subtracting $\hat{y}$, the adjusted time series with outliers removed, from $y$, the original, unadjusted time series.
+-->
 
+## Toy example
 **\autoref{explanationfigure}** shows a toy example for anomaly detection in a time series using the classic example of the exchange rate between the Euro and the Swiss Franc (CHF).
@@
 A full listing of these anomalies can be found in the **Supplementary Materials**.
-We begin by highlighting significant fertility and mortality anomaloies across all three types of outliers (Additive Outliers, Level Shift Outliers, and Temporary Change outliers) with plausible explanations.
+We begin by highlighting significant fertility and mortality anomalies across all three types of outliers (Additive Outliers, Level Shift Outliers, and Temporary Change outliers) with plausible explanations.
 We then highlight two strong anomalies that bely explanation.
@@
 The migration associated with Hurricane Katrina has garnered most of the attention of social scientists [@fussellRecoveryMigrationCity2014;@horiDisplacementDynamicsSouthern2009], but the hurricane had a very clear impact on fertility behaviors too, creating a mini "baby bust" in Louisiana with the departure of many people in their childbearing years.
-We estimate `r la_averbirths` fewer births in Louisiana compared to the counterfactual, likely attributable to the Hurricane.
+We estimate `r la_averbirths` fewer births in Louisiana compared to the counter-factual, likely attributable to the Hurricane.
 
@@
 Here we detect a strong (*t*= `r conn_tstat` for `r conn_births` births/month) level shift toward lower fertility beginning in August 2009 (2009:08) that continues through the end of the period (**\autoref{fig:fertla}b**).
-This LS toward lower fertility reduced the number of births by `r conn_tot` or `r conn_per`) lower than the counterfactual time series.
+This LS toward lower fertility reduced the number of births by `r conn_tot` or `r conn_per`) lower than the counter-factual time series.
 The stock market crash of 2008, where the Dow Jones Industrial Average had the largest single-day loss up to that point, occurred just 11 months before we detect a shift toward lower fertility.
 It is possible that the trend toward lower fertility in August 2009 and the stock market crash in September 2008 are linked, although, as stated previously, we are speculating only to illustrate how causal inference could work in an abductive research agenda.
-A potentially interesting future line of research would be to link fertility anomalies to the kinds of economic shocks that may lead to outmigration of those in their childbearing years.
+A potentially interesting future line of research would be to link fertility anomalies to the kinds of economic shocks that may lead to out-migration of those in their childbearing years.
 
@@
 This is contrast to **\autoref{fig:ferthawaii}b**, where we identify a strange mortality reduction in Ohio (*t*-stat: `r oh_tstat`).
-This LS is more than `r oh_deaths` deaths per month less than the counterfactual time series, suggesting had nearly `r oh_tot` fewer deaths since February 2015 than expected.
+This LS is more than `r oh_deaths` deaths per month less than the counter-factual time series, suggesting had nearly `r oh_tot` fewer deaths since February 2015 than expected.
 This is the single-largest LS among all states.
@@
 
-```{r TrueAnomalies, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Hawaii state fertility (a) and Ohio state mortality (b).} Here we detect one outlier, an additive outlier (AO) in May 2014 in Hawaii (a), representing a large, unexplainable 14% increase in expected births in that month.
+```{r TrueAnomalies, echo= FALSE, message = FALSE, warning = FALSE, fig.height = 5, fig.cap= paste("\\textbf{Anomaly detection in Hawaii state fertility (a) and Ohio state mortality (b).} Here we detect one outlier, an additive outlier (AO) in May 2014 in Hawaii (a), representing a large, unexplainable 14% increase in expected births in that month.
 We also detect a strong level shift (LS) reduction (b), representing a large, unexplainable reduction of nearly 40,000 deaths in Ohio.
@@
 This algorithm is one of many causal inference approaches that are freely and commercially available (see @rcausalimpact).
-In our initial work, we experimented with other approaches, such as Google’s causalimpact algorithm, which uncovered the same patterns we briefly discuss in this paper.
+In our initial work, we experimented with other approaches, such as Google’s CausalImpact algorithm, which uncovered the same patterns we briefly discuss in this paper.
 By making more use of causal inference techniques and abductive modeling approaches, we argue that social scientists will be able to better understand how events or policy implementation can impact important outcomes.

### 2020-08-03 MainDocument/MainDocument.tex
Updated.

--- 2020-02-24 MANUSCRIPT/manuscript.tex
@@
 {
-  \title{\bf Causal Inference in Population Trends: Searching for
-Demographic Anomalies in Big Data}
+  \title{\bf Causal Inference in Population Trends: Searching for Demographic
+Anomalies in Big Data}
 
@@
   \begin{center}
-    {\LARGE\bf Causal Inference in Population Trends: Searching for
-Demographic Anomalies in Big Data}
+    {\LARGE\bf Causal Inference in Population Trends: Searching for Demographic
+Anomalies in Big Data}
   \end{center}
@@
 \noindent%
-{\it Keywords:} mortality, fertility, causal inference, abductive
-reasoning, big data
+{\it Keywords:} mortality, fertility, causal inference, abductive reasoning, big data
 \vfill
@@
 
-The search for causality is one of the primary motiviations behind
-social science \citep{smith2011commentary, pearl2009causal}. Causal
-Inference, the identification of causal mechanisms, attempts to uncover
-the underlying mechanisms that result in changes in a phenomenon and has
-a long history as an approach to inquiry in the social sciences
+The search for causality is one of the primary motivations behind social
+science \citep{smith2011commentary, pearl2009causal}. Causal Inference,
+the identification of causal mechanisms, attempts to uncover the
+underlying mechanisms that result in changes in a phenomenon and has a
+long history as an approach to inquiry in the social sciences
 \citep{grimmer2015ppsp}. Often, the identification of a casual mechanism
@@
 Unfortunately, both RCTs and natural experiments have drawbacks. RCTs
-are often prehibitively expensive for many scientists, precluding their
+are often prohibitively expensive for many scientists, precluding their
 widespread adoption in the social sciences \citep{west2008ajph}. RCTs
 are also sometimes impossible to deploy in the ``real world''. For
-instance, an RCT to study country-level austeriy measures would be
+instance, an RCT to study country-level austerity measures would be
 difficult, if not impossible, to implement. Natural experiments are a
@@
 identify anomalous demographic behavior. In essence, we identify
-\emph{effects without knowledge of the cause}. We ask one fundamanetal
+\emph{effects without knowledge of the cause}. We ask one fundamental
 question regarding demographic anomalies: What are the hidden baby
@@
 \citep{rcore} via the tsoutliers package \citep{tsoutliers2019}. This
-algorithm iteratively uses ARIMA models to 1) produce a counter-factual
-time series to initially detect an outlier or anomaly, and 2) refit the
-ARIMA with the outliers removed. Here we briefly summarize and describe
-the method.
+algorithm iteratively uses ARIMA models to 1) identify potential
+outliers or anomalies and 2) refit the ARIMA with the outliers removed
+to produce a counter-factual time series. Here we briefly summarize and
+describe the method.
 
@@
 ARIMA models. If a series of values, \(y_t^*\), is subject to \(m\)
-interventions or outliers at time points \(t_1,t_2,…,t_m\), then
-\(y_t^*\) can be defined as
+interventions or outliers at time points \(t_1,t_2,…,t_m\) with weights
+\(\omega\) then \(y_t^*\) can be defined as
 
-\[y_t^* = \sum_{j=1}^{m} \omega_jL_j(B)I_t(t_j) + \frac{\theta(B)}{\phi(B)\alpha(B)}\alpha_t\]
+\begin{equation}
+\label{eq:arima}
+y_t^* = \sum_{j=1}^{m} \omega_jL_j(B)I_t(t_j) + \frac{\theta(B)}{\phi(B)\alpha(B)}\alpha_t 
+\end{equation}
 
 Where \(I_t(t_j)\) is an indicator variable with a value of 1 at
-observation \(t_j\) and where the \(j\)th outlier arises, \(\phi(B)\) is
-an autoregressive polynomial with all roots outside the unit circle,
+observation \(t_j\) and where the \(j\)th outlier arises,\\
+\(\phi(B)\) is an autoregressive polynomial with all roots outside the
+unit circle,\\
 \(\theta(B)\) is a moving average polynomial with all roots outside the
-unit circle, and \(\alpha(B)\) is an autoregressive polynomial with all
-roots on the unit circle.
+unit circle,\\
+and \(\alpha(B)\) is an autoregressive polynomial with all roots on the
+unit circle.
 
-We examine three types of outliers at time point \(t_m\): 1) additive
-outliers (AO), defined as \(L_j(B)=1\); 2) level shift outliers (LS),
-defined as \(L_j(B) = 1/(1-B)\); and 2) temporary change outliers (TC),
-defined as \(L_j(B) = 1/(1-\delta B)\).
+We examine three types of outliers at time point \(t_m\):\\
+1. additive outliers (AO), defined as \(L_j(B)=1\);\\
+2. level shift outliers (LS), defined as \(L_j(B) = 1/(1-B)\); and\\
+3. temporary change outliers (TC), defined as
+\(L_j(B) = 1/(1-\delta B)\) where \(\delta\) is equal to 0.7.
 
@@
 
-An outlier is detected using a regression equation
+An outlier is detected with the estimated residuals using a regression
+equation
 
-\[ \pi(B)y_t^* \equiv \hat{e} = \sum_{j=1}^m \omega_j \pi(B)L_j(B)I_t(t_j) + \alpha_t \]
+\begin{equation}
+\label{eq:residuals}
+\pi(B)y_t^* \equiv \hat{e} = \sum_{j=1}^m \omega_j \pi(B)L_j(B)I_t(t_j) + \alpha_t
+\end{equation} where \(\pi(B)=\sum_{i=o}^{inf} \pi_iB^i\).
+
+Equations \ref{eq:arima} and \ref{eq:residuals} allow for an automatic
+detection of anomalies iterated over a two-stage process.
+
+In stage 1, anomalies are located. First, an ARIMA model is fit to the
+time series using the \texttt{forecast} package in R
+\citep{Rforecast, hymdman2008} where the best performing ARIMA model is
+selected based on the Bayesian information criterion (BIC). Next, the
+residuals from the forecast are checked for their significance using
+equation \ref{eq:residuals} where only anomalies above a critical
+\emph{t}-static are considered ``true'' anomalies (\(|\tau| \geq 3.5\)).
+We choose this critical value based on Chen and Liu's
+\citeyearpar{chen1993joint} recommendation to minimize Type I or
+false-positive anomalies. Finally, two additional rules are implemented:
+If multiple anomalies are detected at the same time point, only the most
+significant anomaly is selected and if anomalies of the same type at
+consecutive time periods are detected, only the anomaly with highest
+\emph{t}-statistic is selected.
+
+In stage 2, anomalies are removed from the time series and a new ARIMA
+model is chosen and fit. The selection of the initial ARIMA model could
+have been affected by the presence of the anomalies, making some
+anomalies spuriously identified. To correct for this, a new ARIMA model
+is fit accounting for additional regression effects in equation
+\ref{eq:arima} from the list of candidate anomalies identified in stage
+1, effectively removing the anomalies from the time series. Each anomaly
+is then reassessed under the new model and those anomalies that are no
+longer significant are removed.
+
+These two stages are then iterated until no additional anomalies are
+detected.
+
+\hypertarget{robustness-check}{%
+\subsection{Robustness Check}\label{robustness-check}}
+
+To determine the false-positive (Type I) and false-negative (Type II)
+error rates under a critical value of 3.5, we ran two simulations. For
+the false-positive error we simulated 1000 time series with 180 time
+steps without introducing any anomalies and ran the \texttt{tsoutliers}
+algorithm on them. Any anomalies we detect in these time series are
+spurious anomalies. For the false-negative error, we simulated 1000 time
+series with 180 time steps but introduced a single anomaly of the same
+magnitude at the same time point in each time series. If the
+\texttt{tsoutliers} algorithm fails to detect the anomaly, it would be
+considered a false-negative error.
+
+In our anomaly-less time series, we detect 208 anomalies on the 180,000
+candidate time steps for a false-positive error rate of 0.116\%. In our
+time series containing a single anomaly, we detect 817 anomalies out of
+the 1000 candidate time series, for a false-negative error rate of
+18.3\%. This provides a statistical power of 0.817. These results give
+us confidence that the anomalies we detect in this paper are true
+anomalies.
 
@@
 \centering
-\includegraphics{./manuscript_files/figure-latex/Figure1.png}
+\includegraphics{../MainDocument/MainDocument_files/figure-latex/Figure1.png}
 \caption{\textbf{``Toy'' example of anomaly detection using the exchange
@@
 
-where \(\pi(B)=\sum_{i=o}^{inf} \pi_iB^i\). The identification of
-outliers then involves a three step process. First, (1) the algorithm
-identifies all potential outliers, \(t_j\) and \(L_j (B)\). Next, (2) we
-compute joint estimates of model parameters and outlier effects to
-identify potentially spurious outliers. Finally, (3) the outliers and
-effects are re-estimated without spurious outliers. Of importance here
-is step 2, which relies on some critical value above which an outlier at
-time point \(m\) is considered spurious. Based on Chen and Liu's
-\citeyearpar{chen1993joint} recommendation, we set a critical value of
-3.5 which generally minimizes the possibility of Type I errors or false
-positive outliers.
-
-We report outliers using a simple t-statistic and report the size of the
-outlier, or its impact on the time series, by subtracting \(\hat{y}\),
-the adjusted time series with outliers removed, from \(y\), the
-original, unadjusted time series.
+\hypertarget{toy-example}{%
+\subsection{Toy example}\label{toy-example}}
 
@@
 in the \textbf{Supplementary Materials}. We begin by highlighting
-significant fertility and mortality anomaloies across all three types of
+significant fertility and mortality anomalies across all three types of
 outliers (Additive Outliers, Level Shift Outliers, and Temporary Change
@@
 people in their childbearing years. We estimate 4,411 fewer births in
-Louisiana compared to the counterfactual, likely attributable to the
+Louisiana compared to the counter-factual, likely attributable to the
 Hurricane.
@@
 (\textbf{\autoref{fig:fertla}b}). This LS toward lower fertility reduced
-the number of births by 28,752 or 9\%) lower than the counterfactual
+the number of births by 28,752 or 9\%) lower than the counter-factual
 time series. The stock market crash of 2008, where the Dow Jones
@@
 future line of research would be to link fertility anomalies to the
-kinds of economic shocks that may lead to outmigration of those in their
-childbearing years.
+kinds of economic shocks that may lead to out-migration of those in
+their childbearing years.
 
@@
 \centering
-\includegraphics{manuscript_files/figure-latex/FertilityAnomalies-1.pdf}
+\includegraphics{MainDocument_files/figure-latex/FertilityAnomalies-1.pdf}
 \caption{\textbf{Anomaly detection in Louisiana (a) and Connecticut (b) state fertility, 2003-2018.}
@@
 \centering
-\includegraphics{manuscript_files/figure-latex/MortalityAnomalies-1.pdf}
+\includegraphics{MainDocument_files/figure-latex/MortalityAnomalies-1.pdf}
 \caption{\textbf{Anomaly Detection for New York (a) and New Hampshire (b) state mortality, 1999-2016.}
@@
 identify a strange mortality reduction in Ohio (\emph{t}-stat: -4.27).
-This LS is more than 1,157 deaths per month less than the counterfactual
-time series, suggesting had nearly 40,000 fewer deaths since February
-2015 than expected. This is the single-largest LS among all states. We
-could not identify the potential policies Ohio might have put into place
-to provide such a strong mortality protection.
+This LS is more than 1,157 deaths per month less than the
+counter-factual time series, suggesting had nearly 40,000 fewer deaths
+since February 2015 than expected. This is the single-largest LS among
+all states. We could not identify the potential policies Ohio might have
+put into place to provide such a strong mortality protection.
 
@@
 \centering
-\includegraphics{manuscript_files/figure-latex/TrueAnomalies-1.pdf}
+\includegraphics{MainDocument_files/figure-latex/TrueAnomalies-1.pdf}
 \caption{\textbf{Anomaly detection in Hawaii state fertility (a) and Ohio state mortality (b).}
@@
 
-\caption{\label{tab:unnamed-chunk-6}\textbf{Summary by Demographic Component.}  Here we can see there 22 Fertility and 156 Mortality anomalies among the state-level time series, totalling more than 200k anomalous births and 600k anomalous deaths.
+\caption{\label{tab:unnamed-chunk-7}\textbf{Summary by Demographic Component.}  Here we can see there 22 Fertility and 156 Mortality anomalies among the state-level time series, totalling more than 200k anomalous births and 600k anomalous deaths.
 \label{sumtable}}
@@
 \centering
-\includegraphics{manuscript_files/figure-latex/AnomalyMap-1.pdf}
+\includegraphics{MainDocument_files/figure-latex/AnomalyMap-1.pdf}
 \caption{\textbf{Summary of Anomalies by type, demographic component, and State.}
@@
 commercially available (see \citet{rcausalimpact}). In our initial work,
-we experimented with other approaches, such as Google's causalimpact
+we experimented with other approaches, such as Google's CausalImpact
 algorithm, which uncovered the same patterns we briefly discuss in this
@@
 \bibliographystyle{agsm}
-\bibliography{mybibfile}
+\bibliography{../MainDocument/mybibfile.bib}
 

### 2020-08-03 R/SANDBOX/PR_mort.R

@@ -14,7 +14,7 @@ z <- ts(PRdeaths$Deaths,
 # Generating the TS outliers using just the `Deaths`
 outlier.morttots <- tsoutliers::tso(ts(PRdeaths$Deaths, 
                                        start = year(min(PRdeaths$Month3)), 
-                                       end = c(year(max(PRdeaths$Month3)),12), frequency = 12),  maxit.iloop=10)
+                                       end = c(year(max(PRdeaths$Month3)),12), frequency = 12),  cval = 3.5, maxit.iloop=10)
 
 # Generating the TS outliers using the `Deaths` and the regressor variable `Population_Estimate`
 # outlier.mortots.popest <- tsoutliers::tso(ts(PRdeaths$Deaths), xreg = PRdeaths$Population_Estimate,  maxit.iloop=10)
@@ -24,6 +24,6 @@ plot(outlier.morttots)
 outlier.morttots
 # Summing up the difference in the Mortality using only the `Deaths`
 sum(outlier.morttots$y[225:240] - outlier.morttots$yadj[225:240])
-
+sum(outlier.morttots$y[72:87] - outlier.morttots$yadj[72:87])
 # Summing up the difference using the `Deaths` and the `Population_Estimate` regressor
-sum(outlier.mortots.popest$y - outlier.mortots.popest$yadj)
+sum(outlier.morttots$y - outlier.morttots$yadj)

### 2020-08-03 R/SANDBOX/clusteringanalysis.R

@@ -1,0 +1,133 @@
+
+library(tsoutliers)
+set.seed(1)
+
+z <- mortality_anomalies %>%
+  group_by(ind) %>%
+  dplyr::summarise(number = n(),
+                   targets = sum(coefhat))
+
+z2 <- fertility_anomalies %>%
+  group_by(ind) %>%
+  dplyr::summarise(number = n(),
+                   targets = sum(coefhat))
+
+
+mort2 <- fertility_anomalies %>%
+  mutate(direction = if_else(coefhat <0, "dip", "spike")) %>%
+  group_by(time, direction) %>%
+  dplyr::summarise(number = n(),
+                   targets = sum(coefhat))
+
+ggplot(data = mort2, aes(x = time, y = number, fill = direction)) +
+  geom_bar(stat = "identity") +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 90)) +
+  labs(title = "# of States exhibiting a mortality anomaly")
+
+ggplot(data = fertility_anomalies, aes(x = ind)) +
+  geom_bar(stat = "count") +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 90)) +
+  labs(title = "# of States exhibiting a fertility anomaly")
+
+mort <- mortality_anomalies %>%
+  mutate(Clusters = case_when(
+    ind %in% c(13,14,15) ~ "2000:January",
+    ind %in% c(60) ~ "2003:December",
+    ind %in% c(73,74,75) ~ "2005:January",
+    ind %in% c(109,110,111) ~ "2008:January",
+    ind %in% c(169) ~ "2013:January",
+    ind %in% c(191,192,193,194) ~ "2014:November"
+  )) %>%
+  na.omit %>%
+  mutate(GEOID = substr(loc,2,3),
+         number = "a") %>%
+  dplyr::select(GEOID, Clusters, number, coefhat, code = STATE_ABBREV) %>%
+  distinct() %>%
+  group_by(GEOID, Clusters, number, code) %>%
+  dplyr::summarise(coefhat = sum(coefhat))
+  # pivot_wider(names_from = Clusters, values_from = number )
+  # filter(Clusters == "January 2000")
+
+data(state_laea) 
+
+states <- left_join(state_laea, mort) %>%
+  na.omit
+
+states2 <- state_laea
+
+ggplot(data=states) +
+  # geom_bar(mapping=aes(x=number, fill=number))
+  geom_sf(data = states2, color = NA, ) +
[…]

### 2020-08-03 R/SANDBOX/exportfigures.R

@@ -1,0 +1,85 @@
+
+dat_fert <-
+  ###FERTILITY
+  bind_rows(read_tsv("./R/DATA-RAW/Fertility/Natality, 2003-2006,states.txt"),
+            read_tsv("./R/DATA-RAW/Fertility/Natality, 2007-2017,states.txt"))%>%
+  mutate(Month2 = str_pad(`Month Code`,2 , pad = "0"),
+         Month3 = as.yearmon(paste0(Year,"-", Month2)),
+         location = paste0("X", `State Code`)
+  ) %>%
+  dplyr::select(Year = Month3, location, Target = Births) %>%
+  na.omit %>%
+  arrange(location, Year)
+
+dat_mort <- 
+  ###MORTALITY 
+  read_tsv("./R/DATA-RAW/Mortality/Underlying Cause of Death, 1999-2017,states.txt") %>%
+  mutate(Month2 = substr(`Month Code`,6,7),
+         Month3 = as.yearmon(paste0(Year,"-", Month2)),
+         # location = paste0("X", `County Code`)
+         location = paste0("X", `State Code`)
+  ) %>%
+  dplyr::select(Year = Month3, location, Target = Deaths) %>%
+  na.omit %>%
+  arrange(location, Year)
+
+refstates <- read_csv("./R/DATA-RAW/ref_states.csv") %>%
+  mutate(loc = paste0("X", str_pad(loc, 2, pad = "0")),
+         location = loc)
+  
+
+set.seed(1)
+
+fertility_anomalies <- read_rds( "./R/DATA-PROCESSED/fertility_anomalies")%>%
+  left_join(., refstates)
+df_sums2 <- read_rds("./R/DATA-PROCESSED/fertility_anomalies_sums") %>%
+  left_join(., refstates)
+
+
+lista <- unique(fertility_anomalies$location)
+pdf(file = "fertility_anomalies.pdf",
+    width=11, height=8.5)
+
+for(this.state in lista){
+  dat2 <- filter(dat_fert, location == this.state)
+  d3 <- filter(df_sums2, loc == this.state)
+  dat3 <- ts(dat2$Target, start = year(min(dat2$Year)), end = c(year(max(dat2$Year)),12), frequency = 12)
+  outlier.county <- tsoutliers::tso(dat3,types = c("AO","LS","TC"), cval = 3.5, maxit.iloop=10)
+  a <-  function() {
+    plot(outlier.county)
[…]

### 2020-08-03 R/SCRIPTS/000-Libraries.R

@@ -65,7 +65,9 @@ pkgs <- c(
   "lubridate",
   "tsoutliers",
   "numform",
-  "statebins"
+  "statebins",
+  "flipTime",
+  "gridGraphics"
 )
 
 # Install missing packages

### 2020-09-01 MainDocument/MainDocument.Rmd
Minor Edits.

@@
 abstract: "The proliferation of big data, wider access to advanced computing platforms, and the development of powerful statistical algorithms can uncover hidden anomalies in social data, previously dismissed as noise.
-Here, we combine causal inference techniques and abductive reasoning to identify fertility and mortality anomalies on twenty years of demographic data in the United States.
+Here, we combine causal inference techniques and abductive reasoning to identify fertility and mortality anomalies on twenty years of complete demographic data in the United States.
 We uncover real, “hidden” baby booms/busts and mortality spikes/dips, distinguishable from regular trend variations.
@@
 
+## Replication Files
+All data and code necessary to reproduce the reported results are licensed under the CC-BY-4.0 license and are publicly available in a replication repository located at https://osf.io/fp52x/?view_only=754d9a72a2ea4f6b8e0c193dc9a590d1.
+The analyses were performed in *R* [@rcite].
+
 # Results

### 2020-09-01 MainDocument/MainDocument.tex
Minor Edits.

@@
 Here, we combine causal inference techniques and abductive reasoning to
-identify fertility and mortality anomalies on twenty years of
+identify fertility and mortality anomalies on twenty years of complete
 demographic data in the United States. We uncover real, ``hidden'' baby
@@
 series for a total of 20,808 state-months of data.
+
+\hypertarget{replication-files}{%
+\subsection{Replication Files}\label{replication-files}}
+
+All data and code necessary to reproduce the reported results are
+licensed under the CC-BY-4.0 license and are publicly available in a
+replication repository located at
+\url{https://osf.io/fp52x/?view_only=754d9a72a2ea4f6b8e0c193dc9a590d1}.
+The analyses were performed in \emph{R} \citep{rcite}.
 

### 2020-09-11 MainDocument/MainDocument.Rmd
New small section on Abductive reasoning.

@@
 From explanations of phenomena that may have previously gone unnoticed, social scientists may be able to better forecast populations and provide policy solutions for impending problems such as climate change.
+
+# Abductive Reasoning and Causal Inference
+
+Scientific inquiry falls under one of three modes of logical reasoning: deduction, induction, and abduction.
+Deductive reasoning starts with a theory and makes a prediction of what the observations should be if that theory is correct.
+The classic example of deductive reasoning essentially goes (a) "all men are mortal", (b) "Socrates is a man", (c) "Therefore, Socrates is mortal." Deductive reasoning has a long history in the sciences and most hypothesis testing in the social sciences follows this logical reasoning.
+For example, John Snow used deductive reasoning in to stop the 1854 London cholera outbreak.
+He hypothesized that cholera was waterborne and deduced that a bad water pump was responsible for the outbreak.
+
+Inductive reasoning is the opposite of deductive reasoning.
+Rather than starting with a theory and making a prediction, induction starts with an observation and makes a theory based on the likeliest reason.
+For example, the Hispanic Mortality Paradox first appeared in the scientific literature in the mid-1980s [@markides1986health] as an observation that hispanics in the United States had, paradoxically, better health outcomes than expected given their socioeconomic status.
+Based on this observation, numerous theories arose to explain this paradox.
+
+Both deductive and inductive reasoning are well established in science but are not without their shortcomings.
+In both approaches, it is impossible, or at least imporbable, for the premise to be true and the conclusion false.
+Furthermore, if a hypothesis is rejected under a deductive or inductive approach, there is no alternative explanation -- there is either the presence or absence of support.
+
+Abductive reasoning is a third, less well-known way that reasons the best explanation given the observations [@walton2014abductive].
+It is most concerned with generating and evaluating explanatory hypotheses [@Crowder2017] by looking for a pattern and suggesting a hypothesis [@fann2012peirce;@pierce1878deduction].
+Unlike the more rigid deductive and inductive reasoning, abductive reasoning allows the scientist to move between deduction and induction and fits well during exploratory data analysis [@yu1994abduction;@haig2015commentary;@tukey1977exploratory].
+During exploratory data analysis, abductive reasoning helps find a pattern and suggest a hypothesis, deduction helps refine and build a testable hypothesis, and induction is the empirical support [@yu1994abduction].
+
+Abductive reasoning and exploratory data analysis are best employed in a data rich environment -- the very data environment social scientists and demographers find themselves in the present moment.
+Additionally, in a data rich environment, a hypothesis is a hidden liability [@yanai_hypothesis_2020]because it focuses our attention on only that which we seek^[Yanai and Lercher [-@yanai_hypothesis_2020] ask if focusing on a specific hypothesis prevents new discovery?
+They created a simple dataset that if plotted would reveal an image of a gorilla waving.
+Students instructed to find support for specific hypotheses were less likely to discover the gorilla compared to those who were simply asked "What do you conclude from the dataset?"].
+Focusing on specific hypotheses often prevents us from finding other interesting phenomena.
+Afterall, it was the publication of the hispanic paradox itself that has generated more than thirty years of devoted scholarship.
+When paired with causal inference techniques, abductive reasoning provides an empirical platform for the discovery of potentially interesting social phenomena; it provides an important tool for sifting large amounts of data to find interesting results in the absence of hypothesis testing.
 

### 2020-09-11 MainDocument/MainDocument.tex
New small section on Abductive reasoning.

@@
 problems such as climate change.
+
+\hypertarget{abductive-reasoning-and-causal-inference}{%
+\section{Abductive Reasoning and Causal
+Inference}\label{abductive-reasoning-and-causal-inference}}
+
+Scientific inquiry falls under one of three modes of logical reasoning:
+deduction, induction, and abduction. Deductive reasoning starts with a
+theory and makes a prediction of what the observations should be if that
+theory is correct. The classic example of deductive reasoning
+essentially goes (a) ``all men are mortal'', (b) ``Socrates is a man'',
+(c) ``Therefore, Socrates is mortal.'' Deductive reasoning has a long
+history in the sciences and most hypothesis testing in the social
+sciences follows this logical reasoning. For example, John Snow used
+deductive reasoning in to stop the 1854 London cholera outbreak. He
+hypothesized that cholera was waterborne and deduced that a bad water
+pump was responsible for the outbreak.
+
+Inductive reasoning is the opposite of deductive reasoning. Rather than
+starting with a theory and making a prediction, induction starts with an
+observation and makes a theory based on the likeliest reason. For
+example, the Hispanic Mortality Paradox first appeared in the scientific
+literature in the mid-1980s \citep{markides1986health} as an observation
+that hispanics in the United States had, paradoxically, better health
+outcomes than expected given their socioeconomic status. Based on this
+observation, numerous theories arose to explain this paradox.
+
+Both deductive and inductive reasoning are well established in science
+but are not without their shortcomings. In both approaches, it is
+impossible, or at least imporbable, for the premise to be true and the
+conclusion false. Furthermore, if a hypothesis is rejected under a
+deductive or inductive approach, there is no alternative explanation --
+there is either the presence or absence of support.
+
+Abductive reasoning is a third, less well-known way that reasons the
+best explanation given the observations \citep{walton2014abductive}. It
+is most concerned with generating and evaluating explanatory hypotheses
+\citep{Crowder2017} by looking for a pattern and suggesting a hypothesis
+\citep{fann2012peirce, pierce1878deduction}. Unlike the more rigid
+deductive and inductive reasoning, abductive reasoning allows the
+scientist to move between deduction and induction and fits well during
+exploratory data analysis
+\citep{yu1994abduction, haig2015commentary, tukey1977exploratory}.
+During exploratory data analysis, abductive reasoning helps find a
+pattern and suggest a hypothesis, deduction helps refine and build a
+testable hypothesis, and induction is the empirical support
+\citep{yu1994abduction}.
+
+Abductive reasoning and exploratory data analysis are best employed in a
+data rich environment -- the very data environment social scientists and
+demographers find themselves in the present moment. Additionally, in a
+data rich environment, a hypothesis is a hidden liability
+\citep{yanai_hypothesis_2020}because it focuses our attention on only
+that which we seek\footnote{Yanai and Lercher
+  \citeyearpar{yanai_hypothesis_2020} ask if focusing on a specific
+  hypothesis prevents new discovery? They created a simple dataset that
+  if plotted would reveal an image of a gorilla waving. Students
+  instructed to find support for specific hypotheses were less likely to
+  discover the gorilla compared to those who were simply asked ``What do
+  you conclude from the dataset?''}. Focusing on specific hypotheses
+often prevents us from finding other interesting phenomena. Afterall, it
+was the publication of the hispanic paradox itself that has generated
+more than thirty years of devoted scholarship. When paired with causal
+inference techniques, abductive reasoning provides an empirical platform
+for the discovery of potentially interesting social phenomena; it
+provides an important tool for sifting large amounts of data to find
+interesting results in the absence of hypothesis testing.
 

### 2020-09-14 MainDocument/MainDocument.Rmd
Update MainDocument.Rmd

@@
 
-# abstract: "The proliferation of big data, wider access to advanced computing platforms, and the development of powerful statistical algorithms can uncover hidden anomalies in demographic data that have previously been dismissed as noise.
+# abstract: "The proliferation of big data, wider access to advanced computing platforms, and the development of powerful statistical algorithms allows us to uncover hidden anomalies in demographic data that have previously been dismissed as noise.
 Here, we demonstrate how social scientists can use causal inference techniques and abductive reasoning to identify fertility and mortality anomalies in state-level data trends by investigating demographic time series data since 1999.
@@
 
-The search for causality is one of the primary motivations behind social science [@smith2011commentary; @pearl2009causal].
-Causal Inference, the identification of causal mechanisms, attempts to uncover the underlying mechanisms that result in changes in a phenomenon and has a long history as an approach to inquiry in the social sciences [@grimmer2015ppsp].
+The search for causality is one of the primary motivations of social scientists [@smith2011commentary; @pearl2009causal].
+Causal inference, the identification of causal mechanisms, attempts to uncover the underlying mechanisms that result in changes in a phenomenon and has a long history as an approach to inquiry in the social sciences [@grimmer2015ppsp].
 Often, the identification of a casual mechanism is accomplished via a randomized control trial (RCT) or expert knowledge applied to a natural experiment [@salganik2019bit].
@@
 RCTs are often prohibitively expensive for many scientists, precluding their widespread adoption in the social sciences [@west2008ajph].
-RCTs are also sometimes impossible to deploy in the "real world".
-For instance, an RCT to study country-level austerity measures would be difficult, if not impossible, to implement.
-Natural experiments are a cheaper alternative, but provide less certainty about causation, fewer opportunities to replicate work, and require expert knowledge of the natural experiment to be properly employed [@pearl2018book].
+RCTs are also sometimes impossible to deploy in the "real world." For instance, an RCT to study country-level austerity measures would be difficult, if not impossible, to implement.
+RCTs to study fertility would be immoral.
+Natural experiments are a cheaper alternative but provide less certainty about causation, fewer opportunities to replicate work, and require expert knowledge of the natural experiment to be properly employed [@pearl2018book].
 
@@
 
-Additionally, both RCTs and natural experiments follow traditional hypothesis testing, the deductive scientific paradigm where a question is first posed and scientists generate or find data to answer the question.
-Such approaches, while extensively utilized and time-tested, are likely to miss important phenomena that might go unnoticed or suffer from extensive p-hacking [@head2015extent;@nuzzo2014scientific;@ruggles2014big].
-Abductive reasoning, sometimes described as “inferring cause from effect” [@Crowder2017], moves the scientific process from the inductive to the deductive, and sometimes back and forth, to reach conclusions [@bryant2014realm] The physical sciences use abductive reasoning more fully compared to the social sciences.
+Additionally, both RCTs and natural experiments are based on traditional hypothesis-based testing, the deductive scientific paradigm where a question is first posed and scientists generate or find data to answer the question.
+Such approaches, while extensively utilized and time-tested, increase the likelihood that a researcher will miss important phenomena that might go unnoticed because it falls outside the hypothesis, or the study could suffer from extensive p-hacking [@head2015extent;@nuzzo2014scientific;@ruggles2014big].
+At the same time, purely inductive data mining can result in false assertions or spurious findings, especially in the hands of those who are not field experts [@yanai_hypothesis_2020].
+Abductive reasoning, sometimes described as “inferring cause from effect” [@Crowder2017], moves the scientific process from the inductive to the deductive, and sometimes back and forth, to reach conclusions [@bryant2014realm] Physical scientists use abductive reasoning more commonly than social scientists.
 The discovery of Neptune in 1846 based on unexplained perturbations of the orbit of Uranus [@popper2005logic], the discovery of galactic migration of planets [@gomes2005n], and the current generation of gravitational wave detectors such as the Laser Interferometer Gravitational-Wave Observatory (LIGO) [@harry2010advanced] demonstrate the power of the abductive approach in the physical sciences.
@@
 The modern proliferation of data, advances in high performance (“super”) computing, and the development of powerful statistical algorithms provide an environment ripe for wedding causal inference with abductive reasoning [@van2016data;@zikopoulos2011], with the potential for researchers to find hidden or understudied social phenomena [@bohon2018demography].
-While the availability of Big Data and high performance computing allows novel exploration of data through causal inference [@bohon2018demography;@rcausalimpact;@shiffrin2016drawing], relatively few studies utilize causal inference techniques in the study of demographic phenomena.
+While the availability of big data and high performance computing allows novel exploration of data through causal inference [@bohon2018demography;@rcausalimpact;@shiffrin2016drawing], relatively few studies utilize causal inference techniques in the study of demographic phenomena.
 However, understanding social phenomena using these advances reveals important insights into society [@angrist1989lifetime;@mas2009peers] and allows us to better monitor population trends [@nobles2019;@torche2015hidden].
@@
 
-In this paper, we combine a modern statistical outlier detection algorithm [@chen1993joint], abductive reasoning, and large demographic data sets on more than 20 years of mortality/fertility to identify anomalous demographic behavior.
+In this paper, we combine a modern statistical outlier detection algorithm [@chen1993joint], abductive reasoning, and large demographic data sets on more than 20 years of mortality and fertility trends to identify anomalous demographic behavior.
 In essence, we identify *effects without knowledge of the cause*.
@@
 Deductive reasoning starts with a theory and makes a prediction of what the observations should be if that theory is correct.
-The classic example of deductive reasoning essentially goes (a) "all men are mortal", (b) "Socrates is a man", (c) "Therefore, Socrates is mortal." Deductive reasoning has a long history in the sciences and most hypothesis testing in the social sciences follows this logical reasoning.
-For example, John Snow used deductive reasoning in to stop the 1854 London cholera outbreak.
+The classic example of deductive reasoning essentially goes (a) "all men are mortal," (b) "Socrates is a man," (c) "therefore, Socrates is mortal." Deductive reasoning has a long history in the sciences and most hypothesis testing in the social sciences follows this logical reasoning.
+For example, John Snow used deductive reasoning to stop the 1854 London cholera outbreak.
 He hypothesized that cholera was waterborne and deduced that a bad water pump was responsible for the outbreak.
 
-Inductive reasoning is the opposite of deductive reasoning.
+Inductive reasoning is directionally opposite of deductive reasoning.
 Rather than starting with a theory and making a prediction, induction starts with an observation and makes a theory based on the likeliest reason.
-For example, the Hispanic Mortality Paradox first appeared in the scientific literature in the mid-1980s [@markides1986health] as an observation that hispanics in the United States had, paradoxically, better health outcomes than expected given their socioeconomic status.
+For example, the Hispanic Mortality Paradox first appeared in the scientific literature in the mid-1980s [@markides1986health] as an observation that Hispanics in the United States had better health outcomes than expected given their socioeconomic status.
 Based on this observation, numerous theories arose to explain this paradox.
@@
 Both deductive and inductive reasoning are well established in science but are not without their shortcomings.
-In both approaches, it is impossible, or at least imporbable, for the premise to be true and the conclusion false.
+In both approaches, it is impossible, or at least improbable, for the premise to be true and the conclusion false.
 Furthermore, if a hypothesis is rejected under a deductive or inductive approach, there is no alternative explanation -- there is either the presence or absence of support.
@@
 Abductive reasoning and exploratory data analysis are best employed in a data rich environment -- the very data environment social scientists and demographers find themselves in the present moment.
-Additionally, in a data rich environment, a hypothesis is a hidden liability [@yanai_hypothesis_2020]because it focuses our attention on only that which we seek^[Yanai and Lercher [-@yanai_hypothesis_2020] ask if focusing on a specific hypothesis prevents new discovery?
-They created a simple dataset that if plotted would reveal an image of a gorilla waving.
+In a data rich environment, a hypothesis is a hidden liability [@yanai_hypothesis_2020] because it focuses our attention on only that which we seek.
+To demonstrate this, ^[Yanai and Lercher [-@yanai_hypothesis_2020] ask if focusing on a specific hypothesis prevents new discovery.
+By way of illustration, they created a simple dataset that if plotted would reveal an image of a gorilla waving.
 Students instructed to find support for specific hypotheses were less likely to discover the gorilla compared to those who were simply asked "What do you conclude from the dataset?"].
 Focusing on specific hypotheses often prevents us from finding other interesting phenomena.
-Afterall, it was the publication of the hispanic paradox itself that has generated more than thirty years of devoted scholarship.
-When paired with causal inference techniques, abductive reasoning provides an empirical platform for the discovery of potentially interesting social phenomena; it provides an important tool for sifting large amounts of data to find interesting results in the absence of hypothesis testing.
+After all, it was the publication of the Hispanic paradox itself that has generated more than thirty years of devoted scholarship.
+At the same time, when paired with causal inference techniques, abductive reasoning provides an empirical platform for the discovery of potentially interesting social phenomena; it provides an important tool for sifting large amounts of data to find interesting results without the blinders of hypothesis testing.
 
@@
 ```
-In our anomaly-less time series, we detect `r nrow(results_noutliers)` anomalies on the 180,000 candidate time steps for a false-positive error rate of `r prettyNum(nrow(results_noutliers) / 180000*100, digits = 3)`%.
+In our sans-anomaly time series, we detect `r nrow(results_noutliers)` anomalies on the 180,000 candidate time steps for a false-positive error rate of `r prettyNum(nrow(results_noutliers) / 180000*100, digits = 3)`%.
 In our time series containing a single anomaly, we detect `r nrow(results_AO[which(results_AO$time == 150),])` anomalies out of the 1000 candidate time series, for a false-negative error rate of `r (1-(nrow(results_AO[which(results_AO$time == 150),]) / 1000)) * 100`%.
@@
 What policies did NY put into place that might have contributed to this considerable mortality reduction?
-What environmental or economic conditions may have changed may have changed?
+What environmental or economic conditions may have changed?
 These are the kinds of questions that arise from our analyses.
@@
 This is the single-largest LS among all states.
-We could not identify the potential policies Ohio might have put into place to provide such a strong mortality protection.
+We could not identify the potential policies Ohio might have put into place or the events that occurred to provide such a strong mortality protection.
 
@@
 Fertility anomalies overwhelmingly tend toward lower fertility, with only a few fertility anomalies yielding more births.
-This is consistent with research findings that link high severity weather events to significantly lower fertility [@Evans2008Hurricanebirth].
+This is consistent with research findings that link severe weather events to significantly lower fertility [@Evans2008Hurricanebirth].
 Conversely, mortality anomalies tend to be more evenly split between positive and negative anomalies, but tend toward more deaths rather than fewer deaths.
@@
 As **\autoref{fig:summap}** shows, three states exhibited neither mortality nor fertility anomalies: Alaska, North Dakota, and South Dakota.
-Another nine states exhibited just a single anomaly (DE, DC, ID, IL, KS, NV, NM, UT, and WY).
+Another nine states exhibited just a single anomaly (Delaware, District of Colombia, Idaho, Illinois, Kansas, Nevada, New Mexico, Utah, and Wyoming).
 Both New York and Massachusetts exhibited the most anomalies with nine.
@@
 Data scientists frequently claim that the big data revolution is a turning point in scientific discovery that will allow us to solve some of the world’s most pressing problems [@grimmer2015ppsp].
-Social scientists are skeptical of such claims, because they better understand the complexities of the social world and know from experience that data alone is not enough to solve social problems [@bohon2018demography;@grimmer2015ppsp].
+Social scientists are skeptical of such claims, because they better understand the complexities of the social world and know from experience that data alone are not enough to solve social problems [@bohon2018demography;@grimmer2015ppsp].
 Nonetheless, the increased availability of data and (more importantly, we argue) the development of advanced techniques for analyzing these data will enable important discovery [@monroe2015no].

### 2021-03-29 MainDocument/MainDocument.Rmd
Updated.

@@
 - name: Mathew E. Hauer
-  thanks:  Thanks y'all!
+  # thanks:  Thanks y'all!
   affiliation: Department of Sociology, Florida State University
@@
 We identify more than 22 and 156 fertility and mortality anomalies, totaling more than 200k and 600k anomalous births and deaths, respectively.
-Notable detectable mortality anomalies include the September 11 2001 terrorist attack in New York, the emergence and acceleration of the opioid epidemic in New Hampshire, and the excess mortality due to Hurricane Maria in Puerto Rico.
+Notable detectable mortality anomalies include the September 11 2001 terrorist attack in New York and the emergence and acceleration of the opioid epidemic in New Hampshire.
 Notable fertility anomalies include the “missing births” in Louisiana after Hurricane Katrina and the reduction in fertility behavior after the September 2008 stock market crash in Connecticut, amongst others.
-Our combined causal inference and abductive reasoning approach can be readily adapted to find other, undiscovered social phenomena or to evaluate the efficacy of important public policies."
+The combined causal inference and abductive reasoning approach can be readily adapted to find other, undiscovered social phenomena or to evaluate the efficacy of important public policies."
 
@@
 
+ countylist = unique(dat_fert$location)
+
+noamom_fert <- setdiff(countylist, unique(anom_fert$location))
+
+noamom_mort <- setdiff(countylist, unique(anom_mort$location))
+
+  results_noutliers <- read_csv("../R/DATA-PROCESSED/robustness_nooutliers.csv")
+  results_AO <-   read_csv("../R/DATA-PROCESSED/robustness_AO.csv")
+
+z2 <- results_noutliers %>%
+  group_by(type) %>%
+  dplyr::summarise(tot = n()) %>%
+  mutate(timesteps = if_else(type == "fertility", 180*1000*33, 228*1000*8)) %>%
+  dplyr::summarise(tot = sum(tot),
+                   timesteps = sum(timesteps)) %>%
+  mutate(FNR = tot/timesteps)
+
 abstract <- rbind(anom_mort, anom_fert) %>%
@@
 The search for causality is one of the primary motivations of social scientists [@smith2011commentary; @pearl2009causal].
-Causal inference, the identification of causal mechanisms, attempts to uncover the underlying mechanisms that result in changes in a phenomenon and has a long history as an approach to inquiry in the social sciences [@grimmer2015ppsp].
+Causal inference, the identification of causal effects, attempts to uncover the underlying mechanisms that result in changes in a phenomenon and has a long history as an approach to inquiry in the social sciences [@grimmer2015ppsp].
 Often, the identification of a casual mechanism is accomplished via a randomized control trial (RCT) or expert knowledge applied to a natural experiment [@salganik2019bit].
@@
 At the same time, purely inductive data mining can result in false assertions or spurious findings, especially in the hands of those who are not field experts [@yanai_hypothesis_2020].
-Abductive reasoning, sometimes described as “inferring cause from effect” [@Crowder2017], moves the scientific process from the inductive to the deductive, and sometimes back and forth, to reach conclusions [@bryant2014realm] Physical scientists use abductive reasoning more commonly than social scientists.
+Abductive reasoning, sometimes described as “inferring cause from effect” [@Crowder2017], moves the scientific process from the inductive to the deductive, and sometimes back and forth, to reach conclusions [@bryant2014realm].
+Physical scientists use abductive reasoning more commonly than social scientists.
 The discovery of Neptune in 1846 based on unexplained perturbations of the orbit of Uranus [@popper2005logic], the discovery of galactic migration of planets [@gomes2005n], and the current generation of gravitational wave detectors such as the Laser Interferometer Gravitational-Wave Observatory (LIGO) [@harry2010advanced] demonstrate the power of the abductive approach in the physical sciences.
@@
 We ask one fundamental question regarding demographic anomalies: What are the hidden baby booms/busts and mortality spikes/dips in the United States over the last twenty years?
-We do not necessarily know the causes of these anomalies but identifying them allows scientists with more detailed knowledge of local population dynamics, state-level policy making, or macro-economics to explain these phenomena post-hoc.
+We do not necessarily know the causes of these anomalies and suggest causes for select anomalies.
+Even without knowledge of the causes, simply identification of anomalies allows scientists with more detailed knowledge of local population dynamics, state-level policy making, or macro-economics to explain these phenomena post-hoc.
 From explanations of phenomena that may have previously gone unnoticed, social scientists may be able to better forecast populations and provide policy solutions for impending problems such as climate change.
@@
 
-## Robustness Check
-To determine the false-positive (Type I) and false-negative (Type II) error rates under a critical value of 3.5, we ran two simulations.
+<!-- To determine the false-positive (Type I) and false-negative (Type II) error rates under a critical value of 3.5, we ran two simulations.
 For the false-positive error we simulated 1000 time series with 180 time steps without introducing any anomalies and ran the `tsoutliers` algorithm on them.
@@
 If the `tsoutliers` algorithm fails to detect the anomaly, it would be considered a false-negative error.
+-->
 
-```{r , include = FALSE, cache = TRUE}
+<!-- ```{r , include = FALSE, cache = TRUE} -->
 
-set.seed(1)
-results_noutliers <- data.frame()
-# y2 <- matrix(nrow = 1000, ncol=180)
-for(i in 1:1000){
-  tryCatch({
-y <- arima.sim(model = list(ar = 0.7, ma = -0.4), n = 180)
+<!-- set.seed(1) -->
+<!-- results_noutliers <- data.frame() -->
+<!-- # y2 <- matrix(nrow = 1000, ncol=180) -->
+<!-- for(i in 1:1000){ -->
+<!--   tryCatch({ -->
+<!-- y <- arima.sim(model = list(ar = 0.7, ma = -0.4), n = 180) -->
 
-df2<- tso(y, cval = 3.5)$outliers
-df2$iteration <- i
-results_noutliers <- bind_rows(results_noutliers, df2)
-# print(i)
-  } , error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
-}
+<!-- df2<- tso(y, cval = 3.5)$outliers -->
+<!-- df2$iteration <- i -->
+<!-- results_noutliers <- bind_rows(results_noutliers, df2) -->
+<!-- # print(i) -->
+<!--   } , error=function(e){cat("ERROR :",conditionMessage(e), "\n")}) -->
+<!-- } -->
 
-# (a <- nrow(results_noutliers) / (180*1000))
-# 
-# a * 20000
+<!-- # (a <- nrow(results_noutliers) / (180*1000)) -->
+<!-- #  -->
+<!-- # a * 20000 -->
 
-results_AO <- data.frame()
-# y2 <- matrix(nrow = 1000, ncol=180)
-for(i in 1:1000){
-  tryCatch({
-    y <- arima.sim(model = list(ar = 0.7, ma = -0.4), n = 180)
-    y[150] <- 4
+<!-- results_AO <- data.frame() -->
+<!-- # y2 <- matrix(nrow = 1000, ncol=180) -->
+<!-- for(i in 1:1000){ -->
+<!--   tryCatch({ -->
+<!--     y <- arima.sim(model = list(ar = 0.7, ma = -0.4), n = 180) -->
+<!--     y[150] <- 4 -->
 
-    df2<- tso(y, cval = 3.5)$outliers
-    df2$iteration <- i
-    results_AO <- bind_rows(results_AO, df2)
-    # print(i)
-  } , error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
-}
-```
-In our sans-anomaly time series, we detect `r nrow(results_noutliers)` anomalies on the 180,000 candidate time steps for a false-positive error rate of `r prettyNum(nrow(results_noutliers) / 180000*100, digits = 3)`%.
+<!--     df2<- tso(y, cval = 3.5)$outliers -->
+<!--     df2$iteration <- i -->
+<!--     results_AO <- bind_rows(results_AO, df2) -->
+<!--     # print(i) -->
+<!--   } , error=function(e){cat("ERROR :",conditionMessage(e), "\n")}) -->
+<!-- } -->
+<!-- ``` -->
+<!-- In our sans-anomaly time series, we detect `r nrow(results_noutliers)` anomalies on the 180,000 candidate time steps for a false-positive error rate of `r prettyNum(nrow(results_noutliers) / 180000*100, digits = 3)`%.
 In our time series containing a single anomaly, we detect `r nrow(results_AO[which(results_AO$time == 150),])` anomalies out of the 1000 candidate time series, for a false-negative error rate of `r (1-(nrow(results_AO[which(results_AO$time == 150),]) / 1000)) * 100`%.
@@
 These results give us confidence that the anomalies we detect in this paper are true anomalies.
+-->
 
-![**"Toy" example of anomaly detection using the exchange rate between the Euro and the Swiss Franc.** The Swiss Franc was pegged to the Euro at 1.2 until January 15 2015 when the peg was removed and the currency was allowed to trade more freely.
+<!-- ![**"Toy" example of anomaly detection using the exchange rate between the Euro and the Swiss Franc.** The Swiss Franc was pegged to the Euro at 1.2 until January 15 2015 when the peg was removed and the currency was allowed to trade more freely.
 This is a very strong anomly in the time series (*t*-stat = -144.07) detectable using statistical methods and visually.
 The countefactural estimate with the outlier removed would simply have kept the currency exchange at 1.2 ($\hat{y}$).
-\label{explanationfigure}](../MainDocument/MainDocument_files/figure-latex/Figure1.png)
+\label{explanationfigure}](../MainDocument/MainDocument_files/figure-latex/Figure1.pdf){ width=100% height=40%} -->
 
@@
 
-stockdat <- tribble(
-  ~Date, ~Value,
-'2014-12-27',	1.203880343,
-'2014-12-28',	1.202523908,
-'2014-12-29',	1.202714098,
-'2014-12-30',	1.201988512,
-'2014-12-31',	1.202535102,
-'2015-01-01',	1.201727726,
-'2015-01-02',	1.201794127,
-'2015-01-04',	1.20132752,
-'2015-01-05',	1.20100032,
-'2015-01-06',	1.20071649,
-'2015-01-07',	1.200762136,
-'2015-01-08',	1.200787933,
-'2015-01-09',	1.20079581,
-'2015-01-11',	1.200989867,
-'2015-01-12',	1.2006267,
-'2015-01-13',	1.20058359,
-'2015-01-14',	1.200605219,
-'2015-01-15',	0.99417984,
-'2015-01-16',	0.992258896,
-'2015-01-18',	0.992353824,
-'2015-01-19',	1.019880601,
-'2015-01-20',	1.00940244,
-'2015-01-21',	0.999117945,
-'2015-01-22',	0.988725725,
-'2015-01-23',	0.939377117,
-'2015-01-24',	0.987456601,
-'2015-01-25',	0.980699544,
-'2015-01-26',	1.013809279,
-'2015-01-27',	1.02203072,
-'2015-01-28',	1.0234588,
-'2015-01-29',	1.0455615
-)
-dat3 <- ts(stockdat$Value)
-stockdat_outlier <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)
+# stockdat <- tribble(
+#   ~Date, ~Value,
+# '2014-12-27',	1.203880343,
+# '2014-12-28',	1.202523908,
+# '2014-12-29',	1.202714098,
+# '2014-12-30',	1.201988512,
+# '2014-12-31',	1.202535102,
+# '2015-01-01',	1.201727726,
+# '2015-01-02',	1.201794127,
+# '2015-01-04',	1.20132752,
+# '2015-01-05',	1.20100032,
+# '2015-01-06',	1.20071649,
+# '2015-01-07',	1.200762136,
+# '2015-01-08',	1.200787933,
+# '2015-01-09',	1.20079581,
+# '2015-01-11',	1.200989867,
+# '2015-01-12',	1.2006267,
+# '2015-01-13',	1.20058359,
+# '2015-01-14',	1.200605219,
+# '2015-01-15',	0.99417984,
+# '2015-01-16',	0.992258896,
+# '2015-01-18',	0.992353824,
+# '2015-01-19',	1.019880601,
+# '2015-01-20',	1.00940244,
+# '2015-01-21',	0.999117945,
+# '2015-01-22',	0.988725725,
+# '2015-01-23',	0.939377117,
+# '2015-01-24',	0.987456601,
+# '2015-01-25',	0.980699544,
+# '2015-01-26',	1.013809279,
+# '2015-01-27',	1.02203072,
+# '2015-01-28',	1.0234588,
+# '2015-01-29',	1.0455615
+# )
+# dat3 <- ts(stockdat$Value)
+# stockdat_outlier <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)
+
+# download.file("https://seer.cancer.gov/popdata/yr1969_2019.19ages/la.1969_2019.19ages.txt.gz", "./R/DATA-RAW/la.1969_2019.19ages.txt.gz")
+# ## UNZIPPING THE DATA FILE
+# gunzip("./R/DATA-RAW/la.1969_2019.19ages.txt.gz", overwrite = TRUE, remove = TRUE)
+
+K05_pop<- read.table("../R/DATA-RAW/la.1969_2019.19ages.txt") 
+K05_pop$V1 <- as.character(K05_pop$V1) # SETTING THE ENTIRE SINGLE VARIABLE INTO A CHARACTER
+K05_pop$YEAR <- as.numeric(substr(K05_pop$V1,1,4)) # SEPARATING THE YEAR AND SETTING IT AS A NUMBER
+K05_pop$STATEID <- substr(K05_pop$V1, 5,6) # SEPARATING THE 2 CHARACTER STATE ABBREVIATION
+K05_pop$STATE <- substr(K05_pop$V1, 7,8) # SEPARATING THE 2-DIGIT STATE CODE
+K05_pop$COUNTY <- substr(K05_pop$V1,9,11) # SEPARATING THE 3-DIGIT COUNTY CODE
+K05_pop$REGISTRY <- substr(K05_pop$V1, 12,12) # REGISTRY IS A THROW AWAY VARIABLE REFERING TO ODD GEOGRAPHIES
+K05_pop$RACE <- substr(K05_pop$V1, 14,14) # SEPARATING OUT THE RACE CODES.
+K05_pop$ORIGIN <- substr(K05_pop$V1, 15,15) # SEPARATING OUT HISPANIC ORIGIN. THIS VARIABLE IS NOT APPLICABLE IN THE 1969-2016 DATA
+K05_pop$SEX <- substr(K05_pop$V1, 16,16) # SEPARATING OUT THE SEX DATA
+K05_pop$POPULATION <- as.numeric(substr(K05_pop$V1, 19, 30)) # SEPARATING THE ACTUAL POPULATION ESTIMATES.
+
+# Setting the groupings
+GROUPING <- c("STATE", "COUNTY", "YEAR")
+
+K05_pop <- K05_pop %>%
+  group_by(.dots = GROUPING) %>%
+  dplyr::summarise(POPULATION = sum(POPULATION))
+
+start <- 1980
+K05_pop$GEOID <- paste0(K05_pop$STATE, K05_pop$COUNTY) # SETTING THE 5-DIGIT FIPS CODE
+K05_pop2 <- filter(K05_pop, GEOID == "22071",
+                  YEAR >= start)
+
+dat3 <- ts(K05_pop2$POPULATION)
+a<- tsoutliers::tso(dat3,types = c("AO","LS","TC"))
+
+K05_pop2$label = ifelse(K05_pop2$YEAR == 2006, "Anomaly Detected\n t = -96.4", "")
+
+b <- as.data.frame(forecast(auto.arima(K05_pop2$POPULATION[1:26]), h=14)) %>%
+  mutate(YEAR = seq(2006, 2019,1)) %>%
+  dplyr::select(Point.Forecast = `Point Forecast`, YEAR)
+
+b<- rbind(b, data.frame(YEAR = 2005, `Point Forecast` = 494294))
+
+```
+
+```{r ToyExample, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{An example of anomaly detection using Hurricanes Katrina and Rita in Orleans Parish Louisiana in 2005.} Hurricane Katrina struck Louisiana in 2005 and we use it as a toy example to illustrate our approach.
+This figure shows the annual time series of total population in Orleans Parish between 1980 and 2019.
+Between 1990 and 2005, Orleans Parish total population changed from 495k to 494k, suggesting a possible 'plateau' in the population (illustrated with the dotted 'counterfactual').
+Hurricane Katrina and the widespread population loss of more than 200k people represent a very strong anomaly (t=-96.39) \\label{explanationfigure}")}
+
+ggplot(data = K05_pop2, aes( x = YEAR, y = POPULATION)) +
+  geom_line() +
+  geom_point(data = K05_pop2[which(K05_pop2$label == "Anomaly Detected\n t = -96.4"),], aes(size=2), shape=1) + 
+  geom_line(data = b, aes(y = Point.Forecast, x=YEAR), linetype=2) +
+  scale_y_continuous(label=comma) +
+  geom_label_repel(aes(label = label),
+                   box.padding   = 0.4, 
+                   point.padding = 0.6,
+                   segment.color = 'grey50') +
+  annotate(geom='text', x=2021, y=500000, label=TeX("Counterfactual estimate, $\\hat{y}$", output='character'), parse=TRUE, hjust = 0) +
+  annotate(geom='text', x=2021, y=390000, label=TeX("Observed data, $y$", output='character'), parse=TRUE, hjust = 0) +
+  scale_x_continuous(breaks = seq(1980, 2020, 10), limits = c(1980,2040), expand= c(0,0)) +
+   scale_y_continuous(label=comma, limits=c(0, 565000), breaks = seq(0,570000,100000)) +
+  coord_capped_cart(bottom='right') +
+  theme_classic() +
+  theme(legend.position = "none")+
+  labs(x = "Year",
+       y = "Total Population")
+
 ```
@@
 ## Toy example
-**\autoref{explanationfigure}** shows a toy example for anomaly detection in a time series using the classic example of the exchange rate between the Euro and the Swiss Franc (CHF).
-On January 15 2015, the Swiss National Bank removed the currency peg of 1.20 francs per Euro, exposing the Franc to the volatility of the currency market.
-The Franc immediately began trading a reduced rate compared to the Euro.
-What would have been the Euro/CHF exchange rate had the Swiss National Bank *not* removed the peg?
-A simple counter-factual estimate would be to keep the exchange rate at 1.20 ($\hat{y}$, sometimes called a "synthetic control" [@abadie2010synthetic]).
+**\autoref{explanationfigure}** shows a toy example for anomaly detection in a time series using the example of Hurricane Katrina in Orleans Parish.
+On August 23 2005, Hurricane Katrina, a category 5 hurricane, struck southern Louisiana causing widespread damage and destruction in Orleans Parish in particular.
+The displacement from the Hurricane and the federal response were well documented [@horiDisplacementDynamicsSouthern2009; @fussellRecoveryMigrationCity2014] and Census estimates suggest Orleans Parish lost more than 200,000 residents between 2005 and 2006.
+What would have been Orleans' population estimate had Hurricane Katrina *not* occurred?
+A simple counter-factual estimate might keep population size just under 500,000 people ($\hat{y}$).
 
-In the Swiss Franc example, we have knowledge of the Swiss Bank's activities after the fact or ex-post-facto to to create the counter-factual time series $\hat{y}$ of 1.20.
-But is this anomaly detectable without knowledge of the Swiss Bank's activities?
-In other words, can we detect the reduction of the exchange rate on January 15 using only the time series?
+In the Hurricane Katrina example, we have knowledge of the impact of Hurricane Katrina after the fact or ex-post-facto in order to create the counter-factual time series $\hat{y}$.
+But is this anomaly detectable without knowledge of the Hurricane Katrina?
+In other words, can we detect the reduction of Orleans Parish's population between 2005 and 2006 using only the time series?
 Absolutely.
-the tsoutliers package identifies January 15 2015 as an extremely strong level-shift outlier (*t*-stat = `r round( stockdat_outlier$outliers$tstat[1], digits = 2)`).
-The real world is hardly this simplified where a direct intervention is known and is testable.
+the tsoutliers package identifies 2006 as an extremely strong level-shift outlier (*t*-stat = `r round( a$outliers$tstat[1], digits = 2)`).
+The real world is hardly this simplified where a direct intervention is known, is large, and is testable.
 
@@
 
+## Robustness Check
+Chen and Liu [-@chen1993joint] report results of a simulation exercise of varying critical value levels.
+They simulated various anomaly-free time series and injected specific anomalies of varying $\sigma$ levels.
+For a critical value of 3.5 -- the level Chen and Liu recommend to minimize both false-positive (Type I) and false-negative (Type II) error rates -- they report Type I error rates between 0.0\% and 0.1\% and Type II error rates or the statistical power between 40\% and 60\% with injected anomalies equal to 4$\sigma$.
+We follow their lead by simulating time series to test for Type I and Type II errors.
+We base our simulation on the `r length(noamom_fert) + length(noamom_mort)` time series that do not contain any anomalies (see **Results** below; `r length(noamom_fert)` without a single fertility anomaly and `r length(noamom_mort)` without a single mortality anomaly).
+
+To test for false-positive rates, we simulate 1000 time series for each of the `r length(noamom_fert) + length(noamom_mort)` state-components (fertility or mortality) using the fitted ARIMA model for each state-component.
+This allows the simulated time series to mimic the underlying true time series (see **Supplementary Material** for the underlying code).
+Any anomalies we detect in the `r prettyNum(z2$timesteps, big.mark=',')` candidate time steps are spurious anomalies.
+We detect `r prettyNum(z2$tot, big.mark=',')` anomalies for a false-positive error rate of `r percent(z2$FNR, accuracy = 0.001)`.
+
+To test for false-negative rates, we use the underlying true, sans-anomaly time series but introduce a single anomaly (magnitude equal to 3.5$\sigma$) at the same time point in each series.
+If the `tsoutliers` algorithm fails to detect the anomaly, it would be considered a false-negative error.
+We detect `r  prettyNum(nrow(results_AO[which(results_AO$time == 150),]), big.mark=',')` out of `r prettyNum((length(noamom_fert) + length(noamom_mort))*1000,, big.mark=',')` candidate time series giving us a false-negative error rate of `r percent((1-(nrow(results_AO[which(results_AO$time == 150),]) / ((length(noamom_fert) + length(noamom_mort))*1000))), 0.1)`.
+This provides a statistical power of `r percent(((nrow(results_AO[which(results_AO$time == 150),]) / ((length(noamom_fert) + length(noamom_mort))*1000))), 0.1)`.
+
+Based on these tests we think it is clear that our probability of committing a Type I error (or claiming an anomaly is true when in fact is not) is very low.
+But, based on our false-negative error rate, it is possible that the anomalies we do detect are *conservative*.
+Our statistical power is of just `r percent(((nrow(results_AO[which(results_AO$time == 150),]) / ((length(noamom_fert) + length(noamom_mort))*1000))), 0.1)` suggests a higher than normal likelihood that true effects are going undetected.
+We are unlikely to commit an error of commission (ie, including a false anomaly) but are more likely to commit an error of omission (ie omitting true anomalies).
+Thus our results should be considered conservative.
+
 ## Replication Files
-All data and code necessary to reproduce the reported results are licensed under the CC-BY-4.0 license and are publicly available in a replication repository located at https://osf.io/fp52x/?view_only=754d9a72a2ea4f6b8e0c193dc9a590d1.
-The analyses were performed in *R* [@rcite].
+All data and code necessary to reproduce the reported results are licensed under the CC-BY-4.0 license and are publicly available as https://osf.io/yc3r4/?view_only=3ce3ca267c7047e0942940ebd08846d0.
+The analyses were performed in *R* [@rcore].
 
@@
 We begin by highlighting significant fertility and mortality anomalies across all three types of outliers (Additive Outliers, Level Shift Outliers, and Temporary Change outliers) with plausible explanations.
-We then highlight two strong anomalies that bely explanation.
+We then highlight two strong anomalies that belie explanation.
 Finally, we conclude with a summary of the anomalies we detect.
@@
 conn_per <- percent(abs(sum(outlier.fert.conn$y - outlier.fert.conn$yadj) / sum(outlier.fert.conn$yadj[80:180])))
+
+mag_la1 <- round(outlier.fert.la$outliers$coefhat[1],1)
+mag_la2 <- round(outlier.fert.la$outliers$coefhat[2],1)
+# int_la1 <- round(outlier.fert.la$outliers$coefhat[1] - confint(outlier.fert.la$fit, level=0.95)[5],1)
+# int_la2 <- round(outlier.fert.la$outliers$coefhat[1] - confint(outlier.fert.la$fit, level=0.95)[6],1)
 ```
@@
 
-```{r FertilityAnomalies, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Louisiana (a) and Connecticut (b) state fertility, 2003-2018.} The top part of each panel contains the original time series (light gray), the corrected, counter-factual time series in the absence of anomalies (blue), and the red dots correspond to the onset of detected anomalies.
+```{r , include=FALSE}
+n.sims <- 1e3
+
+# This function will simulate 100 time series with similar values as the target database and make a ggplot figure version of tsoutliers baseplot but with an 95th percentile confidence interval.
+makefigure <- function(this.outlier, titles){
+set.seed(1)
+sims.hist <- matrix(NA,nrow=n.sims,ncol=length(this.outlier$y))
+# for ( ii in 1:n.sims ) {
+#   # setWinProgressBar(pb,ii,paste(ii,"of",n.sims))
+#   sims.hist[ii,] <- this.outlier$yadj+rnorm(length(this.outlier$yadj),0,sd(this.outlier$yadj))
+#   model <- ets(ts(sims.hist[ii,],frequency=frequency(this.outlier$yadj)))
+# }
+
+   createsim <- function(ii){
+    # set.seed(runif(1,1,1000))
+    arima.sim(n=length(this.outlier$y), model=list(this.outlier$fit), sd = sdval) + as.vector(this.outlier$yadj)
+   }
+     sdval <- sqrt(this.outlier$fit$sigma2)
+  for ( ii in 1:n.sims ) {
+    sims.hist[ii,] <- as.vector(createsim())
+  }
+
+conf <- c(0.95)
+ylim <- range(
+  apply(sims.hist,2,quantile,prob=c((1-max(conf))/2,1-(1-max(conf))/2)))
+boundaries.hist <- apply(sims.hist,2,quantile,prob=c((1-rev(conf)[1])/2,1-(1-rev(conf)[1])/2))
+
+a <- data.frame(Y=as.matrix(this.outlier$yadj),
+                date = time(this.outlier$yadj),
+                ymin = as.matrix(boundaries.hist[1,]),
+                ymax = as.matrix(boundaries.hist[2,])
+)
+b <- data.frame(y = as.matrix(this.outlier$y),
+                date = time(this.outlier$y))
+pts <- data.frame(date = this.outlier$times,
+                  y = this.outlier$y[this.outlier$outliers$ind])
+
+effectsizes <- data.frame(date =time(this.outlier$y),
+                           y= this.outlier$y - this.outlier$yadj)
+
+top <- ggplot(data= a, aes(x = date, y=Y)) +
+  geom_line(data= b, aes(x=date, y=y), color = "red") +
+  geom_line() +
+  geom_ribbon(aes(ymin = ymin, ymax = ymax), alpha = 0.2) +
+  geom_point(data=pts, aes(x=date, y =y), color = "red", fill = NA, size = 2, shape =21) +
+  theme_bw() +
+  scale_y_continuous(labels=comma) +
+  labs(x="", y ="",
+       title = paste(titles, "<br>
+       <span style='font-size:10pt'><span style='color:#FF0000'>**Original**</span> and **Adjusted**  series</span>"
+       # caption = "Shaded region is the 95th percentile confidence interval"
+       )) +
+  theme(
+    plot.title = element_markdown(lineheight = 1.2),
+    axis.title.x=element_blank(),
+    axis.text.x=element_blank(),
+    axis.ticks.x = element_blank(),
+    plot.margin = unit(c(0.1, 0, 0, 0), "cm")
+  )
+
+bot <- 
+  ggplot(data = effectsizes, aes(x=date, y=y)) +
+  geom_line(color="blue") +
+  theme_bw() +
+  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
+  scale_y_continuous(labels=comma,
+                     position = "right") +
+  labs(x="", y ="")
+
+together <- plot_grid(top, bot, ncol=1,
+           align = "v", rel_heights = c(1,0.6))
+
+return(together)
+}
+
+```
+
+```{r FertilityAnomalies, echo= FALSE, message = FALSE, warning = FALSE, fig.height = 6, fig.cap= paste("\\textbf{Anomaly detection in Louisiana (a) and Connecticut (b) state fertility, 2003-2018.} The top part of each panel contains the original time series (light gray), the adjusted, counter-factual time series in the absence of anomalies (blue), and the red dots correspond to the onset of detected anomalies.
 The bottom part of each panel contains the magnitude and type of the outlier in red.
@@
 
-a <-  function() {
-  plot(outlier.fert.la, cex.axis=0.8)
-  title(main = "Louisiana - Fertility", adj = 0.9)
-}
+a<- makefigure(outlier.fert.la, "Louisiana - Fertility")
+b <- makefigure(outlier.fert.conn, "Connecticut - Fertility")
+# a <-  function() {
+#   plot(outlier.fert.la, cex.axis=0.8)
+#   title(main = "Louisiana - Fertility", adj = 0.9)
+# }
+# 
+# b <- function(){
+#   plot(outlier.fert.conn, cex.axis=0.8) +
+#   title(main = "Connecticut - Fertility", adj = 0.88)
+# }
 
-b <- function(){
-  plot(outlier.fert.conn, cex.axis=0.8) +
-  title(main = "Connecticut - Fertility", adj = 0.88)
-}
+# plot_grid(ggdraw(a) ,
+#                ggdraw(b), 
+#                ncol=1,
+#                labels= "auto")
 
-plot_grid(ggdraw(a) ,
-               ggdraw(b), 
+together <- plot_grid(a,
+               b, 
                ncol=1,
-               labels= "auto")
+               labels= "auto",
+               rel_heights = c(1, 1))
+ggdraw(add_sub(together, "Shaded region is the 95th percentile confidence interval", size = 8 ,x=1, hjust=1.1))
 ```
@@
 
-```{r MortalityAnomalies, echo= FALSE, message = FALSE, warning = FALSE,dev=c('pdf','png'), fig.height = 5, fig.cap= paste("\\textbf{Anomaly Detection for New York (a) and New Hampshire (b) state mortality, 1999-2016.} In New York (a), we detect additive outliers (AO) in January 2000, September 2001, January 2005, and January 2013; temporary change (TC) outliers in February 2007 and January 2015; and a level shift (LS) starting in February 2004.
+```{r MortalityAnomalies, echo= FALSE, message = FALSE, warning = FALSE,dev=c('pdf','png'), fig.height = 6, fig.cap= paste("\\textbf{Anomaly Detection for New York (a) and New Hampshire (b) state mortality, 1999-2016.} In New York (a), we detect additive outliers (AO) in January 2000, September 2001, January 2005, and January 2013; temporary change (TC) outliers in February 2007 and January 2015; and a level shift (LS) starting in February 2004.
 In New Hampshire (b), we detect two outliers, both level shift outliers (LS) in April 2010 and again in November 2014.
@@
 
-a <-  function() {
-plot(outlier.mort.ny, cex = 0.8)
-title("New York - Mortality", adj = 0.9, line =2)
-}
+# a <-  function() {
+# plot(outlier.mort.ny, cex = 0.8)
+# title("New York - Mortality", adj = 0.9, line =2)
+# }
+# 
+# b <-  function() {
+#   plot(outlier.mort.nh, cex = 0.8)
+#   title("New Hampshire - Mortality", adj = 0.82,line =2)
+# }
+# 
+# plot_grid(ggdraw(a) ,
+#                ggdraw(b), 
+#                ncol=1,
+#                labels= "auto")
 
-b <-  function() {
-  plot(outlier.mort.nh, cex = 0.8)
-  title("New Hampshire - Mortality", adj = 0.82,line =2)
-}
+a<- makefigure(outlier.mort.ny, "New York - Mortality")
+b <- makefigure(outlier.mort.nh, "New Hampshire - Mortality")
 
-plot_grid(ggdraw(a) ,
-               ggdraw(b), 
+together <- plot_grid(a,
+               b, 
                ncol=1,
-               labels= "auto")
+               labels= "auto",
+               rel_heights = c(1, 1))
+ggdraw(add_sub(together, "Shaded region is the 95th percentile confidence interval", size = 8 ,x=1, hjust=1.1))
+
 ```
@@
 This is contrast to **\autoref{fig:ferthawaii}b**, where we identify a strange mortality reduction in Ohio (*t*-stat: `r oh_tstat`).
-This LS is more than `r oh_deaths` deaths per month less than the counter-factual time series, suggesting had nearly `r oh_tot` fewer deaths since February 2015 than expected.
+This LS is more than `r oh_deaths` deaths per month less than the counter-factual time series, suggesting Ohio had nearly `r oh_tot` fewer deaths since February 2015 than expected.
 This is the single-largest LS among all states.
@@
 
-```{r TrueAnomalies, echo= FALSE, message = FALSE, warning = FALSE, fig.height = 5, fig.cap= paste("\\textbf{Anomaly detection in Hawaii state fertility (a) and Ohio state mortality (b).} Here we detect one outlier, an additive outlier (AO) in May 2014 in Hawaii (a), representing a large, unexplainable 14% increase in expected births in that month.
+```{r TrueAnomalies, echo= FALSE, message = FALSE, warning = FALSE, fig.height = 7, fig.cap= paste("\\textbf{Anomaly detection in Hawaii state fertility (a) and Ohio state mortality (b).} Here we detect one outlier, an additive outlier (AO) in May 2014 in Hawaii (a), representing a large, unexplainable 14% increase in expected births in that month.
 We also detect a strong level shift (LS) reduction (b), representing a large, unexplainable reduction of nearly 40,000 deaths in Ohio.
@@
 
-a <-  function() {
-plot(outlier.fert.hi, cex.axis=0.8)
-  title(main = "Hawaii - Fertility", adj = 0.88)
-}
+# a <-  function() {
+# plot(outlier.fert.hi, cex.axis=0.8)
+#   title(main = "Hawaii - Fertility", adj = 0.88)
+# }
+# 
+# b <-  function() {
+# plot(outlier.mort.oh, cex.axis=0.8)
+#   title(main = "Ohio - Mortality", adj = 0.9)
+# }
 
-b <-  function() {
-plot(outlier.mort.oh, cex.axis=0.8)
-  title(main = "Ohio - Mortality", adj = 0.9)
-}
+# 
+# plot_grid(ggdraw(a) ,
+#                ggdraw(b), 
+#                ncol=1,
+#                labels= "auto")
 
-plot_grid(ggdraw(a) ,
-               ggdraw(b), 
+a<- makefigure(outlier.fert.hi, "Hawaii - Fertility")
+b <- makefigure(outlier.mort.oh, "Ohio - Mortality")
+
+together <- plot_grid(a,
+               b, 
                ncol=1,
-               labels= "auto")
+               labels= "auto",
+               rel_heights = c(1, 1))
+ggdraw(add_sub(together, "Shaded region is the 95th percentile confidence interval", size = 8 ,x=1, hjust=1.1))
+
 ```

### 2021-03-29 MainDocument/MainDocument.tex
Updated.

@@
 %% load any required packages here
+
+% Pandoc citation processing
 
@@
 {
-  \title{\bf Causal Inference in Population Trends: Searching for Demographic
-Anomalies in Big Data}
+  \title{\bf Causal Inference in Population Trends: Searching for
+Demographic Anomalies in Big Data}
 
   \author{
-        Mathew E. Hauer \thanks{Thanks y'all!} \\
+        Mathew E. Hauer \\
     Department of Sociology, Florida State University\\
@@
 and deaths, respectively. Notable detectable mortality anomalies include
-the September 11 2001 terrorist attack in New York, the emergence and
-acceleration of the opioid epidemic in New Hampshire, and the excess
-mortality due to Hurricane Maria in Puerto Rico. Notable fertility
+the September 11 2001 terrorist attack in New York and the emergence and
+acceleration of the opioid epidemic in New Hampshire. Notable fertility
 anomalies include the ``missing births'' in Louisiana after Hurricane
 Katrina and the reduction in fertility behavior after the September 2008
-stock market crash in Connecticut, amongst others. Our combined causal
+stock market crash in Connecticut, amongst others. The combined causal
 inference and abductive reasoning approach can be readily adapted to
@@
 
-The search for causality is one of the primary motivations behind social
-science \citep{smith2011commentary, pearl2009causal}. Causal Inference,
-the identification of causal mechanisms, attempts to uncover the
+The search for causality is one of the primary motivations of social
+scientists \citep{smith2011commentary, pearl2009causal}. Causal
+inference, the identification of causal effects, attempts to uncover the
 underlying mechanisms that result in changes in a phenomenon and has a
@@
 widespread adoption in the social sciences \citep{west2008ajph}. RCTs
-are also sometimes impossible to deploy in the ``real world''. For
+are also sometimes impossible to deploy in the ``real world.'' For
 instance, an RCT to study country-level austerity measures would be
-difficult, if not impossible, to implement. Natural experiments are a
-cheaper alternative, but provide less certainty about causation, fewer
-opportunities to replicate work, and require expert knowledge of the
-natural experiment to be properly employed \citep{pearl2018book}.
+difficult, if not impossible, to implement. RCTs to study fertility
+would be immoral. Natural experiments are a cheaper alternative but
+provide less certainty about causation, fewer opportunities to replicate
+work, and require expert knowledge of the natural experiment to be
+properly employed \citep{pearl2018book}.
 
-Additionally, both RCTs and natural experiments follow traditional
-hypothesis testing, the deductive scientific paradigm where a question
-is first posed and scientists generate or find data to answer the
-question. Such approaches, while extensively utilized and time-tested,
-are likely to miss important phenomena that might go unnoticed or suffer
-from extensive p-hacking
-\citep{head2015extent, nuzzo2014scientific, ruggles2014big}. Abductive
-reasoning, sometimes described as ``inferring cause from effect''
-\citep{Crowder2017}, moves the scientific process from the inductive to
-the deductive, and sometimes back and forth, to reach conclusions
-\citep{bryant2014realm} The physical sciences use abductive reasoning
-more fully compared to the social sciences. The discovery of Neptune in
-1846 based on unexplained perturbations of the orbit of Uranus
-\citep{popper2005logic}, the discovery of galactic migration of planets
-\citep{gomes2005n}, and the current generation of gravitational wave
-detectors such as the Laser Interferometer Gravitational-Wave
-Observatory (LIGO) \citep{harry2010advanced} demonstrate the power of
-the abductive approach in the physical sciences.
+Additionally, both RCTs and natural experiments are based on traditional
+hypothesis-based testing, the deductive scientific paradigm where a
+question is first posed and scientists generate or find data to answer
+the question. Such approaches, while extensively utilized and
+time-tested, increase the likelihood that a researcher will miss
+important phenomena that might go unnoticed because it falls outside the
+hypothesis, or the study could suffer from extensive p-hacking
+\citep{head2015extent, nuzzo2014scientific, ruggles2014big}. At the same
+time, purely inductive data mining can result in false assertions or
+spurious findings, especially in the hands of those who are not field
+experts \citep{yanai_hypothesis_2020}. Abductive reasoning, sometimes
+described as ``inferring cause from effect'' \citep{Crowder2017}, moves
+the scientific process from the inductive to the deductive, and
+sometimes back and forth, to reach conclusions \citep{bryant2014realm}.
+Physical scientists use abductive reasoning more commonly than social
+scientists. The discovery of Neptune in 1846 based on unexplained
+perturbations of the orbit of Uranus \citep{popper2005logic}, the
+discovery of galactic migration of planets \citep{gomes2005n}, and the
+current generation of gravitational wave detectors such as the Laser
+Interferometer Gravitational-Wave Observatory (LIGO)
+\citep{harry2010advanced} demonstrate the power of the abductive
+approach in the physical sciences.
 
@@
 researchers to find hidden or understudied social phenomena
-\citep{bohon2018demography}. While the availability of Big Data and high
+\citep{bohon2018demography}. While the availability of big data and high
 performance computing allows novel exploration of data through causal
@@
 algorithm \citep{chen1993joint}, abductive reasoning, and large
-demographic data sets on more than 20 years of mortality/fertility to
-identify anomalous demographic behavior. In essence, we identify
-\emph{effects without knowledge of the cause}. We ask one fundamental
-question regarding demographic anomalies: What are the hidden baby
-booms/busts and mortality spikes/dips in the United States over the last
-twenty years? We do not necessarily know the causes of these anomalies
-but identifying them allows scientists with more detailed knowledge of
-local population dynamics, state-level policy making, or macro-economics
-to explain these phenomena post-hoc. From explanations of phenomena that
-may have previously gone unnoticed, social scientists may be able to
-better forecast populations and provide policy solutions for impending
-problems such as climate change.
+demographic data sets on more than 20 years of mortality and fertility
+trends to identify anomalous demographic behavior. In essence, we
+identify \emph{effects without knowledge of the cause}. We ask one
+fundamental question regarding demographic anomalies: What are the
+hidden baby booms/busts and mortality spikes/dips in the United States
+over the last twenty years? We do not necessarily know the causes of
+these anomalies and suggest causes for select anomalies. Even without
+knowledge of the causes, simply identification of anomalies allows
+scientists with more detailed knowledge of local population dynamics,
+state-level policy making, or macro-economics to explain these phenomena
+post-hoc. From explanations of phenomena that may have previously gone
+unnoticed, social scientists may be able to better forecast populations
+and provide policy solutions for impending problems such as climate
+change.
 
@@
 theory is correct. The classic example of deductive reasoning
-essentially goes (a) ``all men are mortal'', (b) ``Socrates is a man'',
-(c) ``Therefore, Socrates is mortal.'' Deductive reasoning has a long
+essentially goes (a) ``all men are mortal,'' (b) ``Socrates is a man,''
+(c) ``therefore, Socrates is mortal.'' Deductive reasoning has a long
 history in the sciences and most hypothesis testing in the social
 sciences follows this logical reasoning. For example, John Snow used
-deductive reasoning in to stop the 1854 London cholera outbreak. He
+deductive reasoning to stop the 1854 London cholera outbreak. He
 hypothesized that cholera was waterborne and deduced that a bad water
@@
 
-Inductive reasoning is the opposite of deductive reasoning. Rather than
-starting with a theory and making a prediction, induction starts with an
-observation and makes a theory based on the likeliest reason. For
-example, the Hispanic Mortality Paradox first appeared in the scientific
-literature in the mid-1980s \citep{markides1986health} as an observation
-that hispanics in the United States had, paradoxically, better health
+Inductive reasoning is directionally opposite of deductive reasoning.
+Rather than starting with a theory and making a prediction, induction
+starts with an observation and makes a theory based on the likeliest
+reason. For example, the Hispanic Mortality Paradox first appeared in
+the scientific literature in the mid-1980s \citep{markides1986health} as
+an observation that Hispanics in the United States had better health
 outcomes than expected given their socioeconomic status. Based on this
@@
 but are not without their shortcomings. In both approaches, it is
-impossible, or at least imporbable, for the premise to be true and the
+impossible, or at least improbable, for the premise to be true and the
 conclusion false. Furthermore, if a hypothesis is rejected under a
@@
 data rich environment -- the very data environment social scientists and
-demographers find themselves in the present moment. Additionally, in a
-data rich environment, a hypothesis is a hidden liability
-\citep{yanai_hypothesis_2020}because it focuses our attention on only
-that which we seek\footnote{Yanai and Lercher
+demographers find themselves in the present moment. In a data rich
+environment, a hypothesis is a hidden liability
+\citep{yanai_hypothesis_2020} because it focuses our attention on only
+that which we seek. To demonstrate this, \footnote{Yanai and Lercher
   \citeyearpar{yanai_hypothesis_2020} ask if focusing on a specific
-  hypothesis prevents new discovery? They created a simple dataset that
-  if plotted would reveal an image of a gorilla waving. Students
-  instructed to find support for specific hypotheses were less likely to
-  discover the gorilla compared to those who were simply asked ``What do
-  you conclude from the dataset?''}. Focusing on specific hypotheses
-often prevents us from finding other interesting phenomena. Afterall, it
-was the publication of the hispanic paradox itself that has generated
-more than thirty years of devoted scholarship. When paired with causal
-inference techniques, abductive reasoning provides an empirical platform
-for the discovery of potentially interesting social phenomena; it
-provides an important tool for sifting large amounts of data to find
-interesting results in the absence of hypothesis testing.
+  hypothesis prevents new discovery. By way of illustration, they
+  created a simple dataset that if plotted would reveal an image of a
+  gorilla waving. Students instructed to find support for specific
+  hypotheses were less likely to discover the gorilla compared to those
+  who were simply asked ``What do you conclude from the dataset?''}.
+Focusing on specific hypotheses often prevents us from finding other
+interesting phenomena. After all, it was the publication of the Hispanic
+paradox itself that has generated more than thirty years of devoted
+scholarship. At the same time, when paired with causal inference
+techniques, abductive reasoning provides an empirical platform for the
+discovery of potentially interesting social phenomena; it provides an
+important tool for sifting large amounts of data to find interesting
+results without the blinders of hypothesis testing.
 
@@
 
-\hypertarget{robustness-check}{%
-\subsection{Robustness Check}\label{robustness-check}}
-
-To determine the false-positive (Type I) and false-negative (Type II)
-error rates under a critical value of 3.5, we ran two simulations. For
-the false-positive error we simulated 1000 time series with 180 time
-steps without introducing any anomalies and ran the \texttt{tsoutliers}
-algorithm on them. Any anomalies we detect in these time series are
-spurious anomalies. For the false-negative error, we simulated 1000 time
-series with 180 time steps but introduced a single anomaly of the same
-magnitude at the same time point in each time series. If the
-\texttt{tsoutliers} algorithm fails to detect the anomaly, it would be
-considered a false-negative error.
-
-In our anomaly-less time series, we detect 208 anomalies on the 180,000
-candidate time steps for a false-positive error rate of 0.116\%. In our
-time series containing a single anomaly, we detect 817 anomalies out of
-the 1000 candidate time series, for a false-negative error rate of
-18.3\%. This provides a statistical power of 0.817. These results give
-us confidence that the anomalies we detect in this paper are true
-anomalies.
-
 \begin{figure}
 \centering
-\includegraphics{../MainDocument/MainDocument_files/figure-latex/Figure1.png}
-\caption{\textbf{``Toy'' example of anomaly detection using the exchange
-rate between the Euro and the Swiss Franc.} The Swiss Franc was pegged
-to the Euro at 1.2 until January 15 2015 when the peg was removed and
-the currency was allowed to trade more freely. This is a very strong
-anomly in the time series (\emph{t}-stat = -144.07) detectable using
-statistical methods and visually. The countefactural estimate with the
-outlier removed would simply have kept the currency exchange at 1.2
-(\(\hat{y}\)). \label{explanationfigure}}
+\includegraphics{MainDocument_files/figure-latex/ToyExample-1.pdf}
+\caption{\textbf{An example of anomaly detection using Hurricanes Katrina and Rita in Orleans Parish Louisiana in 2005.}
+Hurricane Katrina struck Louisiana in 2005 and we use it as a toy
+example to illustrate our approach. This figure shows the annual time
+series of total population in Orleans Parish between 1980 and 2019.
+Between 1990 and 2005, Orleans Parish total population changed from 495k
+to 494k, suggesting a possible `plateau' in the population (illustrated
+with the dotted `counterfactual'). Hurricane Katrina and the widespread
+population loss of more than 200k people represent a very strong anomaly
+(t=-96.39) \label{explanationfigure}}
 \end{figure}
@@
 \textbf{\autoref{explanationfigure}} shows a toy example for anomaly
-detection in a time series using the classic example of the exchange
-rate between the Euro and the Swiss Franc (CHF). On January 15 2015, the
-Swiss National Bank removed the currency peg of 1.20 francs per Euro,
-exposing the Franc to the volatility of the currency market. The Franc
-immediately began trading a reduced rate compared to the Euro. What
-would have been the Euro/CHF exchange rate had the Swiss National Bank
-\emph{not} removed the peg? A simple counter-factual estimate would be
-to keep the exchange rate at 1.20 (\(\hat{y}\), sometimes called a
-``synthetic control'' \citep{abadie2010synthetic}).
+detection in a time series using the example of Hurricane Katrina in
+Orleans Parish. On August 23 2005, Hurricane Katrina, a category 5
+hurricane, struck southern Louisiana causing widespread damage and
+destruction in Orleans Parish in particular. The displacement from the
+Hurricane and the federal response were well documented
+\citep{horiDisplacementDynamicsSouthern2009, fussellRecoveryMigrationCity2014}
+and Census estimates suggest Orleans Parish lost more than 200,000
+residents between 2005 and 2006. What would have been Orleans'
+population estimate had Hurricane Katrina \emph{not} occurred? A simple
+counter-factual estimate might keep population size just under 500,000
+people (\(\hat{y}\)).
 
-In the Swiss Franc example, we have knowledge of the Swiss Bank's
-activities after the fact or ex-post-facto to to create the
-counter-factual time series \(\hat{y}\) of 1.20. But is this anomaly
-detectable without knowledge of the Swiss Bank's activities? In other
-words, can we detect the reduction of the exchange rate on January 15
-using only the time series? Absolutely. the tsoutliers package
-identifies January 15 2015 as an extremely strong level-shift outlier
-(\emph{t}-stat = -144.07). The real world is hardly this simplified
-where a direct intervention is known and is testable.
+In the Hurricane Katrina example, we have knowledge of the impact of
+Hurricane Katrina after the fact or ex-post-facto in order to create the
+counter-factual time series \(\hat{y}\). But is this anomaly detectable
+without knowledge of the Hurricane Katrina? In other words, can we
+detect the reduction of Orleans Parish's population between 2005 and
+2006 using only the time series? Absolutely. the tsoutliers package
+identifies 2006 as an extremely strong level-shift outlier
+(\emph{t}-stat = -96.39). The real world is hardly this simplified where
+a direct intervention is known, is large, and is testable.
 
@@
 
+\hypertarget{robustness-check}{%
+\subsection{Robustness Check}\label{robustness-check}}
+
+Chen and Liu \citeyearpar{chen1993joint} report results of a simulation
+exercise of varying critical value levels. They simulated various
+anomaly-free time series and injected specific anomalies of varying
+\(\sigma\) levels. For a critical value of 3.5 -- the level Chen and Liu
+recommend to minimize both false-positive (Type I) and false-negative
+(Type II) error rates -- they report Type I error rates between 0.0\%
+and 0.1\% and Type II error rates or the statistical power between 40\%
+and 60\% with injected anomalies equal to 4\(\sigma\). We follow their
+lead by simulating time series to test for Type I and Type II errors. We
+base our simulation on the 41 time series that do not contain any
+anomalies (see \textbf{Results} below; 33 without a single fertility
+anomaly and 8 without a single mortality anomaly).
+
+To test for false-positive rates, we simulate 1000 time series for each
+of the 41 state-components (fertility or mortality) using the fitted
+ARIMA model for each state-component. This allows the simulated time
+series to mimic the underlying true time series (see
+\textbf{Supplementary Material} for the underlying code). Any anomalies
+we detect in the 7,764,000 candidate time steps are spurious anomalies.
+We detect 6,588 anomalies for a false-positive error rate of 0.085\%.
+
+To test for false-negative rates, we use the underlying true,
+sans-anomaly time series but introduce a single anomaly (magnitude equal
+to 3.5\(\sigma\)) at the same time point in each series. If the
+\texttt{tsoutliers} algorithm fails to detect the anomaly, it would be
+considered a false-negative error. We detect 27,904 out of 41,000
+candidate time series giving us a false-negative error rate of 31.9\%.
+This provides a statistical power of 68.1\%.
+
+Based on these tests we think it is clear that our probability of
+committing a Type I error (or claiming an anomaly is true when in fact
+is not) is very low. But, based on our false-negative error rate, it is
+possible that the anomalies we do detect are \emph{conservative}. Our
+statistical power is of just 68.1\% suggests a higher than normal
+likelihood that true effects are going undetected. We are unlikely to
+commit an error of commission (ie, including a false anomaly) but are
+more likely to commit an error of omission (ie omitting true anomalies).
+Thus our results should be considered conservative.
+
 \hypertarget{replication-files}{%
@@
 All data and code necessary to reproduce the reported results are
-licensed under the CC-BY-4.0 license and are publicly available in a
-replication repository located at
-\url{https://osf.io/fp52x/?view_only=754d9a72a2ea4f6b8e0c193dc9a590d1}.
-The analyses were performed in \emph{R} \citep{rcite}.
+licensed under the CC-BY-4.0 license and are publicly available as
+\url{https://osf.io/yc3r4/?view_only=3ce3ca267c7047e0942940ebd08846d0}.
+The analyses were performed in \emph{R} \citep{rcore}.
 
@@
 outliers) with plausible explanations. We then highlight two strong
-anomalies that bely explanation. Finally, we conclude with a summary of
+anomalies that belie explanation. Finally, we conclude with a summary of
 the anomalies we detect.
@@
 The top part of each panel contains the original time series (light
-gray), the corrected, counter-factual time series in the absence of
+gray), the adjusted, counter-factual time series in the absence of
 anomalies (blue), and the red dots correspond to the onset of detected
@@
 place that might have contributed to this considerable mortality
-reduction? What environmental or economic conditions may have changed
-may have changed? These are the kinds of questions that arise from our
-analyses. The purpose of our paper is not to answer these questions, but
-our findings underscore the need for more research that takes an
-abductive approach. By identifying anomalies through an inductive
-process, researchers can then look for underlying causes. Once those
-causes are identified, researchers can then use a deductive process to
-see if such events predict other (or future) anomalies.
+reduction? What environmental or economic conditions may have changed?
+These are the kinds of questions that arise from our analyses. The
+purpose of our paper is not to answer these questions, but our findings
+underscore the need for more research that takes an abductive approach.
+By identifying anomalies through an inductive process, researchers can
+then look for underlying causes. Once those causes are identified,
+researchers can then use a deductive process to see if such events
+predict other (or future) anomalies.
 
@@
 This LS is more than 1,157 deaths per month less than the
-counter-factual time series, suggesting had nearly 40,000 fewer deaths
-since February 2015 than expected. This is the single-largest LS among
-all states. We could not identify the potential policies Ohio might have
-put into place to provide such a strong mortality protection.
+counter-factual time series, suggesting Ohio had nearly 40,000 fewer
+deaths since February 2015 than expected. This is the single-largest LS
+among all states. We could not identify the potential policies Ohio
+might have put into place or the events that occurred to provide such a
+strong mortality protection.
 
@@
 fertility anomalies yielding more births. This is consistent with
-research findings that link high severity weather events to
-significantly lower fertility \citep{Evans2008Hurricanebirth}.
-Conversely, mortality anomalies tend to be more evenly split between
-positive and negative anomalies, but tend toward more deaths rather than
-fewer deaths. Again, this is not surprising, as the kinds of events that
-increase death (e.g., plant closings are associated with increased
-opioid death \citep{Venkataramani2019}) are more common than the events
-that decrease death (e.g., cannabis legalization is associated with
-decreased opioid deaths \citep{Livingston2017Cannabis}). These anomalous
-deaths and births account for 1.44\% and 0.889\% over the entire time
-series, respectively.
+research findings that link severe weather events to significantly lower
+fertility \citep{Evans2008Hurricanebirth}. Conversely, mortality
+anomalies tend to be more evenly split between positive and negative
+anomalies, but tend toward more deaths rather than fewer deaths. Again,
+this is not surprising, as the kinds of events that increase death
+(e.g., plant closings are associated with increased opioid death
+\citep{Venkataramani2019}) are more common than the events that decrease
+death (e.g., cannabis legalization is associated with decreased opioid
+deaths \citep{Livingston2017Cannabis}). These anomalous deaths and
+births account for 1.44\% and 0.889\% over the entire time series,
+respectively.
 
@@
 mortality nor fertility anomalies: Alaska, North Dakota, and South
-Dakota. Another nine states exhibited just a single anomaly (DE, DC, ID,
-IL, KS, NV, NM, UT, and WY). Both New York and Massachusetts exhibited
-the most anomalies with nine.
+Dakota. Another nine states exhibited just a single anomaly (Delaware,
+District of Colombia, Idaho, Illinois, Kansas, Nevada, New Mexico, Utah,
+and Wyoming). Both New York and Massachusetts exhibited the most
+anomalies with nine.
 
@@
 the complexities of the social world and know from experience that data
-alone is not enough to solve social problems
+alone are not enough to solve social problems
 \citep{bohon2018demography, grimmer2015ppsp}. Nonetheless, the increased

### 2021-03-29 R/SCRIPTS/001-AnomaliesDetection.R

@@ -1,10 +1,11 @@
-###------Mort Loop-----
-## @knitr MortLoop
+###------Anomalies Detection-----
+## @knitr AnomaliesDetection
 
+# Reading the fertility data in
 dat <-
   ###FERTILITY
-  bind_rows(read_tsv("./R/DATA-RAW/Fertility/Natality, 2003-2006,states.txt"),
-            read_tsv("./R/DATA-RAW/Fertility/Natality, 2007-2017,states.txt"))%>%
+  bind_rows(read_tsv("../R/DATA-RAW/Fertility/Natality, 2003-2006,states.txt"),
+            read_tsv("../R/DATA-RAW/Fertility/Natality, 2007-2017,states.txt"))%>%
   mutate(Month2 = str_pad(`Month Code`,2 , pad = "0"),
          Month3 = as.yearmon(paste0(Year,"-", Month2)),
          location = paste0("X", `State Code`)
@@ -13,49 +14,47 @@ dat <-
   na.omit %>%
   arrange(location, Year)
 
+# Generating a list of States
 countylist = unique(dat$location)
 
+# Setting up the data frames to collect the results
 df <- data.frame()
 df_sums <- data.frame()
+# Setting the t-stat level.
 sigma <- 3.5
-
-for(this.state in unique(countylist)){
+# Looping through the states. For illustrative purposes, this loops through the 1st 5 states.
+for(this.state in unique(countylist)[1:5]){
   set.seed(1)
   print(this.state)
   tryCatch({
     df2<- setNames(data.frame(matrix(ncol = 5, nrow = 0)), c("type", "ind", "time", "coefhat", "tstat"))
-    
-    state <- this.state
+        state <- this.state
     dat2 <- filter(dat, location == state)
     dat3 <- ts(dat2$Target, start = year(min(dat$Year)), end = c(year(max(dat$Year)),12), frequency = 12)
     outlier.county <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)
     df2<- outlier.county$outliers
     df2$location <- state
     df <- bind_rows(df, df2)
-    
     df_sums2 <- setNames(data.frame(matrix(ncol=3, nrow=1)), c("y", "yadj", "loc"))
     df_sums2$y <- sum(outlier.county$y)
[…]

### 2021-03-29 R/SCRIPTS/SuppMat_figures.R

@@ -1,0 +1,205 @@
+library(future)
+library(future.apply)
+library(parallel)
+anom_mort <- read_rds("./R/DATA-PROCESSED/mortality_anomalies") %>%
+  na.omit %>%
+  mutate(component = "Mortality")
+
+anom_fert <- read_rds("./R/DATA-PROCESSED/fertility_anomalies") %>%
+  na.omit %>%
+  mutate(component = "Fertility")
+
+### Getting a FIPS list
+# This will download a list of FIPS codes from online.
+fipslist <- read_csv(file="https://www2.census.gov/geo/docs/reference/codes/files/national_county.txt", col_names = FALSE) %>%
+  mutate(GEOID = paste0(X2, X3),
+         STATECOUNTY = paste0(X1, X4)) %>%
+  dplyr::rename(state = X1,
+                STATEID = X2,
+                CNTYID = X3,
+                NAME = X4) %>%
+  filter(!STATEID %in% c("60", "66", "69", "72", "74", "78")) %>%
+  dplyr::select(state, STATEID) %>%
+  distinct()
+
+abstract <- rbind(anom_mort, anom_fert) %>%
+  mutate(STATEID = substr(location,2,3)) %>%
+  left_join(., fipslist) %>%
+  dplyr::select(state, STATEID, component, time, type, tstat, coefhat, -ind, -location) %>%
+  arrange(STATEID, component, time)
+dat_fert <-
+  ###FERTILITY
+  bind_rows(read_tsv("./R/DATA-RAW/Fertility/Natality, 2003-2006,states.txt"),
+            read_tsv("./R/DATA-RAW/Fertility/Natality, 2007-2017,states.txt"))%>%
+  mutate(Month2 = str_pad(`Month Code`,2 , pad = "0"),
+         Month3 = as.yearmon(paste0(Year,"-", Month2)),
+         location = paste0("X", `State Code`)
+  ) %>%
+  dplyr::select(Year = Month3, location, Target = Births) %>%
+  na.omit %>%
+  arrange(location, Year)
+
+dat_mort <- 
+  ###MORTALITY 
+  read_tsv("./R/DATA-RAW/Mortality/Underlying Cause of Death, 1999-2017,states.txt") %>%
+  mutate(Month2 = substr(`Month Code`,6,7),
+         Month3 = as.yearmon(paste0(Year,"-", Month2)),
+         # location = paste0("X", `County Code`)
+         location = paste0("X", `State Code`)
+  ) %>%
[…]

### 2021-03-29 R/SCRIPTS/robustnesscheck.R

@@ -1,0 +1,115 @@
+
+
+countylist = unique(dat_fert$location)
+
+noamom_fert <- setdiff(countylist, unique(anom_fert$location))
+
+noamom_mort <- setdiff(countylist, unique(anom_mort$location))
+
+# dat <- ts(dat$Target)
+# fit <- tso(dat, cval=3.5)
+# sdval <- sqrt(fit$fit$sigma2)
+# arima.sim(n=length(fit$y), model=list(fit$fit), sd = sdval) + as.vector(fit$yadj)
+# results_noutliers <- data.frame()
+# y2 <- matrix(nrow = 1000, ncol=180)
+for(this.state in noamom_mort){
+  print(this.state)
+  dat <- ts(filter(dat_mort, location == this.state)$Target)
+  fit <- tso(dat, cval=3.5)
+  sdval <- sqrt(fit$fit$sigma2)
+
+for(i in 1:1000){
+  tryCatch({
+    # y <- arima.sim(model = list(ar = 0.7, ma = -0.4), n = 180)
+    y <- arima.sim(n=length(fit$y), model=list(fit$fit), sd = sdval) + as.vector(fit$yadj)
+    
+    df2<- tso(y, cval = 3.5)$outliers
+    df2$iteration <- i
+    df2$type <- "mortality"
+    results_noutliers <- bind_rows(results_noutliers, df2)
+    # print(i)
+  } , error=function(e){#cat("ERROR :",conditionMessage(e), "\n")
+    })
+}
+}
+
+for(this.state in noamom_fert){
+  print(this.state)
+  dat <- ts(filter(dat_fert, location == this.state)$Target)
+  fit <- tso(dat, cval=3.5)
+  sdval <- sqrt(fit$fit$sigma2)
+  
+  for(i in 1:1000){
+    tryCatch({
+      # y <- arima.sim(model = list(ar = 0.7, ma = -0.4), n = 180)
+      y <- arima.sim(n=length(fit$y), model=list(fit$fit), sd = sdval) + as.vector(fit$yadj)
+      
+      df2<- tso(y, cval = 3.5)$outliers
+      df2$iteration <- i
+      df2$type <- "fertility"
+      results_noutliers <- bind_rows(results_noutliers, df2)
+      # print(i)
+    } , error=function(e){#cat("ERROR :",conditionMessage(e), "\n")
+    })
+  }
+}
+
+z <- results_noutliers %>%
[…]
